### 2020-06-07 papers/SBLP2020/latex/paper.tex
Need some more fixes on the Agda section on the paper

\documentclass[sigconf]{acmart}

\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{agda}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\setcopyright{acmcopyright}
\copyrightyear{2020}
\acmYear{2020}
%\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[SBLP '20]{SBLP '20: Brazilian Symposium on Programming Languages}{October 19--23, 2020}{Natal, Brazil}
\acmBooktitle{SBLP '20: Brazilian Symposium on Programming Languages,
  October 19--23, 2020, Natal, Brazil}
%\acmPrice{15.00}
%\acmISBN{978-1-4503-XXXX-X/18/06}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% agda stuff                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{newunicodechar}
\newunicodechar{γ}{$\gamma$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{τ}{$\tau$}
\newunicodechar{Π}{$\Pi$}
\newunicodechar{→}{$\rightarrow$}
\newunicodechar{⇒}{$\Rightarrow$}
\newunicodechar{∈}{$\in$}
\newunicodechar{∼}{$\sim$}
\newunicodechar{ₗ}{$_l$}
\newunicodechar{ᵣ}{$_r$}
\newunicodechar{∀}{$\forall$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{ℕ}{$\mathbb{N}$}
\newunicodechar{∷}{$::$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{₄}{$_4$}
\newunicodechar{ₑ}{$_e$}
\newunicodechar{σ}{$\sigma$}
\newunicodechar{Δ}{$\Delta$}
\newunicodechar{ᴺ}{$^{N}$}
\newunicodechar{∘}{$\circ$}
\newunicodechar{δ}{$\delta$}
\newunicodechar{Σ}{$\Sigma$}
\newunicodechar{λ}{$\lambda$}
\newunicodechar{ᶜ}{$^{c}$}
\newunicodechar{⊎}{$\biguplus$}
\newunicodechar{∧}{$\land$}

\begin{document}

\title{A Certified Interpreter for the List Machine Benchmark}

\author{Samuel Feitosa}
\authornotemark[1]
\email{samuel.feitosa@ifsc.edu.br}
\affiliation{%
  \institution{Departamento de Informática}
  \streetaddress{Instituto Federal de Santa Catarina}
  \city{Caçador}
  \state{Santa Catarina}
  \country{Brazil}
}

\author{Rodrigo Ribeiro}
\email{rodrigo.ribeiro@ufop.edu.br}
\affiliation{%
  \institution{Prog. Pós Graduação em Ciência da Computação}
  \streetaddress{Universidade Federal de Ouro Preto}
  \city{Ouro Preto}
  \state{Minas Gerais}
  \country{Brazil}}

\begin{abstract}
This is the abstract...
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011039.10011311</concept_id>
<concept_desc>Software and its engineering~Semantics</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041.10010943</concept_id>
<concept_desc>Software and its engineering~Interpreters</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Semantics}
\ccsdesc[500]{Software and its engineering~Interpreters}
\ccsdesc[500]{Theory of computation~Type theory}

\keywords{Dependent types, formal semantics}

\maketitle

\section{Introduction}

\section{An Overview of Agda}\label{sec:agda}

Agda is a dependently-typed functional programming language based on
Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
and an infinite hierarchy of types of types, \AgdaDatatype{Set l}, where
\AgdaDatatype{l} is a natural number, are built-in. Everything else is a user-defined
type. The type \AgdaDatatype{Set}, also known as \AgdaDatatype{Set₀}, is the type of all
``small'' types, such as \AgdaDatatype{Bool}, \AgdaDatatype{String} and
\AgdaDatatype{List Bool}.  The type \AgdaDatatype{Set₁} is the type of \AgdaDatatype{Set}
and ``others like it'', such as \AgdaDatatype{Set → Bool}, \AgdaDatatype{String → Set}, and
\AgdaDatatype{Set → Set}. We have that \AgdaDatatype{Set l} is an
element of the type \AgdaDatatype{Set (l+1)}, for every $l \geq 0$. This
stratification of types is used to keep Agda consistent as a logical
theory~\cite{Sorensen2006}.

An ordinary (non-dependent) function type is written \AgdaDatatype{A → B} and a
dependent one is written \AgdaDatatype{(x : A) → B}, where type \AgdaDatatype{B}
depends on \AgdaDatatype{x}, or \AgdaDatatype{∀ (x : A) → B}|. Agda allows the
definition of \emph{implicit parameters}, i.e.,  parameters whose values can be
inferred from the context, by surrounding them in curly braces:
\AgdaDatatype{∀ {x : A} → B}. In order to avoid clutter, we'll omit implicit
arguments from the source code presentation. The reader can safely assume
that every free variable in a type is an implicit parameter.

As an example of Agda code, consider the following data type of
length-indexed lists, also known as vectors.

\begin{code}%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[4]\AgdaInductiveConstructor{zero}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\<%
\\
%
\>[4]\AgdaInductiveConstructor{suc}%
\>[9]\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{ℕ}\<%
\\
%
\\[\AgdaEmptyExtraSkip]%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{A}\AgdaSymbol{\}(}\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[4]\AgdaInductiveConstructor{[]}%
\>[8]\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaInductiveConstructor{zero}\<%
\\
%
\>[4]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∷\AgdaUnderscore{}}}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{n}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{n}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{)}\<%
\end{code}
\begin{code}[hide]%
%
\>[2]\AgdaKeyword{infixr}\AgdaSpace{}%
\AgdaNumber{4}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∷\AgdaUnderscore{}}}\<%
\end{code}

Constructor \AgdaInductiveConstructor{[]} builds empty vectors. The cons-operator (\AgdaInductiveConstructor{_∷_})
inserts a new element in front of a vector of $n$ elements (of type
\AgdaDatatype{Vec A n}) and returns a value of type \AgdaDatatype{Vec A (suc n)}. The
\AgdaDatatype{Vec} datatype is an example of a dependent type, i.e., a type
that uses a value (that denotes its length). The usefulness of
dependent types can be illustrated with the definition of a safe list
head function: \AgdaFunction{head} can be defined to accept only non-empty
vectors, i.e.,~values of type \AgdaDatatype{Vec A (suc n)}.
\begin{code}%
%
\>[2]\AgdaFunction{head}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{A}\<%
\\
%
\>[2]\AgdaFunction{head}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor{∷}}\AgdaSpace{}%
\AgdaBound{xs}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{=}\AgdaSpace{}%
\AgdaBound{x}\<%
\end{code}
In \AgdaFunction{head}'s definition, constructor \AgdaInductiveConstructor{[]} is not used. The
Agda type-checker can figure out, from \AgdaFunction{head}'s parameter type,
that argument \AgdaInductiveConstructor{[]} to \AgdaFunction{head} is not type-correct.

Another useful data type is the finite type, \AgdaDatatype{Fin}\footnote{Note that Agda supports the overloading of data type constructor names.
Constructor \AgdaInductiveConstructor{zero} can refer to type \AgdaDatatype{ℕ} or \AgdaDatatype{Fin}, depending on the
context where the name is used.}, which is defined in
Agda's standard library as:

\begin{code}%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[4]\AgdaInductiveConstructor{zero}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{n}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{)}\<%
\\
%
\>[4]\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{n}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaBound{n}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{)}\<%
\end{code}
Type \AgdaDatatype{Fin n} has exactly \AgdaDatatype{n} inhabitants
(elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
An application of such type is to define a safe vector
lookup function, which avoids the access of invalid positions.
\begin{code}%
%
\>[2]\AgdaFunction{lookup}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaBound{n}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{n}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{A}\<%
\\
%
\>[2]\AgdaFunction{lookup}\AgdaSpace{}%
\AgdaInductiveConstructor{zero}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor{∷}}\AgdaSpace{}%
\AgdaSymbol{\AgdaUnderscore{})}\AgdaSpace{}%
\AgdaSymbol{=}\AgdaSpace{}%
\AgdaBound{x}\<%
\\
%
\>[2]\AgdaFunction{lookup}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaInductiveConstructor{suc}\AgdaSpace{}%
\AgdaBound{idx}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{(\AgdaUnderscore{}}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor{∷}}\AgdaSpace{}%
\AgdaBound{xs}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{=}\AgdaSpace{}%
\AgdaFunction{lookup}\AgdaSpace{}%
\AgdaBound{idx}\AgdaSpace{}%
\AgdaBound{xs}\<%
\end{code}
Thanks to the propositions-as-types principle,\footnote{It is also known as
  Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
types as logical formulas and terms as proofs. An example is the
representation of equality as the following Agda type:

\begin{code}%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaOperator{\AgdaDatatype{\AgdaUnderscore{}≡\AgdaUnderscore{}}}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{l}\AgdaSymbol{\}\{}\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaBound{l}\AgdaSymbol{\}(}\AgdaBound{x}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaBound{A}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[4]\AgdaInductiveConstructor{refl}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaBound{x}\AgdaSpace{}%
\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
\AgdaBound{x}\<%
\end{code}

This type is called propositional equality. It defines that there is
a unique evidence for equality, constructor \AgdaInductiveConstructor{refl} (for reflexivity),
that asserts that the only value equal to \AgdaInductiveConstructor{x} is itself.
Given a predicate \AgdaDatatype{P : A → Set} and a vector \AgdaInductiveConstructor{xs}, the
type \AgdaDatatype{All P xs} is used to build proofs that \AgdaDatatype{P} holds for all
elements in \AgdaDatatype{xs} and it is defined as:
\begin{code}%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{All}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{n}\AgdaSymbol{\}(}\AgdaBound{P}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSymbol{)}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Vec}\AgdaSpace{}%
\AgdaBound{A}\AgdaSpace{}%
\AgdaBound{n}\AgdaSpace{}%
\AgdaSymbol{→}%
\>[43]\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[5]\AgdaInductiveConstructor{[]}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{All}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaInductiveConstructor{[]}\<%
\\
%
\>[5]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∷\AgdaUnderscore{}}}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaSymbol{∀}\AgdaSpace{}%
\AgdaSymbol{\{}\AgdaBound{x}\AgdaSpace{}%
\AgdaBound{xs}\AgdaSymbol{\}}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaBound{x}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{All}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaBound{xs}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{All}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor{∷}}\AgdaSpace{}%
\AgdaBound{xs}\AgdaSymbol{)}\<%
\end{code}
The first constructor specifies that \AgdaDatatype{All P} holds for the empty vector and
constructor \AgdaInductiveConstructor{_∷_} builds a proof of \AgdaDatatype{All P (x :: xs)} from proofs of
\AgdaDatatype{P x} and \AgdaDatatype{All P xs}. Since this type has the same structure of vectors,
some functions on \AgdaDatatype{Vec} have similar definitions for type \AgdaDatatype{All}. As an example
used in our formalization, consider the function \AgdaFunction{lookup}, which extracts a
proof of \AgdaDatatype{P} for the element at position \AgdaDatatype{v : Fin n} in a \AgdaDatatype{Vec}:
\begin{spec}
  lookup : ∀ {A n P x}{xs : Vec A n} → Fin n → All P xs → P x
  lookup zero (px ∷ _) = px
  lookup (succ idx) (_ ∷ pxs) = lookup idx pxs
\end{spec}
An important application of dependent types is to encode programming languages
syntax. The role of dependent types in this domain is to encode programs that
only allow well-typed and well-scoped terms~\cite{Benton2012}. Intuitively, we encode
the static semantics of the object language in the host language AST's
constructor, leaving the responsibility of checking type safety to the
host's language type checker. As an example, consider the following simple
expression language.
\begin{spec}
  data Expr : Set where
    True False : Expr
    Num : ℕ → Expr
    _∧_ _+_ : Expr → Expr → Expr
\end{spec}
Using this data type,\footnote{Agda supports the definition of mixfix operators.
We can use underscores to mark arguments positions.} we can construct expressions
that denote terms that should not be considered well-typed like
\AgdaFunction{(Num 1) + True}. Using this approach, we need to specify the static semantics
as another definition, which should consider all possible cases to avoid the
definition of ill-typed terms.

A better approach is to combine the static semantics and language syntax into
a single definition, as shown below.

\begin{code}%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{Ty}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[4]\AgdaInductiveConstructor{Nat}\AgdaSpace{}%
\AgdaInductiveConstructor{Bool}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Ty}\<%
\\
%
\\[\AgdaEmptyExtraSkip]%
%
\>[2]\AgdaKeyword{data}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Ty}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaPrimitiveType{Set}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
\>[2][@{}l@{\AgdaIndent{0}}]%
\>[4]\AgdaInductiveConstructor{True}\AgdaSpace{}%
\AgdaInductiveConstructor{False}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Bool}\<%
\\
%
\>[4]\AgdaInductiveConstructor{Num}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{ℕ}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Nat}\<%
\\
%
\>[4]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∧\AgdaUnderscore{}}}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Bool}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Bool}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Bool}\<%
\\
%
\>[4]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}+\AgdaUnderscore{}}}\AgdaSpace{}%
\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Nat}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Nat}\AgdaSpace{}%
\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Expr}\AgdaSpace{}%
\AgdaInductiveConstructor{Nat}\<%
\end{code}

In this definition, the \AgdaDatatype{Expr} type is indexed by a value of type \AgdaDatatype{Ty} which
indicates the type of the expression being built. In this approach, Agda's
type system can enforce that only well-typed terms could be written.
Agda's type checker will automatically reject a definition which uses the expression
\AgdaFunction{(Num 1) + True}.

%Dependently typed pattern matching is built by using the so-called
%|with| construct, that allows for matching intermediate
%values~\cite{McBride2004}. If the matched value has a dependent type,
%then its result can affect the form of other values. For example,
%consider the following code that defines a type for natural number
%parity. If the natural number is even, it can be represented as the
%sum of two equal natural numbers; if it is odd, it is equal to one
%plus the sum of two equal values. Pattern matching on a value of
%|Parity n| allows to discover if $n = j + j$ or $n = S (k + k)$,
%for some $j$ and $k$ in each branch of |with|.  Note that the
%value of $n$ is specialized accordingly, using information ``learned''
%by the type-checker.
%\begin{spec}
%data Parity : Nat -> Set where
%   Even : forall {n : Nat} -> Parity (n + n)
%   Odd  : forall {n : Nat} -> Parity (S (n + n))
%
%parity : (n : Nat) -> Parity n
%parity = -- definition omitted

%natToBin : Nat -> List Bool
%natToBin zero = []
%natToBin k with (parity k)
%   natToBin (j + j)     | Even = false :: natToBin j
%   natToBin (succ (j + j)) | Odd  = true  :: natToBin j
%\end{spec}

For further information about Agda, see~\cite{Norell2009,Stump16}.

\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}
\end{document}
\endinput

### 2020-06-07 papers/SBLP2020/paper.tex
Using lhs2TeX instead of latex backend for better formatting.

\documentclass[sigconf]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage[utf8x]{inputenc}
\usepackage{ucs}
\usepackage{graphicx}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{ifpdf}
\ifpdf
\usepackage{hyperref}
\fi
\usepackage{float}
\usepackage{proof}

% Copyright
%\setcopyright{none}
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\setcopyright{acmcopyright}
\copyrightyear{2020}
\acmYear{2020}

\acmConference[SBLP '20]{SBLP '20: Brazilian Symposium on Programming Languages}{October 19--23, 2020}{Natal, Brazil}
\acmBooktitle{SBLP '20: Brazilian Symposium on Programming Languages,
  October 19--23, 2020, Natal, Brazil}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% color formatting stuff %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newtheorem{Lemma}{Lemma}
\newtheorem{Theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{Example}{Example}

\usepackage{color}
\newcommand{\redFG}[1]{\textcolor[rgb]{0.6,0,0}{#1}}
\newcommand{\greenFG}[1]{\textcolor[rgb]{0,0.4,0}{#1}}
\newcommand{\blueFG}[1]{\textcolor[rgb]{0,0,0.8}{#1}}
\newcommand{\orangeFG}[1]{\textcolor[rgb]{0.8,0.4,0}{#1}}
\newcommand{\purpleFG}[1]{\textcolor[rgb]{0.4,0,0.4}{#1}}
\newcommand{\yellowFG}[1]{\textcolor{yellow}{#1}}
\newcommand{\brownFG}[1]{\textcolor[rgb]{0.5,0.2,0.2}{#1}}
\newcommand{\blackFG}[1]{\textcolor[rgb]{0,0,0}{#1}}
\newcommand{\whiteFG}[1]{\textcolor[rgb]{1,1,1}{#1}}
\newcommand{\yellowBG}[1]{\colorbox[rgb]{1,1,0.2}{#1}}
\newcommand{\brownBG}[1]{\colorbox[rgb]{1.0,0.7,0.4}{#1}}

\newcommand{\ColourStuff}{
  \newcommand{\red}{\redFG}
  \newcommand{\green}{\greenFG}
  \newcommand{\blue}{\blueFG}
  \newcommand{\orange}{\orangeFG}
  \newcommand{\purple}{\purpleFG}
  \newcommand{\yellow}{\yellowFG}
  \newcommand{\brown}{\brownFG}
  \newcommand{\black}{\blackFG}
  \newcommand{\white}{\whiteFG}
}

\newcommand{\MonochromeStuff}{
  \newcommand{\red}{\blackFG}
  \newcommand{\green}{\blackFG}
  \newcommand{\blue}{\blackFG}
  \newcommand{\orange}{\blackFG}
  \newcommand{\purple}{\blackFG}
  \newcommand{\yellow}{\blackFG}
  \newcommand{\brown}{\blackFG}
  \newcommand{\black}{\blackFG}
  \newcommand{\white}{\blackFG}
}

\ColourStuff

%%%%%%%%%%%%%%%%%%%%
%% lhs2TeX stuff  %%
%%%%%%%%%%%%%%%%%%%%

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

\DeclareMathAlphabet{\mathkw}{OT1}{cmss}{bx}{n}

\newcommand{\D}[1]{\blue{\mathsf{#1}}}
\newcommand{\Con}[1]{\red{\mathsf{#1}}}
\newcommand{\F}[1]{\green{\mathsf{#1}}}
\newcommand{\V}[1]{\purple{\mathit{#1}}}

\begin{document}

\title{A Certified Interpreter for the List Machine Benchmark}

\author{Samuel Feitosa}
\authornotemark[1]
\email{samuel.feitosa@ifsc.edu.br}
\affiliation{%
  \institution{Departamento de Informática}
  \streetaddress{Instituto Federal de Santa Catarina}
  \city{Caçador}
  \state{Santa Catarina}
  \country{Brazil}
}

\author{Rodrigo Ribeiro}
\email{rodrigo.ribeiro@ufop.edu.br}
\affiliation{%
  \institution{Prog. Pós Graduação em Ciência da Computação}
  \streetaddress{Universidade Federal de Ouro Preto}
  \city{Ouro Preto}
  \state{Minas Gerais}
  \country{Brazil}}

\begin{abstract}
This is the abstract...
\end{abstract}

\begin{CCSXML}\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\V{ccs2012>}{}\<[E]%
\\
\>[B]{}\V{concept>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 id>10011007.10011006.10011039.10011311</concept\char95 id>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 desc>Software}\;\V{and}\;\V{its}\;\V{engineering\char126 Semantics</concept\char95 desc>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 significance>500</concept\char95 significance>}{}\<[E]%
\\
\>[B]{}\V{/concept>}{}\<[E]%
\\
\>[B]{}\V{concept>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 id>10011007.10011006.10011041.10010943</concept\char95 id>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 desc>Software}\;\V{and}\;\V{its}\;\V{engineering\char126 Interpreters</concept\char95 desc>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 significance>500</concept\char95 significance>}{}\<[E]%
\\
\>[B]{}\V{/concept>}{}\<[E]%
\\
\>[B]{}\V{concept>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 id>10003752.10003790.10011740</concept\char95 id>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 desc>Theory}\;\V{of}\;\V{computation\char126 Type}\;\V{theory</concept\char95 desc>}{}\<[E]%
\\
\>[B]{}\V{concept\char95 significance>500</concept\char95 significance>}{}\<[E]%
\\
\>[B]{}\V{/concept>}{}\<[E]%
\\
\>[B]{}\V{/ccs2012>}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Semantics}
\ccsdesc[500]{Software and its engineering~Interpreters}
\ccsdesc[500]{Theory of computation~Type theory}

\keywords{Dependent types, formal semantics}

\maketitle

\section{Introduction}

\section{An Overview of Agda}\label{sec:agda}

Agda is a dependently-typed functional programming language based on
Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
natural number, are built-in. Everything else is a user-defined
type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}}, \ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}. We have that \ensuremath{\D{Set}\;\V{l}} is an
element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
stratification of types is used to keep Agda consistent as a logical
theory~\cite{Sorensen2006}.

An ordinary (non-dependent) function type is written \ensuremath{\V{A}\;\to \;\V{B}} and a
dependent one is written \ensuremath{(\V{x}\;\mathbin{:}\;\V{A})\;\to \;\V{B}}, where type \ensuremath{\V{B}} depends on
\ensuremath{\V{x}}, or \ensuremath{\D{\forall}\;(\V{x}\;\mathbin{:}\;\V{A})\;\to \;\V{B}}. Agda allows the definition of \emph{implicit
parameters}, i.e.,  parameters whose values can be inferred from the
context, by surrounding them in curly braces: \ensuremath{\D{\forall}\;\{\mskip1.5mu \V{x}\;\mathbin{:}\;\V{A}\mskip1.5mu\}\;\to \;\V{B}}. To
avoid clutter, we'll omit implicit arguments from the source code
presentation. The reader can safely assume that every free variable in
a type is an implicit parameter.

As an example of Agda code, consider the following data type of
length-indexed lists, also known as vectors.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{\mathbb{N}}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\mathbb{N}}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{\mathbb{N}}{}\<[E]%
\\[\blanklineskip]%
\>[3]{}\mathkw{data}\;\D{Vec}\;(\V{A}\;\mathbin{:}\;\D{Set})\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{\lbrack\:\rbrack}\;{}\<[9]%
\>[9]{}\mathbin{:}\;\D{Vec}\;\V{A}\;\Con{zero}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\V{A}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})
inserts a new element in front of a vector of $n$ elements (of type
\ensuremath{\D{Vec}\;\V{A}\;\V{n}}) and returns a value of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}. The
\ensuremath{\D{Vec}} datatype is an example of a dependent type, i.e., a type
that uses a value (that denotes its length). The usefulness of
dependent types can be illustrated with the definition of a safe list
head function: \ensuremath{\F{head}} can be defined to accept only non-empty
vectors, i.e.,~values of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\F{head}\;\mathbin{:}\;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})\;\to \;\V{A}{}\<[E]%
\\
\>[3]{}\F{head}\;(\V{x}\;\Con{::}\;\V{xs})\;\mathrel{=}\;\V{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
In \ensuremath{\F{head}}'s definition, constructor \ensuremath{\Con{\lbrack\:\rbrack}} is not used. The
Agda type-checker can figure out, from \ensuremath{\F{head}}'s parameter type,
that argument \ensuremath{\Con{\lbrack\:\rbrack}} to \ensuremath{\F{head}} is not type-correct.

Another useful data type is the finite type, \ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of data type constructor names.
Constructor \ensuremath{\Con{zero}} can refer to type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the
context where the name is used.}, which is defined in
Agda's standard library as:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{Fin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Type \ensuremath{\D{Fin}\;\V{n}} has exactly \ensuremath{\V{n}} inhabitants
(elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
An application of such type is to define a safe vector
lookup function, which avoids the access of invalid positions.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\F{lookup}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\V{A}{}\<[E]%
\\
\>[3]{}\F{lookup}\;\Con{zero}\;(\V{x}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{x}{}\<[E]%
\\
\>[3]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{xs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Thanks to the propositions-as-types principle,\footnote{It is also known as
  Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
types as logical formulas and terms as proofs. An example is the
representation of equality as the following Agda type:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{\_ \equiv \_}\;\{\mskip1.5mu \V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{A}\;\mathbin{:}\;\D{Set}\;\V{l}\mskip1.5mu\}\;(\V{x}\;\mathbin{:}\;\V{A})\;\mathbin{:}\;\V{A}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{refl}\;\mathbin{:}\;\V{x}\;\D{\equiv}\;\V{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This type is called propositional equality. It defines that there is
a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
that asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
and a vector \ensuremath{\V{xs}}, the type \ensuremath{\D{All}\;\V{P}\;\V{xs}} is used to build proofs that \ensuremath{\V{P}} holds for all
elements in \ensuremath{\V{xs}} and it is defined as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{All}\;(\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
\>[41]{}\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{3}{}\<[6]%
\>[6]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\V{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
\\
\>[3]{}\hsindent{3}{}\<[6]%
\>[6]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\V{P}\;\V{x}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The first constructor specifies that \ensuremath{\D{All}\;\V{P}} holds for the empty vector and
constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
\ensuremath{\V{P}\;\V{x}} and \ensuremath{\D{All}\;\V{P}\;\V{xs}}. Since this type has the same structure of vectors,
some functions on \ensuremath{\D{Vec}} have similar definitions for type \ensuremath{\D{All}}. As an example
used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\Con{::}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\V{P}\;\V{x}{}\<[E]%
\\
\>[4]{}\F{lookup}\;\Con{zero}\;(\V{px}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{px}{}\<[E]%
\\
\>[4]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{pxs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{pxs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
An important application of dependent types is to encode programming languages
syntax. The role of dependent types in this domain is to encode programs that
only allow well-typed and well-scoped terms~\cite{Benton2012}. Intuitively, we encode
the static semantics of the object language in the host language AST's
constructor, leaving the responsibility of checking type safety to the
host's language type checker. As an example, consider the following simple
expression language.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[4]{}\mathkw{data}\;\D{Expr}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{True}\;\Con{False}\;\mathbin{:}\;\D{Expr}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{Num}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Expr}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{\_\land\_}\;\Con{\_+\_}\;\mathbin{:}\;\D{Expr}\;\to \;\D{Expr}\;\to \;\D{Expr}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Using this data type,\footnote{Agda supports the definition of mixfix operators.
We can use underscores to mark arguments positions.} we can construct expressions
that denote terms that should not be considered well-typed like
\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
as another definition, which should consider all possible cases to avoid the
definition of ill-typed terms.

A better approach is to combine the static semantics and language syntax into
a single definition, as shown below.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[4]{}\mathkw{data}\;\D{Ty}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{Nat}\;\Con{Bool}\;\mathbin{:}\;\D{Ty}{}\<[E]%
\\[\blanklineskip]%
\>[4]{}\mathkw{data}\;\D{Expr}\;\mathbin{:}\;\D{Ty}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{True}\;\Con{False}\;\mathbin{:}\;\D{Expr}\;\Con{Bool}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{Num}\;\mathbin{:}\;\Con{Nat}\;\to \;\D{Expr}\;\Con{Nat}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{\_\land\_}\;\mathbin{:}\;\D{Expr}\;\Con{Bool}\;\to \;\D{Expr}\;\Con{Bool}\;\to \;\D{Expr}\;\Con{Bool}{}\<[E]%
\\
\>[4]{}\hsindent{3}{}\<[7]%
\>[7]{}\Con{\_+\_}\;\mathbin{:}\;\D{Expr}\;\Con{Nat}\;\to \;\D{Expr}\;\Con{Nat}\;\to \;\D{Expr}\;\Con{Nat}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In this definition, the \ensuremath{\D{Expr}} type is indexed by a value of type \ensuremath{\D{Ty}} which
indicates the type of the expression being built. In this approach, Agda's
type system can enforce that only well-typed terms could be written.
%A definition which uses the expression |(Num 1) + True| will be rejected by Agda's type checker automatically.
Agda's type checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}.

%Dependently typed pattern matching is built by using the so-called
%|with| construct, that allows for matching intermediate
%values~\cite{McBride2004}. If the matched value has a dependent type,
%then its result can affect the form of other values. For example,
%consider the following code that defines a type for natural number
%parity. If the natural number is even, it can be represented as the
%sum of two equal natural numbers; if it is odd, it is equal to one
%plus the sum of two equal values. Pattern matching on a value of
%|Parity n| allows to discover if $n = j + j$ or $n = S (k + k)$,
%for some $j$ and $k$ in each branch of |with|.  Note that the
%value of $n$ is specialized accordingly, using information ``learned''
%by the type-checker.
%\begin{spec}
%data Parity : Nat -> Set where
%   Even : forall {n : Nat} -> Parity (n + n)
%   Odd  : forall {n : Nat} -> Parity (S (n + n))
%
%parity : (n : Nat) -> Parity n
%parity = -- definition omitted

%natToBin : Nat -> List Bool
%natToBin zero = []
%natToBin k with (parity k)
%   natToBin (j + j)     | Even = false :: natToBin j
%   natToBin (succ (j + j)) | Odd  = true  :: natToBin j
%\end{spec}

For further information about Agda, see~\cite{Norell2009,Stump16}.

\bibliographystyle{ACM-Reference-Format}
\bibliography{main}
\end{document}
\endinput

### 2020-06-08 papers/SBLP2020/paper.tex
Some improvements on text. Put Samuels coding of list benchmark rules

@@
 \usepackage{ifpdf}
+\usepackage{mathtools}
+\usepackage{semantic}
+
 \ifpdf
@@
 \begin{abstract}
-This is the abstract...
+Formal models are an important tool in the programming languages research
+community. However, such models are full of intricacies and, due to that,
+they are subject to subtle errors. Such failures motivated the usage of
+tools to ensure the correctness of these formalisms. One way to eliminate
+such errors is to encode models in a dependently-typed language in order
+to ensure ``correctness-by-construction''.
+
+In this work, we use follow this idea to build the verified interpreter
+for the list machine benchmark in the Agda programming language.
 \end{abstract}
@@
 
-\section{An Overview of Agda}\label{sec:agda}
+The development of new programming language design, linguistic construct
+or type system involves its careful formalization in order to express
+its core ideas in a concise way. However, such models have many details
+and complexities which hinders its correctness assurances.
+Because of such problems, the programming languages research community
+started to use tools, like proof assistants~\cite{Stump16,Chlipala13},
+and benchmark problems to validate them and stress its suitability for
+such tasks~\cite{Aydemir05,Pientka18,Appel07}.
 
+The rest of this paper is organized as follows. Section~\ref{sec:background}
+presents a brief introduction to Agda and reviews the list machine benchmark.
+In Section~\ref{sec:typing}, we
+describe the type system for PEGs and its relation with the original
+well-formedness predicate proposed by Ford~\cite{Ford2004}.
+Section~\ref{sec:interpreter} describes the intrinsically typed syntax for PEGs
+and its interpreter. Limitations of our approach are discussed
+in Section~\ref{sec:discussion}. Related work is discussed in
+Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.
+
+All the source code in this article has been formalized in Agda
+version 2.6.0 using Standard Library 1.0. All source code produced,
+including the \LaTeX~ source of this
+article, are available on-line~\cite{peg-rep}.
+
+\section{Background}\label{sec:background}
+
+\paragraph{An Overview of Agda}
 Agda is a dependently-typed functional programming language based on
@@
 used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
-proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\Con{::}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
+proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
 \begin{hscode}\SaveRestoreHook
@@
 
+\paragraph{The List Machine Benchmark}
+
+\begin{equation}
+\inference{}
+          {(r, (\iota_1;\iota_2);\iota_3) \xmapsto{p} (r, \iota_1;(\iota_2;\iota_3))}[step-seq]
+\end{equation}
+
+\begin{equation}
+\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_0]=r'}
+          {(r, (\textbf{fetch-field}~v~0~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-0]
+\end{equation}
+
+\begin{equation}
+\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_1]=r'}
+          {(r, (\textbf{fetch-field}~v~1~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-1]
+\end{equation}
+
+\begin{equation}
+\inference{r(v_0) = a_0~~~r(v_1)=a_1~~~r[v':=cons(a_0,a_1)]=r'}
+          {(r, (\textbf{cons}~v_0~v_1~v';\iota)) \xmapsto{p} (r', \iota)}[step-cons]
+\end{equation}
+
+\begin{equation}
+\inference{r(v) = cons(a_0,a_1)}
+          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota)}[step-branch-not-taken]
+\end{equation}
+
+\begin{equation}
+\inference{r(v) = nil~~~p(l)=\iota'}
+          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota')}[step-branch-taken]
+\end{equation}
+
+\begin{equation}
+\inference{p(l)=\iota'}
+          {(r, \textbf{jump}~l) \xmapsto{p} (r, \iota')}[step-jump]
+\end{equation}
+
+\begin{equation}
+\inference{(r, \iota) \xmapsto{p} (r', \iota')~~~(p, r', \iota')}
+          {(p, r, \iota) \Downarrow}[run-step]
+\end{equation}
+
+\begin{equation}
+\inference{}
+          {(p, r, \textbf{halt}) \Downarrow}[run-halt]
+\end{equation}
+
+\begin{equation}
+\inference{\{\}[\textbf{v}_0:=\text{nil}]=r~~~p(\textbf{L}_0)=\iota~~~(p,r,\iota)\Downarrow}
+          {p \Downarrow}[run-prog]
+\end{equation}
+
+%\subsection{Typing}
+
+%\subsubsection{Subtyping}
+
+\begin{equation}
+\inference{}
+          {\tau \subset \tau}[subtype-refl]
+\end{equation}
+
+\begin{equation}
+\inference{}
+          {\text{nil} \subset \text{list}~\tau}[subtype-nil]
+\end{equation}
+
+\begin{equation}
+\inference{\tau \subset \tau'}
+          {\text{list}~\tau \subset \text{list}~\tau'}[subtype-list]
+\end{equation}
+
+\begin{equation}
+\inference{\tau \subset \tau'}
+          {\text{listcons}~\tau \subset \text{list}~\tau'}[subtype-listcons]
+\end{equation}
+
+\begin{equation}
+\inference{\tau \subset \tau'}
+          {\text{listcons}~\tau \subset \text{listcons}~\tau'}[subtype-listmixed]
+\end{equation}
+
+%\subsubsection{Instruction typing}
+
+\begin{equation}
+\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'~~~\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[check-instr-seq]
+\end{equation}
+
+\begin{equation}
+\inference{\Gamma(v) = \text{list}~\tau~~~\Pi(l) = \Gamma_1~~~\Gamma[v:=\text{nil}]=\Gamma'~~~\Gamma' \subset \Gamma_1}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}(v: \text{listcons}~\tau,~\Gamma')}[check-instr-branch-list]
+\end{equation}
+
+\begin{equation}
+\inference{\Gamma(v) = \text{listcons}~\tau~~~\Pi(l) = \Gamma_1~~~\Gamma[v:=\text{nil}]=\Gamma'~~~\Gamma' \subset \Gamma_1}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[check-instr-branch-listcons]
+\end{equation}
+
+\begin{equation}
+\inference{\Gamma(v) = \text{nil}~~~\Pi(l) = \Gamma_1~~~\Gamma \subset \Gamma_1}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[check-instr-branch-nil]
+\end{equation}
+
+\begin{equation}
+\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}[check-instr-fetch-0]
+\end{equation}
+
+\begin{equation}
+\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\text{list}~\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}[check-instr-fetch-1]
+\end{equation}
+
+\begin{equation}
+\inference{\Gamma(v_0) = \tau_0~~~~~~~~~\Gamma(v_1) = \tau_1 \\ (\text{list}~\tau_0) \sqcap \tau_1=\text{list}~\tau~~~\Gamma[v:=\text{listcons}~\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{cons}~v_0~v_1~v \}\Gamma'}[check-instr-cons]
+\end{equation}
+
+%\subsubsection{Block typings}
+
+\begin{equation}
+\inference{}
+          {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[check-block-halt]
+\end{equation}
+
+\begin{equation}
+\inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'~~~\Pi;\Gamma'\vdash_{\text{block}} \iota_2}
+          {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[check-block-seq]
+\end{equation}
+
+\begin{equation}
+\inference{\Pi(l)=\Gamma_1~~~\Gamma \subset \Gamma_1}
+          {\Pi;\Gamma\vdash_{\text{block}} \textbf{jump}~l}[check-block-jump]
+\end{equation}
+
+%\subsubsection{Program typings}
+
+\begin{equation}
+\inference{\Pi(l)=\Gamma~~~\Pi;\Gamma\vdash_{\text{block}} \iota~~~\Pi\vdash_{\text{blocks}} p}
+          {\Pi\vdash_{\text{blocks}} l: \iota;~p}[check-blocks-label]
+\end{equation}
+
+\begin{equation}
+\inference{}
+          {\Pi\vdash_{\text{blocks}} \textbf{end}}[check-blocks-empty]
+\end{equation}
+
 \bibliographystyle{ACM-Reference-Format}

### 2020-06-08 papers/SBLP2020/paper.tex
Fixed some typos.

@@
 \begin{abstract}
-Formal models are an important tool in the programming languages research
+Formal models are important tools in the programming languages research
 community. However, such models are full of intricacies and, due to that,
@@
 such errors is to encode models in a dependently-typed language in order
-to ensure ``correctness-by-construction''.
+to ensure its ``correctness-by-construction''.
 
-In this work, we use follow this idea to build the verified interpreter
+In this work, we use this idea to build a verified interpreter
 for the list machine benchmark in the Agda programming language.

### 2020-06-08 papers/SBLP2020/paper.tex
Introduction finished

@@
 
-The development of new programming language design, linguistic construct
+The development of a new programming language design, linguistic construct
 or type system involves its careful formalization in order to express
 its core ideas in a concise way. However, such models have many details
-and complexities which hinders its correctness assurances.
+and complexities which can hinders its correctness assurances.
 Because of such problems, the programming languages research community
@@
 and benchmark problems to validate them and stress its suitability for
-such tasks~\cite{Aydemir05,Pientka18,Appel07}.
+such specification tasks~\cite{Aydemir05,Pientka18,Appel07}.
 
-The rest of this paper is organized as follows. Section~\ref{sec:background}
-presents a brief introduction to Agda and reviews the list machine benchmark.
-In Section~\ref{sec:typing}, we
-describe the type system for PEGs and its relation with the original
-well-formedness predicate proposed by Ford~\cite{Ford2004}.
-Section~\ref{sec:interpreter} describes the intrinsically typed syntax for PEGs
-and its interpreter. Limitations of our approach are discussed
-in Section~\ref{sec:discussion}. Related work is discussed in
-Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.
+A popular approach for specifying formal semantics is the
+use of definitional interpreters, which represents the meaning of a
+programming language as an interpreter written in some
+meta-language~\cite{Reynolds72}. A major advantage of such approach
+is the possibility of validating the semantics through execution.
+Recently, definitional interpreters were used to formalize type
+soundness theorems for some advanced typing features~\cite{Amin17}
+and the semantics of imperative programming languages in which static
+semantics is ensured by dependently-typed
+syntax\footnote{Also known as intrinsically-typed.}~\cite{Poulsen18}.
+In this work, we follow Poulsen et. al and use an intrinsically-typed
+representation to build a definitional interpreter for a low-level virtual
+machine developed by Appel and Leroy as a benchmark problem closer to
+real-world implementations like typed assembly languages~\cite{CraryM99} and
+proof carrying code~\cite{Necula97}.
+
+More specifically, we contribute:
+
+\begin{itemize}
+  \item We show how all the details of the list machine type system
+        can be encoded as dependently-typed syntax which avoids, by construction,
+        the presence of stuck states in its definitional interpreter.
+  \item We provide a provably correct implementations for testing the subtyping
+        relation and to calculate the least common super type of two input
+        types for the machine registers.
+\end{itemize}
+
+The rest of this paper is organized as follows. Section~\ref{sec:agda}
+presents a brief introduction to Agda and Section~\ref{sec:list}
+reviews the list machine benchmark and presents its syntax and type system.
+We describe the intrinsically-typed representation for the list machine syntax,
+the subtyping relation and the least common super type algorithm in
+Section~\ref{sec:typing}. The list machine semantics and its realization as
+a definitional interpreter are presented in Section~\ref{sec:semantics}.
+Related work is discussed in Section~\ref{sec:related},
+and Section~\ref{sec:conclusion} concludes.
 
 All the source code in this article has been formalized in Agda
-version 2.6.0 using Standard Library 1.0. All source code produced,
-including the \LaTeX~ source of this
-article, are available on-line~\cite{peg-rep}.
+version 2.6.1 using Standard Library 1.3. All source code produced,
+including the \LaTeX~source of this article, are available
+on-line~\cite{list-rep}.
 
-\section{Background}\label{sec:background}
+\section{An Overview of Agda}\label{sec:agda}
 
-\paragraph{An Overview of Agda}
 Agda is a dependently-typed functional programming language based on
@@
 
-\paragraph{The List Machine Benchmark}
+\section{The List Machine Benchmark}\label{sec:list}
 
@@
 
+\section{Instrinsically-typed syntax}\label{sec:typing}
+
+\section{A definitional interpreter}\label{sec:semantics}
+
+\section{Related work}\label{sec:related}
+
+\section{Conclusion}\label{sec:conclusion}
+
 \bibliographystyle{ACM-Reference-Format}

### 2020-06-08 papers/SBLP2020/paper.tex
Starting the section about the benchmark

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Title and authors %%
+%%%%%%%%%%%%%%%%%%%%%%%
+
 \title{A Certified Interpreter for the List Machine Benchmark}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Abstract          %%
+%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{abstract}
@@
 \end{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
+%% ACM codes for areas %%
+%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \maketitle
+
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Introduction      %%
+%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 on-line~\cite{list-rep}.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Overview of Agda  %%
+%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Another useful data type is the finite type, \ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of data type constructor names.
-Constructor \ensuremath{\Con{zero}} can refer to type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the
-context where the name is used.}, which is defined in
-Agda's standard library as:
-
+Another useful data type is the finite type,
+\ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of
+data type constructor names. Constructor \ensuremath{\Con{zero}} can refer to
+type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the context where the name
+is used.}, which is defined in Agda's standard library as:
 \begin{hscode}\SaveRestoreHook
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% List machine benchmark %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{The List Machine Benchmark}\label{sec:list}
+
+The list machine is a simple pointer virtual machine whose values
+are empty lists and cons-cells:
+\[
+a ::= nil\,\mid\,cons(a_1,a_2)
+\]
+We let meta-variable $a$ to denote an arbitrary value, $v$ denotes a variable
+and $l$ a program label. Following common practice, all meta-variables can
+appear primed or subscripted. The syntax of the virtual machine instructions are
+presented next and their meaning is as usual.
+\[
+\begin{array}{rcll}
+  i & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
+    & \mid & \text{branch-if-nil $v$}            & \text{(if $v = nil$ goto $l$)}\\
+    & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
+    & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
+    & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in v')} \\
+    & \mid & \text{halt}                         & \text{(finishes execution)}\\
+    & \mid & i_1;i_2                             & \text{(sequential composition)}\\
+  p & ::=  & l_i \,:\,i\,;\,p                    & \text{(program: sequence of blocks)}\\
+    & \mid & \text{end}                          & \text{(end of block list)}\\
+\end{array}
+\]
+Programs are just a sequence of blocks which begin with a unique label.
+
+Each program variable is assigned a list type which is used to guarantee the safety of execute
+fetch-field that demands non-empty list arguments and branch instructions refine the argument type
+according to which branch is taken. In order to express such refinements, types are subject to a
+subtyping relation. We let the meta-variable $\tau$ denote an arbitrary type. 
+\[
+\begin{array}{rcll}
+  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
+       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
+       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
+\end{array}
+\]
+We let notation $\tau \subset \tau'$
+denote the subtyping judgement which is defined as follows.
+\[
+\begin{array}{ccc}
+  \inference{}
+            {\tau\subset \tau}
+            [refl]
+  &
+  \inference{}
+            {nil \subset \tau}
+            [nil]
+  &
+  \inference{\tau \subset \tau'}
+            {list\:\tau\subset list\:\tau'}
+            [list]\\\\
+  \multicolumn{3}{c}{
+  \inference{\tau \subset \tau'}
+            {listcons\:\tau\subset list\:\tau'}
+            [listcons]} \\\\
+            \multicolumn{3}{c}{
+            \inference{\tau \subset \tau'}
+            {listcons\:\tau\subset listcons\:\tau'}
+            [mixed]} \\\\
+\end{array}
+\]
+Basically, the subtyping relation specifies that $nil$ (empty list type) is
+subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
 

### 2020-06-09 papers/SBLP2020/paper.tex
Small changes

@@
         can be encoded as dependently-typed syntax which avoids, by construction,
-        the presence of stuck states in its definitional interpreter.
+        the presence of stuck states in its definitional interpreter. In essence,
+        the defined interpreter is a type soundness theorem for the
+        list machine~\cite{Amin17}.
   \item We provide a provably correct implementations for testing the subtyping
         relation and to calculate the least common super type of two input
-        types for the machine registers.
+        types.
 \end{itemize}
@@
 %A definition which uses the expression |(Num 1) + True| will be rejected by Agda's type checker automatically.
-Agda's type checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}.
+Agda's type checker will automatically reject a definition which
+uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}.
 
@@
 subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
-
-\begin{equation}
-\inference{}
-          {(r, (\iota_1;\iota_2);\iota_3) \xmapsto{p} (r, \iota_1;(\iota_2;\iota_3))}[step-seq]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_0]=r'}
-          {(r, (\textbf{fetch-field}~v~0~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-0]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_1]=r'}
-          {(r, (\textbf{fetch-field}~v~1~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-1]
-\end{equation}
-
-\begin{equation}
-\inference{r(v_0) = a_0~~~r(v_1)=a_1~~~r[v':=cons(a_0,a_1)]=r'}
-          {(r, (\textbf{cons}~v_0~v_1~v';\iota)) \xmapsto{p} (r', \iota)}[step-cons]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = cons(a_0,a_1)}
-          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota)}[step-branch-not-taken]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = nil~~~p(l)=\iota'}
-          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota')}[step-branch-taken]
-\end{equation}
-
-\begin{equation}
-\inference{p(l)=\iota'}
-          {(r, \textbf{jump}~l) \xmapsto{p} (r, \iota')}[step-jump]
-\end{equation}
-
-\begin{equation}
-\inference{(r, \iota) \xmapsto{p} (r', \iota')~~~(p, r', \iota')}
-          {(p, r, \iota) \Downarrow}[run-step]
-\end{equation}
-
-\begin{equation}
-\inference{}
-          {(p, r, \textbf{halt}) \Downarrow}[run-halt]
-\end{equation}
-
-\begin{equation}
-\inference{\{\}[\textbf{v}_0:=\text{nil}]=r~~~p(\textbf{L}_0)=\iota~~~(p,r,\iota)\Downarrow}
-          {p \Downarrow}[run-prog]
-\end{equation}
-
-%\subsection{Typing}
-
-%\subsubsection{Subtyping}
-
-\begin{equation}
-\inference{}
-          {\tau \subset \tau}[subtype-refl]
-\end{equation}
-
-\begin{equation}
-\inference{}
-          {\text{nil} \subset \text{list}~\tau}[subtype-nil]
-\end{equation}
-
-\begin{equation}
-\inference{\tau \subset \tau'}
-          {\text{list}~\tau \subset \text{list}~\tau'}[subtype-list]
-\end{equation}
-
-\begin{equation}
-\inference{\tau \subset \tau'}
-          {\text{listcons}~\tau \subset \text{list}~\tau'}[subtype-listcons]
-\end{equation}
-
-\begin{equation}
-\inference{\tau \subset \tau'}
-          {\text{listcons}~\tau \subset \text{listcons}~\tau'}[subtype-listmixed]
-\end{equation}
+The other rules shows the that the $list$ and $listcons$ type constructors are
+compatible with subtyping relation.
 

### 2020-06-17 papers/SBLP2020/paper.tex
Some little updates

--- 2020-06-08 papers/SBLP2020/paper.tex
@@
 %\setcopyright{none}
-\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
+%\setcopyright{acmcopyright}
+\setcopyright{acmlicensed}
 %\setcopyright{rightsretained}
@@
 Formal models are important tools in the programming languages research
-community. However, such models are full of intricacies and, due to that,
+community. However, such models are full of intrincacies and, due to that,
 they are subject to subtle errors. Such failures motivated the usage of
@@
 subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
+The other rules specifies that type constructors $list$ and $listcons$ respect
+the subtyping relation.
+
+Using the previously defined type syntax and its subtyping relation, we can
+define the typing rules for the list machine. Following common practice, we
+let meta-variable $\Gamma$ denote an environment between variable names and
+its type. Notation $\{\}$ denote an empty environment and $v : \tau , \Gamma'$
+denote the operation of including a new entry for variable $v$ with type $\tau$.
+Subtyping is extended to contexts as follows.
+\[
+\begin{array}{cc}
+  \inference{}
+            {\Gamma \subset_{env} \{\}}[b1]
+  &
+  \inference{\Gamma_1 = (v : \tau') , \Gamma' & \tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
+            {\Gamma_1 \subset_{env} v : \tau , \Gamma_2}
+            [b2]
+\end{array}
+\]
+Variable $\Pi$ denote \emph{program typings}, i.e. finite mappings between labels and typing contexts $\Gamma$.
+Notation $\Pi(l) = \Gamma$ denotes that $\Gamma$ stores the types of variables on the entry point of the block
+labeled by $l$.
 
@@
 
-%\subsection{Typing}
-
-%\subsubsection{Subtyping}
-
-\begin{equation}
-\inference{}
-          {\tau \subset \tau}[subtype-refl]
-\end{equation}
-
-\begin{equation}
-\inference{}
-          {\text{nil} \subset \text{list}~\tau}[subtype-nil]
-\end{equation}
-
-\begin{equation}
-\inference{\tau \subset \tau'}
-          {\text{list}~\tau \subset \text{list}~\tau'}[subtype-list]
-\end{equation}
-
-\begin{equation}
-\inference{\tau \subset \tau'}
-          {\text{listcons}~\tau \subset \text{list}~\tau'}[subtype-listcons]
-\end{equation}
-
-\begin{equation}
-\inference{\tau \subset \tau'}
-          {\text{listcons}~\tau \subset \text{listcons}~\tau'}[subtype-listmixed]
-\end{equation}
-
 %\subsubsection{Instruction typing}
@@
 
+In this section we present the design choices and the steps we took to represent the intrinsically-typed syntax for the list-machine benchmark.
+We present here the Agda code used in our definitions, not necessarily in a strict lexically-scoped order.
+
+Some definitions and rules have been slightly tweaked so that they are accepted by the Agda's type-checker.
+As a design choice, we dropped all names, using \emph{de Bruijn} indices~\cite{DEBRUIJN72} to represent both \emph{name bindings} for labels and variables.
+This way, we guarantee that names are always well-scoped.
+
+We started our formalization by defining a type \ensuremath{\D{Ty}}, indicating the possible types for the list-machine language.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Ty}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{nil}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{list}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{listcons}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+We internalize the list-machine type judgments for blocks and instructions in Agda together with its syntax in such a way where only well-typed terms that satisfy typing judgments have meaning.
+This approach makes the AST contain both syntactic and semantic properties.
+
+To be well-typed, the list-machine syntax needs to refer to information from two sources: (1) a type context encoded as a list of types to store variable types; and (2) and a program context encoded as a vector\footnote{We use the \ensuremath{\D{Vec}} datatype indexed by a \ensuremath{\V{n}} which is bound on the module definition and represents the number of labels in the current program.} of type contexts to represent the types of the variables on entry to each basic block.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\D{Ctx}\;\mathbin{:}\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{Ctx}\;\mathrel{=}\;\D{List}\;\D{Ty}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\D{PCtx}\;\mathbin{:}\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{PCtx}\;\mathrel{=}\;\D{Vec}\;\D{Ctx}\;\V{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+As we saw in the previous section, the typing rules of the list-machine language were split in two segments, one for instructions and one for blocks.
+We defined two datatypes (\ensuremath{\D{\_\vdash\_\Rightarrow\_}} and \ensuremath{\D{Block}}) to hold the well-typed terms accordingly, representing each judgment of the static semantics as a syntactical constructor.
+In Agda we use \emph{indexed inductive types} to define a intrinsically-typed syntax.
+
+Both definitions are \emph{parameterized} by a program context and a typing context, and \emph{indexed}\footnote{An index can vary in the result types of the different constructors, while a parameter cannot.} by a resulting typing context.
+The intuition is that, under program-typing \ensuremath{\V{\Pi}}, the \emph{Hoare triple} \ensuremath{\V{\Gamma}\;\{\mskip1.5mu \V{\iota}\mskip1.5mu\}\;\V{\Gamma'}} relates precondition \ensuremath{\V{\Gamma}} to postcondition \ensuremath{\V{\Gamma'}}.
+It is important to note that instead of manipulating syntax directly, the meta-program manipulates structures representing the type judgments as well.
+Such representation scheme makes the Agda's type-checker allow only well-typed blocks and instructions to be created and manipulated.
+
+The representation of instructions is defined as follows.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{\_\vdash\_\Rightarrow\_}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx})\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}seq}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\;\V{\Gamma''}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma'}\;\D{\Rightarrow}\;\V{\Gamma''}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma''}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}branch\textrm{-}list}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{l}\;\V{\Gamma'}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\Con{list}\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma'}\;\rightarrow\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{nil}))\;\D{\subset}\;\V{\Gamma'}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}branch\textrm{-}listcons}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{l}\;\V{\Gamma_1}\;\V{x}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{nil}))\;\D{\subset}\;\V{\Gamma_1}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}branch\textrm{-}nil}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma_1}\;\V{l}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{nil})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;((\V{x'}\;\red{,}\,\;\V{\tau})\;\Con{::}\;\V{\Gamma}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau'}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x'}\;\red{,}\,\;\V{\tau})){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;((\V{x'}\;\red{,}\,\;\Con{list}\;\V{\tau})\;\Con{::}\;\V{\Gamma}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau'}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x'}\;\red{,}\,\;\Con{list}\;\V{\tau})){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}cons\textrm{-}new}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau_0}\;\V{\tau_1}\;\V{x₀}\;\V{x₁}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcap}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;((\V{x'}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\Con{::}\;\V{\Gamma}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}cons\textrm{-}upd}\;{}\<[26]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau_0}\;\V{\tau_1}\;\V{\tau_2}\;\V{x₀}\;\V{x₁}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau_2})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcap}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x'}\;\red{,}\,\;\Con{listcons}\;\V{\tau})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+In our approach, all name binding is done with statically checked \emph{de Bruijn} indices~\cite{DEBRUIJN72}, a technique for handling binding by using a nameless, position-dependent naming scheme.
+For example, we use a well-typed \emph{de Bruijn} index \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma}}, which witnesses the existence of an element \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})} in \ensuremath{\V{\Gamma}}, as defined by the standard library \ensuremath{\D{\_\hspace{-2pt}\in\hspace{-2pt}\_}} operator.
+This technique is well-known for avoiding out-of-bound errors.
+
+\paragraph{Sequencing instructions}{The constructor \ensuremath{\Con{instr\textrm{-}seq}} can be used to express a sequence of instructions.
+From the execution of two instructions, it produces a modified typing context containing the changes performed by both instructions.}
+
+\paragraph{Conditional jump}{There are three constructors related to a conditional jump.
+They are used to perform a jump to a label \ensuremath{\V{l}} when the received variable is \ensuremath{\Con{nil}}.
+All these constructors type-check the typing context of the intended label with the current typing context.
+We use \ensuremath{\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}}, meaning that there exist a typing context \ensuremath{\V{\Gamma_1}} in program typing \ensuremath{\V{\Pi}} related to label \ensuremath{\V{l}}.
+And we use \ensuremath{\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}} as a proof of subtyping between \ensuremath{\V{\Gamma}} and \ensuremath{\V{\Gamma_1}}.}
+
+\paragraph{Fetching information from list}{There are four constructors which can be used to fetch information from a given list.
+The constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}} receives a non-empty list (\ensuremath{\Con{listcons}}), and is used to retrieve the head of this list and store it in a fresh new variable.
+The resulting typing context adds the information about the new variable.
+Constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}} is also used to retrieve the head element of a list, however storing its value in an existing variable, represented by the \emph{de Bruijn} index \ensuremath{\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma}}.
+The constructors \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}} are similar, fetching the tail of a list instead of the head.}
+
+\paragraph{List construction}{The \ensuremath{\Con{instr\textrm{-}cons\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}cons\textrm{-}upd}} constructors are used to create a new list.
+The first creates a new variable, and the second updates a existing variable.
+The list is created from two variables, \ensuremath{(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}} and \ensuremath{(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}}, which are represented as \emph{de Bruijn} indices.
+The type of the new list is defined by the least common supertype\footnote{A complete explanation about the least common supertype can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the constructor \ensuremath{\D{\_\sqcap\_\sim\_}}\footnote{The code of this definition is omitted here, but can be found in our online repository.}.
+The resulting typing context adds information about the newly created list.}
+
+The representation of blocks is defined as follows.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{42}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Block}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx})\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;{}\<[42]%
+\>[42]{}\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{block\textrm{-}halt}\;{}\<[15]%
+\>[15]{}\mathbin{:}\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{block\textrm{-}seq}\;{}\<[15]%
+\>[15]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\;\V{\Gamma''}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;{}\<[E]%
+\\
+\>[15]{}\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma'}\;\V{\Gamma''}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma''}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{block\textrm{-}jump}\;{}\<[15]%
+\>[15]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\Gamma_1}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;{}\<[E]%
+\\
+\>[15]{}\rightarrow\;\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Constructor \ensuremath{\Con{block\textrm{-}halt}} can be used to stop the execution of a given block, \ensuremath{\Con{block\textrm{-}seq}} has a similar meaning to instruction sequence, and \ensuremath{\Con{block\textrm{-}jump}} is used to perform a direct jump (without any condition), receiving a label and checking if the current context typing is subtype of the intended one.
+
+And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label. We use the \ensuremath{\D{All}} datatype to express this relationship.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\D{Program}\;\mathbin{:}\;\D{PCtx}\;\rightarrow\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
 \section{A definitional interpreter}\label{sec:semantics}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Val}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{nil}\;\mathbin{:}\;\D{Val}\;\Con{nil}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{[]v}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\D{Val}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\D{Val}\;\V{t}\;\rightarrow\;\D{Val}\;(\Con{list}\;\V{t})\;\rightarrow\;\D{Val}\;(\Con{listcons}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{\_::v\_}\;\mathbin{:}\;{}\<[11]%
+\>[11]{}\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\D{Val}\;\V{t}\;\rightarrow\;\D{Val}\;(\Con{list}\;\V{t})\;\rightarrow\;\D{Val}\;(\Con{list}\;\V{t}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\D{Env}\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\D{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\D{PEnv}\;\mathbin{:}\;\D{PCtx}\;\rightarrow\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{PEnv}\;\V{\Pi}\;\mathrel{=}\;\D{Allv}\;\D{Env}\;\V{\Pi}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{run\textrm{-}step}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{Fuel}\;\rightarrow\;\D{Program}\;\V{\Pi}\;\rightarrow\;\D{Env}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[11]{}\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\;\rightarrow\;\D{Maybe}\;(\D{Env}\;\V{\Gamma'}){}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;\Con{zero}\;\V{p}\;\V{env}\;\V{b}\;\mathrel{=}\;\Con{nothing}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;\Con{block\textrm{-}halt}\;\mathrel{=}\;\Con{just}\;\V{env}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}seq}\;\V{i₁}\;\V{i₂})\;\V{b})\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;\V{i₁}\;(\Con{block\textrm{-}seq}\;\V{i₂}\;\V{b})){}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;\{\mskip1.5mu \V{\Pi}\mskip1.5mu\}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{\tau}\mskip1.5mu\}\;\{\mskip1.5mu \V{i}\mskip1.5mu\}\;\V{v}\;\V{l}\;\V{s})\;\V{b}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{[]v}\;\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{l})\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{\subset\hspace{-3pt}\textrm{-}Ctx}\;\V{s}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v}\;\Con{nil}))\;(\F{lookupA}\;\V{i}\;\V{p}){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\V{v₁}\;\Con{::\hspace{-2pt}v}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v}\;(\V{v₁}\;\Con{::}\;\V{v₂}))\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\V{v}\;\V{l}\;\V{s})\;\V{b})\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;\V{env}\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}branch\textrm{-}nil}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{i}\mskip1.5mu\}\;\V{v}\;\V{l}\;\V{s})\;\V{b})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{l})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{\subset\hspace{-3pt}\textrm{-}Ctx}\;\V{s}\;\V{env})\;(\F{lookupA}\;\V{i}\;\V{p}){}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{v})\;\V{b}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\V{v₁}\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{v}\;\V{v'})\;\V{b}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;\V{v₁})\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;\V{v})\;\V{b}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\V{v₂}\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;\V{v}\;\V{v'})\;\V{b}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;\V{v₂})\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\V{v₀}\;\V{v₁}\;\V{s})\;\V{b})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;((\F{\textrm{<:-}val}\;(\F{list\textrm{-<:-}inv}\;(\F{lub\textrm{-}subtype\textrm{-}left}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\F{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}cons\textrm{-}upd}\;\V{v₀}\;\V{v₁}\;\V{v'}\;\V{s})\;\V{b})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;(\F{\textrm{<:-}val}\;(\F{list\textrm{-<:-}inv}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\F{lub\textrm{-}subtype\textrm{-}left}\;(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s})))){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\F{lookup}\;\V{env}\;\V{v₀})\;\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\F{lookup}\;\V{env}\;\V{v₁})))\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}jump}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{i}\mskip1.5mu\}\;\V{l}\;\V{s})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{l})\;\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{\subset\hspace{-3pt}\textrm{-}Ctx}\;\V{s}\;\V{env})\;(\F{lookupA}\;\V{i}\;\V{p}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 

### 2020-06-17 papers/SBLP2020/paper.tex
Small updates

@@
   &
-  \inference{\Gamma_1 = (v : \tau') , \Gamma' & \tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
-            {\Gamma_1 \subset_{env} v : \tau , \Gamma_2}
+  \inference{\tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
+            {v : \tau' , \Gamma' \subset_{env} v : \tau , \Gamma_2}
             [b2]
@@
 \]
-Variable $\Pi$ denote \emph{program typings}, i.e. finite mappings between labels and typing contexts $\Gamma$.
-Notation $\Pi(l) = \Gamma$ denotes that $\Gamma$ stores the types of variables on the entry point of the block
+Variable $\Pi$ denote \emph{program typings}, i.e. finite mappings between labels and typing contexts $\Gamma$,
+where notation $\Pi(l) = \Gamma$ denotes that $\Gamma$ stores the types of variables on the entry point of the block
 labeled by $l$.

### 2020-06-18 papers/SBLP2020/paper.tex
Finished the list machine section

@@
         types for the machine registers.
+  \item We provide a correct by construction implementation of a type checker for
+        list machine programs. Our type checker produces, as result, an intrinsically-typed
+        representation of the machine code.
 \end{itemize}
@@
 \begin{array}{rcll}
-  i & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
+  \iota & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
         & \mid & \text{branch-if-nil $v$}            & \text{(if $v = nil$ goto $l$)}\\
@@
         & \mid & \text{halt}                         & \text{(finishes execution)}\\
-    & \mid & i_1;i_2                             & \text{(sequential composition)}\\
-  p & ::=  & l_i \,:\,i\,;\,p                    & \text{(program: sequence of blocks)}\\
+        & \mid & \iota_1;\iota_2                             & \text{(sequential composition)}\\
+      p & ::=  & l_i \,:\,\iota\,;\,p                    & \text{(program: sequence of blocks)}\\
         & \mid & \text{end}                          & \text{(end of block list)}\\
@@
 The other rules specifies that type constructors $list$ and $listcons$ respect
-the subtyping relation.
-
-Using the previously defined type syntax and its subtyping relation, we can
-define the typing rules for the list machine. Following common practice, we
-let meta-variable $\Gamma$ denote an environment between variable names and
-its type. Notation $\{\}$ denote an empty environment and $v : \tau , \Gamma'$
-denote the operation of including a new entry for variable $v$ with type $\tau$.
+the subtyping relation. The list common supertype $\tau = \tau_1 \sqcap \tau_2$ of
+$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
+are subtypes of $\tau$.
+
+Following common practice, we let meta-variable $\Gamma$ denote an
+environment between variable names and
+its types. Notation $\{\}$ denote an empty environment, $v : \tau , \Gamma$
+denote the operation of including a new entry for variable $v$ with type $\tau$
+in $\Gamma$ and $\Gamma [v := \tau]$ denotes the environment which is identical to $\Gamma$, except
+by the entry which associates variable $v$ with type $\tau$.
 Subtyping is extended to contexts as follows.
@@
 \]
-Variable $\Pi$ denote \emph{program typings}, i.e. finite mappings between labels and typing contexts $\Gamma$,
-where notation $\Pi(l) = \Gamma$ denotes that $\Gamma$ stores the types of variables on the entry point of the block
-labeled by $l$.
-
-\begin{equation}
+We let variable $\Pi$ denote \emph{program typings}, i.e. finite mappings between
+labels and typing contexts $\Gamma$, where notation $\Pi(l) = \Gamma$ denotes that
+$\Gamma$ stores the types of variables on the entry point of the block labeled by $l$.
+Using the previously defined notations, we can define the typing rules for the list
+machine as a syntax directed judgement $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$
+which intuitively means that the instruction $\iota$ transforms an input typing environment,
+$\Gamma$, into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
+rules for the list machine instructions are as follows.
+
+The first typing rule we consider is the one for sequencing instructions inside a block. Basically,
+the rule just threads the output environment from the first instruction as the input typing for the
+second.
+\[
+\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'~~~\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[seq]
+\]
+The type system proposed by Appel et. al.~\cite{Appel07} has three rules to deal with each of the possible
+types assigned to the branch variable in current typing context. The first rules deal with the $list$ and $listcon$
+types, which specifies that the environment associated to the label $l$, $\Pi(l) = \Gamma_1$, is greater than
+$\Gamma[v := nil]$. The third rule applies whenever $\Gamma(v) = nil$ and it also demands that $\Gamma \subset \Pi(l)$.
+
+\[
+\begin{array}{c}
+\inference{\Gamma(v) = \text{list}~\tau\:\:\:\:\:\:\:\:\Pi(l) = \Gamma_1\\
+                     \Gamma[v:=\text{nil}]=\Gamma'\:\:\:\:\:\:\:\:\Gamma' \subset_{env} \Gamma_1}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}(v: \text{listcons}~\tau,~\Gamma')}
+          [branch-list]
+\\ \\
+
+\inference{\Gamma(v) = \text{listcons}~\tau\:\:\:\:\:\:\:\:\Pi(l) = \Gamma_1\\
+           \Gamma[v:=\text{nil}]=\Gamma'\:\:\:\:\:\:\:\:\Gamma' \subset_{env} \Gamma_1}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[branch-listcons]
+
+\\ \\
+\inference{\Gamma(v) = \text{nil}~~~\Pi(l) = \Gamma_1~~~\Gamma \subset \Gamma_1}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}
+          [branch-nil]
+\end{array}
+\]
+Next, we have the \emph{fetch} instruction which can be used to store the head / tail of a list value in
+a variable. Rule \emph{fetch-0} retrieves the head of a value stored in a variable $v$ and \emph{fetch-1}
+does the same for the tail. Note that both rules demand that $\Gamma(v) = listcons\:\tau$, for some type
+$\tau$.
+\[
+\begin{array}{c}
+\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}[fetch-0]
+          \\ \\
+\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\text{list}~\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}[fetch-1]
+\end{array}
+\]
+The \emph{cons} instruction allows us to build a non-empty list value and this rule uses the least
+common supertype operator to check if the result of the operation is really a list type.
+\[
+\inference{\Gamma(v_0) = \tau_0\:\:\:\:\:\:\:\:\Gamma(v_1) = \tau_1 \\
+           (\text{list}~\tau_0) \sqcap \tau_1=\text{list}~\tau\:\:\:\:\:\:\:\:\Gamma[v:=\text{listcons}~\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{cons}~v_0~v_1~v \}\Gamma'}[cons]
+\]
+The final rules deal with the well-formedness of blocks and programs. The typing rules for \textbf{halt} instruction and
+program \textbf{end} are completely trivial. Rule \emph{block-seq} does the typing context threading between sequential
+instructions inside a block and rule \emph{block-label} recursively applies the judgment $\vdash_{block}$ on the
+input program.
+
+\[
+\begin{array}{cc}
    \inference{}
-          {(r, (\iota_1;\iota_2);\iota_3) \xmapsto{p} (r, \iota_1;(\iota_2;\iota_3))}[step-seq]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_0]=r'}
-          {(r, (\textbf{fetch-field}~v~0~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-0]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = cons(a_0,a_1)~~~r[v':=a_1]=r'}
-          {(r, (\textbf{fetch-field}~v~1~v';\iota)) \xmapsto{p} (r', \iota)}[step-fetch-field-1]
-\end{equation}
-
-\begin{equation}
-\inference{r(v_0) = a_0~~~r(v_1)=a_1~~~r[v':=cons(a_0,a_1)]=r'}
-          {(r, (\textbf{cons}~v_0~v_1~v';\iota)) \xmapsto{p} (r', \iota)}[step-cons]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = cons(a_0,a_1)}
-          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota)}[step-branch-not-taken]
-\end{equation}
-
-\begin{equation}
-\inference{r(v) = nil~~~p(l)=\iota'}
-          {(r, (\textbf{branch-if-nil}~v~l;\iota)) \xmapsto{p} (r, \iota')}[step-branch-taken]
-\end{equation}
-
-\begin{equation}
-\inference{p(l)=\iota'}
-          {(r, \textbf{jump}~l) \xmapsto{p} (r, \iota')}[step-jump]
-\end{equation}
-
-\begin{equation}
-\inference{(r, \iota) \xmapsto{p} (r', \iota')~~~(p, r', \iota')}
-          {(p, r, \iota) \Downarrow}[run-step]
-\end{equation}
-
-\begin{equation}
+             {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[halt]
+   &
+
+   \inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'\\ \Pi;\Gamma'\vdash_{\text{block}} \iota_2}
+             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[block-seq]\\ \\
+
+   \inference{\Pi(l)=\Gamma_1\\ \Gamma \subset_{env} \Gamma_1}
+             {\Pi;\Gamma\vdash_{\text{block}} \textbf{jump}~l}[jump]
+
+   &
    \inference{}
-          {(p, r, \textbf{halt}) \Downarrow}[run-halt]
-\end{equation}
-
-\begin{equation}
-\inference{\{\}[\textbf{v}_0:=\text{nil}]=r~~~p(\textbf{L}_0)=\iota~~~(p,r,\iota)\Downarrow}
-          {p \Downarrow}[run-prog]
-\end{equation}
-
-%\subsubsection{Instruction typing}
-
-\begin{equation}
-\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'~~~\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[check-instr-seq]
-\end{equation}
-
-\begin{equation}
-\inference{\Gamma(v) = \text{list}~\tau~~~\Pi(l) = \Gamma_1~~~\Gamma[v:=\text{nil}]=\Gamma'~~~\Gamma' \subset \Gamma_1}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}(v: \text{listcons}~\tau,~\Gamma')}[check-instr-branch-list]
-\end{equation}
-
-\begin{equation}
-\inference{\Gamma(v) = \text{listcons}~\tau~~~\Pi(l) = \Gamma_1~~~\Gamma[v:=\text{nil}]=\Gamma'~~~\Gamma' \subset \Gamma_1}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[check-instr-branch-listcons]
-\end{equation}
-
-\begin{equation}
-\inference{\Gamma(v) = \text{nil}~~~\Pi(l) = \Gamma_1~~~\Gamma \subset \Gamma_1}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[check-instr-branch-nil]
-\end{equation}
-
-\begin{equation}
-\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\tau]=\Gamma'}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}[check-instr-fetch-0]
-\end{equation}
-
-\begin{equation}
-\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\text{list}~\tau]=\Gamma'}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}[check-instr-fetch-1]
-\end{equation}
-
-\begin{equation}
-\inference{\Gamma(v_0) = \tau_0~~~~~~~~~\Gamma(v_1) = \tau_1 \\ (\text{list}~\tau_0) \sqcap \tau_1=\text{list}~\tau~~~\Gamma[v:=\text{listcons}~\tau]=\Gamma'}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{cons}~v_0~v_1~v \}\Gamma'}[check-instr-cons]
-\end{equation}
-
-%\subsubsection{Block typings}
-
-\begin{equation}
-\inference{}
-          {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[check-block-halt]
-\end{equation}
-
-\begin{equation}
-\inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'~~~\Pi;\Gamma'\vdash_{\text{block}} \iota_2}
-          {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[check-block-seq]
-\end{equation}
-
-\begin{equation}
-\inference{\Pi(l)=\Gamma_1~~~\Gamma \subset \Gamma_1}
-          {\Pi;\Gamma\vdash_{\text{block}} \textbf{jump}~l}[check-block-jump]
-\end{equation}
-
-%\subsubsection{Program typings}
-
-\begin{equation}
-\inference{\Pi(l)=\Gamma~~~\Pi;\Gamma\vdash_{\text{block}} \iota~~~\Pi\vdash_{\text{blocks}} p}
-          {\Pi\vdash_{\text{blocks}} l: \iota;~p}[check-blocks-label]
-\end{equation}
-
-\begin{equation}
-\inference{}
-          {\Pi\vdash_{\text{blocks}} \textbf{end}}[check-blocks-empty]
-\end{equation}
+             {\Pi\vdash_{\text{blocks}} \textbf{end}}[empty] \\ \\
+
+   \multicolumn{2}{c}{
+   \inference{\Pi(l)=\Gamma\:\:\:\:\:\:\:\:\Pi;\Gamma\vdash_{\text{block}} \iota\:\:\:\:\:\:\:\:\Pi\vdash_{\text{blocks}} p}
+     {\Pi\vdash_{\text{blocks}} l: \iota;~p}[blocks-label]
+    }
+\end{array}
+\]
+Inspired by the typing rules designed by Appel et. al., in the next section, we define an intrisinsically
+typed syntax for list machine programs which ensures that only well-typed programs can be built.
 
@@
 We use \ensuremath{\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}}, meaning that there exist a typing context \ensuremath{\V{\Gamma_1}} in program typing \ensuremath{\V{\Pi}} related to label \ensuremath{\V{l}}.
-And we use \ensuremath{\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}} as a proof of subtyping between \ensuremath{\V{\Gamma}} and \ensuremath{\V{\Gamma_1}}.}
+And we use \ensuremath{\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}} as a proof of subtyping between \ensuremath{\V{\Gamma}} and \ensuremath{\V{\Gamma_1}}.
+The operator \ensuremath{\F{\_\hspace{-2pt}::=\hspace{-2pt}\_}} is used to update the context \ensuremath{\V{\Gamma}} in the position defined by the index \ensuremath{\V{idx}}.}
 
@@
 
+This section describes the steps to evaluate a program written using the list-machine language.
+We adapted the small-step semantics presented in Section~\ref{sec:list}, transforming it in a definitional interpreter~\cite{Reynolds72}, which evaluates an intrinsically-typed instruction, transforming a initial memory state into a new one, represented by a run-time environment.
+
+\paragraph{Values and environments}{The interpreter presented next needs to encode a run-time environment to hold values associated to variables and their types.
+This way, we define the notion of a well-typed value as follows.}
+
 \begin{hscode}\SaveRestoreHook
@@
 
+The datatype \ensuremath{\D{Val}} is indexed by a \ensuremath{\D{Ty}}, indicating the type associated to each value.
+The first two represent \ensuremath{\Con{nil}} values associated with type \ensuremath{\Con{nil}} and the empty \ensuremath{\Con{list}}.
+The last two represent non-empty list, considering the types \ensuremath{\Con{listcons}} and \ensuremath{\Con{list}}.
+
+We use the datatype \ensuremath{\D{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed variable environments and well-typed programs.
+Thus, intuitively, \ensuremath{\D{Env}} is like a list of well-typed values.
+And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
+Both type environments are used to type block instructions and sequence of block instructions.
+
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
+
+\paragraph{Fuel based evaluation}{Having all the building blocks to make the complete interpreter for the list-machine language, we can start the definition of the \ensuremath{\F{run\textrm{-}step}} function.
+It is important to note that Agda is a total language, i.e., each program developed in it must terminate and all possible patterns must be matched.
+However, by using the mechanisms for jumping between labels one can write a program which never ends, making it impossible to implement a terminating interpreting function.
+Following the common practice, we define a fuel based evaluator~\cite{Amin17,Owens2016}.
+Basically, what we do is to parameterize the interpreter over a step index of \emph{fuel value} (represented as a natural number $n$), which bounds the amount of work the interpreter is allowed to do, and is decremented on each recursive call.}
+
+The evaluation function is defined with the following type.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{run\textrm{-}step}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{Fuel}\;\rightarrow\;\D{Program}\;\V{\Pi}\;\rightarrow\;\D{Env}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[11]{}\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\;\rightarrow\;\D{Maybe}\;(\D{Env}\;\V{\Gamma'}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+The function \ensuremath{\F{run\textrm{-}step}} receives four arguments and returns a \ensuremath{\D{Maybe}} value.
+The first argument is the \emph{fuel}, used to ensure the evaluator always terminates.
+The second parameter is a \ensuremath{\D{Program}\;\V{\Pi}}, which contains information about all the program blocks.
+The third parameter is the run-time variable environment.
+And the last one is the \ensuremath{\D{Block}} to be evaluated.
+This function returns a modified run-time environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or nothing when the \emph{fuel} runs out.
+
+From now on we describe how we implement some parts\footnote{The complete evaluation function can be found in our online repository.} of the dynamic semantics (reduction rules) of the list-machine language in the function \ensuremath{\F{run\textrm{-}step}}.
+We mix the code with the explanation to make it easier for the reader.
+
+\paragraph{Out of fuel}{The interpreter stops abruptly when the \emph{fuel} counter reaches \ensuremath{\Con{zero}}, and the \ensuremath{\F{run\textrm{-}step}} function returns \ensuremath{\Con{nothing}}.
+This definition makes our evaluation function structurally recursive on the \emph{fuel} argument.
+}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{run\textrm{-}step}\;\Con{zero}\;\V{p}\;\V{env}\;\V{b}\;\mathrel{=}\;\Con{nothing}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+All the next pieces of code match the value \ensuremath{\Con{suc}\;\V{fuel}} for the first argument of \ensuremath{\F{run\textrm{-}step}}, meaning that there is still \emph{fuel} during the recursive processing of this function.
+
+%run-step (suc n) p env block-halt = just env
+%run-step (suc n) p env (block-seq (instr-seq i₁ i₂) b) =
+%  run-step n p env (block-seq i₁ (block-seq i₂ b))
+%run-step {Π} (suc n) p env (block-seq (instr-branch-list {τ} {i} v l s) b)
+%  with lookup env v 
+%... | []v rewrite sym ([]=⇒lookup l) =
+%  run-step n p (⊂-Ctx s (update-env env v nil)) (lookupA i p)
+%... | v₁ ∷v v₂ = run-step n p (update-env env v (v₁ ∷ v₂)) b
+%run-step (suc n) p env (block-seq (instr-branch-listcons v l s) b) =
+% run-step n p env b
+
+\paragraph{Conditional jump}{We show next only the case when the jump actually occurs, following the rule \emph{step-branch-taken}.
+In this case, variable \ensuremath{\V{v}} has value \ensuremath{\Con{nil}}, and the step of evaluation should proceed with the block instruction defined in program \ensuremath{\V{p}}, with environment \ensuremath{\V{env}} respecting the subtyping constraint.
+We use the function \ensuremath{\F{lookupA}} to obtain the block instruction with index \ensuremath{\V{i}} on program \ensuremath{\V{p}}.
+Since we use \emph{de Bruijn} indices to represent the label, only valid values are accepted by the intrinsically-typed syntax.}
 
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{run\textrm{-}step}\;{}\<[11]%
-\>[11]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{Fuel}\;\rightarrow\;\D{Program}\;\V{\Pi}\;\rightarrow\;\D{Env}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[11]{}\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\;\rightarrow\;\D{Maybe}\;(\D{Env}\;\V{\Gamma'}){}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;\Con{zero}\;\V{p}\;\V{env}\;\V{b}\;\mathrel{=}\;\Con{nothing}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;\Con{block\textrm{-}halt}\;\mathrel{=}\;\Con{just}\;\V{env}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}seq}\;\V{i₁}\;\V{i₂})\;\V{b})\;\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;\V{i₁}\;(\Con{block\textrm{-}seq}\;\V{i₂}\;\V{b})){}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;\{\mskip1.5mu \V{\Pi}\mskip1.5mu\}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{\tau}\mskip1.5mu\}\;\{\mskip1.5mu \V{i}\mskip1.5mu\}\;\V{v}\;\V{l}\;\V{s})\;\V{b}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{[]v}\;\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{l})\;\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{\subset\hspace{-3pt}\textrm{-}Ctx}\;\V{s}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v}\;\Con{nil}))\;(\F{lookupA}\;\V{i}\;\V{p}){}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\V{v₁}\;\Con{::\hspace{-2pt}v}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v}\;(\V{v₁}\;\Con{::}\;\V{v₂}))\;\V{b}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\V{v}\;\V{l}\;\V{s})\;\V{b})\;\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;\V{env}\;\V{b}{}\<[E]%
-\\
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
 \>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}branch\textrm{-}nil}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{i}\mskip1.5mu\}\;\V{v}\;\V{l}\;\V{s})\;\V{b})\;{}\<[E]%
@@
 \>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{\subset\hspace{-3pt}\textrm{-}Ctx}\;\V{s}\;\V{env})\;(\F{lookupA}\;\V{i}\;\V{p}){}\<[E]%
-\\
+\ColumnHook
+\end{hscode}\resethooks
+
+\paragraph{Fetching list information}{The next code shows the evaluation of two syntactical constructors, both related to the \emph{step-fetch-field-0} rule.
+The first retrieves the head element of a list, and stores it in a new variable.
+The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time environment \ensuremath{\V{env}}, and this variable is added to the result environment.
+The typed \emph{de Bruijn} indices guarantee that the projected value has the type demanded, since the environment \ensuremath{\V{env}} is typed by the context \ensuremath{\V{\Gamma}}.
+Similarly, the second instruction also retrieves the head element of a list, however it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}.
+This process is done by the function \ensuremath{\F{update\textrm{-}env}}.}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
 \>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{v})\;\V{b}){}\<[E]%
@@
 \>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;\V{v₁})\;\V{b}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;\V{v})\;\V{b}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
-\\
-\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\V{v₂}\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;\V{v}\;\V{v'})\;\V{b}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
-\\
-\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;\V{v₂})\;\V{b}{}\<[E]%
-\\
+\ColumnHook
+\end{hscode}\resethooks
+
+%run-step (suc n) p env (block-seq (instr-fetch-1-new v) b)
+%  with lookup env v
+%...| v₁ ∷ v₂ = run-step n p (v₂ ∷ env) b
+%run-step (suc n) p env (block-seq (instr-fetch-1-upd v v') b)
+%  with lookup env v
+%...| v₁ ∷ v₂ = run-step n p (update-env env v' v₂) b
+
+\paragraph{List creation}{To evaluate the instruction which creates a new list and respect the expected types, we need some extra lemmas.
+First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the result type of this list is the least common subtype between these two.
+As before, we use the \ensuremath{\F{lookup}} function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we extend the run-time environment \ensuremath{\V{env}} with the type of the created list.
+To convince the Agda's type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}}, and others lemmas to deal with the least common subtying, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}}, and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}.
+These lemmas and their proofs can be found in our repository online.
+}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
 \>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\V{v₀}\;\V{v₁}\;\V{s})\;\V{b})\;{}\<[E]%
@@
 \>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}cons\textrm{-}upd}\;\V{v₀}\;\V{v₁}\;\V{v'}\;\V{s})\;\V{b})\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;(\F{\textrm{<:-}val}\;(\F{list\textrm{-<:-}inv}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}(\F{lub\textrm{-}subtype\textrm{-}left}\;(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s})))){}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}(\F{lookup}\;\V{env}\;\V{v₀})\;\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s}){}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}(\F{lookup}\;\V{env}\;\V{v₁})))\;\V{b}{}\<[E]%
-\\
-\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}jump}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{i}\mskip1.5mu\}\;\V{l}\;\V{s})\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{l})\;\mathrel{=}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{\subset\hspace{-3pt}\textrm{-}Ctx}\;\V{s}\;\V{env})\;(\F{lookupA}\;\V{i}\;\V{p}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
+\ColumnHook
+\end{hscode}\resethooks
+
+It is worth noticing that we do not have any error treatment on this interpreter function, except for when we ran out-of-fuel.
+Since we are using an intrinsically-typed syntax, only valid instructions are accepted in each step of evaluation.
+
+%run-step (suc n) p env (block-seq (instr-cons-upd v₀ v₁ v' s) b)
+%  = run-step n p (update-env env v' (<:-val (list-<:-inv
+%    (lub-subtype-left (lub-of-subtype (lub-subtype-left s))))
+%    (lookup env v₀) ∷ <:-val (lub-subtype-right s)
+%    (lookup env v₁))) b
+%run-step (suc n) p env (block-jump {l = i} l s)
+%  rewrite sym ([]=⇒lookup l) =
+%    run-step n p (⊂-Ctx s env) (lookupA i p)
+
+\paragraph{Soundness properties}{Programs written using an intrinsically-typed syntax are type-sound by construction.
+Since only well-typed programs can be expressed, the \emph{preservation} property is enforced by the host-language type-checker~\cite{Amin17}.
+By implementing the interpreter in such a total language like Agda, i.e., specifying the dynamic semantics in a functional way, instead of relational, we also show the \emph{progress} property, without the need for an extra proof~\cite{Owens2016}.
+This approach is promising to be investigated when formalizing even more complex programming languages.}
+
+\section{Type Checker}\label{sec:typechecker}
+
+In practice, a source-code of a programming language runs through several phases, including lexing, parsing, scope checking, and most importantly \emph{type checking}.
+This sections deals with scope and type checking.
+Since we represent programs using a intrinsically-typed syntax, scope and type checking is only a matter of elaborating an untyped syntax to a typed one.
+
+Since we use \emph{de Bruijn} indices to represent labels and variables, the first step to type and scope check them, we need to provide an index from a named variable.
+This process is done by the \ensuremath{\F{lookup\textrm{-}var}} function.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{lookup\textrm{-}var}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\rightarrow\;(\V{x}\;\mathbin{:}\;\D{String})\;{}\<[E]%
+\\
+\>[13]{}\rightarrow\;\D{Maybe}\;(\D{\exists}\;\lambda\;\V{\tau}\;\rightarrow\;(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma}){}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;\Con{\lbrack\:\rbrack}\;\V{x}\;\mathrel{=}\;\Con{nothing}{}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;((\V{y}\;\red{,}\,\;\V{\tau})\;\Con{::}\;\V{\Gamma})\;\V{x}\;\mathkw{with}\;\V{x}\;\D{=}\;\V{y}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{yes}\;\Con{refl}\;\mathrel{=}\;\Con{just}\;(\V{\tau}\;\red{,}\,\;\Con{here}\;\Con{refl}){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;(\Con{no}\;\V{¬p})\;\mathrel{=}\;\F{map}\;{}\<[22]%
+\>[22]{}(\lambda\;\V{v}\;\rightarrow\;\D{proj_1}\;\V{v}\;\red{,}\,\;\Con{there}\;(\D{proj_2}\;\V{v}))\;{}\<[E]%
+\\
+\>[22]{}(\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{x}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+For space reasons, we show how we type check only one instruction.
+Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}} receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value, which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type checks.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\mathbin{:}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx})\;\rightarrow\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\rightarrow\;(\V{v}\;\mathbin{:}\;\D{String})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\rightarrow\;(\V{v'}\;\mathbin{:}\;\D{String})\;\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{fetch\textrm{-}field\textrm{-}0}\;\V{v}\;\V{v'})){}\<[E]%
+\\
+\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{v'}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}\;\red{,}\,\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;(\text{\ttfamily \char34 variable~out~of~scope:~\char34}\;\F{++}\;\V{v}){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{nil}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error\char34}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{list}\;\V{\tau}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error\char34}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{listcons}\;\V{\tau}\;\red{,}\,\;\V{idx})\;\red{,}\,\;{}\<[34]%
+\>[34]{}\Con{nothing}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[34]{}\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{idx})){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{listcons}\;\V{\tau}\;\red{,}\,\;\V{idx})\;\red{,}\,\;{}\<[34]%
+\>[34]{}\Con{just}\;(\V{\tau'}\;\red{,}\,\;\V{idx'})\;\mathrel{=}\;{}\<[E]%
+\\
+\>[34]{}\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{idx}\;\V{idx'})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+In the code above, we use the function \ensuremath{\F{lookup\textrm{-}var}} to provide the \emph{de Bruijn} indices for each variable, and match the first with its possible forms.
+The first three cases indicate type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{nothing}} it means an scope error; (2) and (3) are typing errors, since the type of variable \ensuremath{\V{v}} should be a \ensuremath{\Con{listcons}}.
+Last two cases represent that the instruction is well-typed.
+The process for type-checking different instructions follows a similar setting.
 

### 2020-06-19 papers/SBLP2020/paper.tex
Some little formatting on text

@@
 
-In this section we present the design choices and the steps we took to represent the intrinsically-typed syntax for the list-machine benchmark.
-We present here the Agda code used in our definitions, not necessarily in a strict lexically-scoped order.
-
-Some definitions and rules have been slightly tweaked so that they are accepted by the Agda's type-checker.
-As a design choice, we dropped all names, using \emph{de Bruijn} indices~\cite{DEBRUIJN72} to represent both \emph{name bindings} for labels and variables.
-This way, we guarantee that names are always well-scoped.
-
-We started our formalization by defining a type \ensuremath{\D{Ty}}, indicating the possible types for the list-machine language.
+In this section we present the design choices and the steps we took to represent the
+intrinsically-typed syntax for the list-machine benchmark. We present here the Agda
+code used in our definitions, not necessarily in a strict lexically-scoped order.
+
+Some definitions and rules have been slightly tweaked so that they are accepted by
+the Agda's type-checker. As a design choice, we dropped all names, using
+\emph{de Bruijn} indices~\cite{DEBRUIJN72} to represent both \emph{name bindings}
+for labels and variables. This way, we guarantee that names are always well-scoped.
+
+We started our formalization by defining a type \ensuremath{\D{Ty}}, indicating the possible types
+for the list-machine language.
 
@@
 
-We internalize the list-machine type judgments for blocks and instructions in Agda together with its syntax in such a way where only well-typed terms that satisfy typing judgments have meaning.
-This approach makes the AST contain both syntactic and semantic properties.
-
-To be well-typed, the list-machine syntax needs to refer to information from two sources: (1) a type context encoded as a list of types to store variable types; and (2) and a program context encoded as a vector\footnote{We use the \ensuremath{\D{Vec}} datatype indexed by a \ensuremath{\V{n}} which is bound on the module definition and represents the number of labels in the current program.} of type contexts to represent the types of the variables on entry to each basic block.
+We internalize the list-machine type judgments for blocks and instructions in Agda
+together with its syntax in such a way where only well-typed terms that satisfy typing
+judgments have meaning. This approach makes the AST contain both syntactic and semantic
+properties.
+
+To be well-typed, the list-machine syntax needs to refer to information from two sources:
+(1) a type context encoded as a list of types to store variable types; and (2) and a program
+context encoded as a vector\footnote{We use the \ensuremath{\D{Vec}} datatype indexed by a \ensuremath{\V{n}} which is
+bound on the module definition and represents the number of labels in the current program.}
+of type contexts to represent the types of the variables on entry to each basic block.
 
@@
 
-As we saw in the previous section, the typing rules of the list-machine language were split in two segments, one for instructions and one for blocks.
-We defined two datatypes (\ensuremath{\D{\_\vdash\_\Rightarrow\_}} and \ensuremath{\D{Block}}) to hold the well-typed terms accordingly, representing each judgment of the static semantics as a syntactical constructor.
-In Agda we use \emph{indexed inductive types} to define a intrinsically-typed syntax.
-
-Both definitions are \emph{parameterized} by a program context and a typing context, and \emph{indexed}\footnote{An index can vary in the result types of the different constructors, while a parameter cannot.} by a resulting typing context.
-The intuition is that, under program-typing \ensuremath{\V{\Pi}}, the \emph{Hoare triple} \ensuremath{\V{\Gamma}\;\{\mskip1.5mu \V{\iota}\mskip1.5mu\}\;\V{\Gamma'}} relates precondition \ensuremath{\V{\Gamma}} to postcondition \ensuremath{\V{\Gamma'}}.
-It is important to note that instead of manipulating syntax directly, the meta-program manipulates structures representing the type judgments as well.
-Such representation scheme makes the Agda's type-checker allow only well-typed blocks and instructions to be created and manipulated.
+As we saw in the previous section, the typing rules of the list-machine language were split
+in two segments, one for instructions and one for blocks. We defined two datatypes (\ensuremath{\D{\_\vdash\_\Rightarrow\_}} and \ensuremath{\D{Block}})
+to hold the well-typed terms accordingly, representing each judgment of the static semantics as a
+syntactical constructor. In Agda we use \emph{indexed inductive types} to define a intrinsically-typed syntax.
+
+Both definitions are \emph{parameterized} by a program context and a typing context, and
+\emph{indexed}\footnote{An index can vary in the result types of the different constructors,
+while a parameter cannot.} by a resulting typing context. The intuition is that, under
+program-typing \ensuremath{\V{\Pi}}, the \emph{Hoare triple} \ensuremath{\V{\Gamma}\;\{\mskip1.5mu \V{\iota}\mskip1.5mu\}\;\V{\Gamma'}} relates precondition \ensuremath{\V{\Gamma}} to
+postcondition \ensuremath{\V{\Gamma'}}. It is important to note that instead of manipulating syntax directly,
+the meta-program manipulates structures representing the type judgments as well. Such
+representation scheme makes the Agda's type-checker allow only well-typed blocks and
+instructions to be created and manipulated.
 
@@
 
-In our approach, all name binding is done with statically checked \emph{de Bruijn} indices~\cite{DEBRUIJN72}, a technique for handling binding by using a nameless, position-dependent naming scheme.
-For example, we use a well-typed \emph{de Bruijn} index \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma}}, which witnesses the existence of an element \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})} in \ensuremath{\V{\Gamma}}, as defined by the standard library \ensuremath{\D{\_\hspace{-2pt}\in\hspace{-2pt}\_}} operator.
-This technique is well-known for avoiding out-of-bound errors.
-
-\paragraph{Sequencing instructions}{The constructor \ensuremath{\Con{instr\textrm{-}seq}} can be used to express a sequence of instructions.
-From the execution of two instructions, it produces a modified typing context containing the changes performed by both instructions.}
-
-\paragraph{Conditional jump}{There are three constructors related to a conditional jump.
-They are used to perform a jump to a label \ensuremath{\V{l}} when the received variable is \ensuremath{\Con{nil}}.
-All these constructors type-check the typing context of the intended label with the current typing context.
-We use \ensuremath{\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}}, meaning that there exist a typing context \ensuremath{\V{\Gamma_1}} in program typing \ensuremath{\V{\Pi}} related to label \ensuremath{\V{l}}.
-And we use \ensuremath{\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}} as a proof of subtyping between \ensuremath{\V{\Gamma}} and \ensuremath{\V{\Gamma_1}}.
-The operator \ensuremath{\F{\_\hspace{-2pt}::=\hspace{-2pt}\_}} is used to update the context \ensuremath{\V{\Gamma}} in the position defined by the index \ensuremath{\V{idx}}.}
-
-\paragraph{Fetching information from list}{There are four constructors which can be used to fetch information from a given list.
-The constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}} receives a non-empty list (\ensuremath{\Con{listcons}}), and is used to retrieve the head of this list and store it in a fresh new variable.
-The resulting typing context adds the information about the new variable.
-Constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}} is also used to retrieve the head element of a list, however storing its value in an existing variable, represented by the \emph{de Bruijn} index \ensuremath{\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma}}.
-The constructors \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}} are similar, fetching the tail of a list instead of the head.}
-
-\paragraph{List construction}{The \ensuremath{\Con{instr\textrm{-}cons\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}cons\textrm{-}upd}} constructors are used to create a new list.
-The first creates a new variable, and the second updates a existing variable.
-The list is created from two variables, \ensuremath{(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}} and \ensuremath{(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}}, which are represented as \emph{de Bruijn} indices.
-The type of the new list is defined by the least common supertype\footnote{A complete explanation about the least common supertype can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the constructor \ensuremath{\D{\_\sqcap\_\sim\_}}\footnote{The code of this definition is omitted here, but can be found in our online repository.}.
-The resulting typing context adds information about the newly created list.}
+In our approach, all name binding is done with statically checked \emph{de Bruijn}
+indices~\cite{DEBRUIJN72}, a technique for handling binding by using a nameless,
+position-dependent naming scheme. For example, we use a well-typed \emph{de Bruijn}
+index \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma}}, which witnesses the existence of an element \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})} in \ensuremath{\V{\Gamma}},
+as defined by the standard library \ensuremath{\D{\_\hspace{-2pt}\in\hspace{-2pt}\_}} operator. This technique is well-known for
+avoiding out-of-bound errors.
+
+\paragraph{Sequencing instructions}{The constructor \ensuremath{\Con{instr\textrm{-}seq}} can be used to
+express a sequence of instructions. From the execution of two instructions, it
+produces a modified typing context containing the changes performed by
+both instructions.}
+
+\paragraph{Conditional jump}{There are three constructors related to a conditional
+jump. They are used to perform a jump to a label \ensuremath{\V{l}} when the received variable is \ensuremath{\Con{nil}}.
+All these constructors type-check the typing context of the intended label with the
+current typing context. We use \ensuremath{\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}}, meaning that there exist a typing context
+\ensuremath{\V{\Gamma_1}} in program typing \ensuremath{\V{\Pi}} related to label \ensuremath{\V{l}}. And we use \ensuremath{\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}} as a proof of
+subtyping between \ensuremath{\V{\Gamma}} and \ensuremath{\V{\Gamma_1}}. The operator \ensuremath{\F{\_\hspace{-2pt}::=\hspace{-2pt}\_}} is used to update the context \ensuremath{\V{\Gamma}}
+in the position defined by the index \ensuremath{\V{idx}}.}
+
+\paragraph{Fetching information from list}{There are four constructors which can be used
+to fetch information from a given list. The constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}} receives a
+non-empty list (\ensuremath{\Con{listcons}}), and is used to retrieve the head of this list and store it
+in a fresh new variable. The resulting typing context adds the information about the new
+variable. Constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}} is also used to retrieve the head element of a
+list, however storing its value in an existing variable, represented by the \emph{de Bruijn}
+index \ensuremath{\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma}}.
+The constructors \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}}
+are similar, fetching the tail of a list instead of the head.}
+
+\paragraph{List construction}{The \ensuremath{\Con{instr\textrm{-}cons\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}cons\textrm{-}upd}} constructors are
+used to create a new list. The first creates a new variable, and the second updates a
+existing variable. The list is created from two variables, \ensuremath{(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}} and \ensuremath{(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}},
+which are represented as \emph{de Bruijn} indices. The type of the new list is defined by
+the least common supertype\footnote{A complete explanation about the least common supertype
+can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the
+constructor \ensuremath{\D{\_\sqcap\_\sim\_}}\footnote{The code of this definition is omitted here, but can be found
+in our online repository.}. The resulting typing context adds information about the
+newly created list.}
 
@@
 
-Constructor \ensuremath{\Con{block\textrm{-}halt}} can be used to stop the execution of a given block, \ensuremath{\Con{block\textrm{-}seq}} has a similar meaning to instruction sequence, and \ensuremath{\Con{block\textrm{-}jump}} is used to perform a direct jump (without any condition), receiving a label and checking if the current context typing is subtype of the intended one.
-
-And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label. We use the \ensuremath{\D{All}} datatype to express this relationship.
+Constructor \ensuremath{\Con{block\textrm{-}halt}} can be used to stop the execution of a given block, \ensuremath{\Con{block\textrm{-}seq}} has a
+similar meaning to instruction sequence, and \ensuremath{\Con{block\textrm{-}jump}} is used to perform a direct jump
+(without any condition), receiving a label and checking if the current context typing is
+subtype of the intended one.
+
+And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label.
+We use the \ensuremath{\D{All}} datatype to express this relationship.
 
@@
 This section describes the steps to evaluate a program written using the list-machine language.
-We adapted the small-step semantics presented in Section~\ref{sec:list}, transforming it in a definitional interpreter~\cite{Reynolds72}, which evaluates an intrinsically-typed instruction, transforming a initial memory state into a new one, represented by a run-time environment.
-
-\paragraph{Values and environments}{The interpreter presented next needs to encode a run-time environment to hold values associated to variables and their types.
-This way, we define the notion of a well-typed value as follows.}
+We adapted the small-step semantics presented in Section~\ref{sec:list}, transforming it in a
+definitional interpreter~\cite{Reynolds72}, which evaluates an intrinsically-typed instruction,
+transforming a initial memory state into a new one, represented by a run-time environment.
+
+\paragraph{Values and environments}{The interpreter presented next needs to encode a run-time
+environment to hold values associated to variables and their types. This way, we define the
+notion of a well-typed value as follows.}
 
@@
 
-The datatype \ensuremath{\D{Val}} is indexed by a \ensuremath{\D{Ty}}, indicating the type associated to each value.
-The first two represent \ensuremath{\Con{nil}} values associated with type \ensuremath{\Con{nil}} and the empty \ensuremath{\Con{list}}.
-The last two represent non-empty list, considering the types \ensuremath{\Con{listcons}} and \ensuremath{\Con{list}}.
-
-We use the datatype \ensuremath{\D{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed variable environments and well-typed programs.
-Thus, intuitively, \ensuremath{\D{Env}} is like a list of well-typed values.
-And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
-Both type environments are used to type block instructions and sequence of block instructions.
+The datatype \ensuremath{\D{Val}} is indexed by a \ensuremath{\D{Ty}}, indicating the type associated to each value. The
+first two represent \ensuremath{\Con{nil}} values associated with type \ensuremath{\Con{nil}} and the empty \ensuremath{\Con{list}}. The last
+two represent non-empty list, considering the types \ensuremath{\Con{listcons}} and \ensuremath{\Con{list}}.
+
+We use the datatype \ensuremath{\D{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
+variable environments and well-typed programs. Thus, intuitively, \ensuremath{\D{Env}} is like a
+list of well-typed values. And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
+Both type environments are used to type block instructions and sequence of
+block instructions.
 
@@
 
-\paragraph{Fuel based evaluation}{Having all the building blocks to make the complete interpreter for the list-machine language, we can start the definition of the \ensuremath{\F{run\textrm{-}step}} function.
-It is important to note that Agda is a total language, i.e., each program developed in it must terminate and all possible patterns must be matched.
-However, by using the mechanisms for jumping between labels one can write a program which never ends, making it impossible to implement a terminating interpreting function.
-Following the common practice, we define a fuel based evaluator~\cite{Amin17,Owens2016}.
-Basically, what we do is to parameterize the interpreter over a step index of \emph{fuel value} (represented as a natural number $n$), which bounds the amount of work the interpreter is allowed to do, and is decremented on each recursive call.}
+\paragraph{Fuel based evaluation}{Having all the building blocks to make the
+complete interpreter for the list-machine language, we can start the definition of
+the \ensuremath{\F{run\textrm{-}step}} function. It is important to note that Agda is a total language,
+i.e., each program developed in it must terminate and all possible patterns must
+be matched. However, by using the mechanisms for jumping between labels one can
+write a program which never ends, making it impossible to implement a terminating
+interpreting function. Following the common practice, we define a fuel based
+evaluator~\cite{Amin17,Owens2016}. Basically, what we do is to parameterize the
+interpreter over a step index of \emph{fuel value} (represented as a natural
+number $n$), which bounds the amount of work the interpreter is allowed to do,
+and is decremented on each recursive call.}
 
@@
 
-The function \ensuremath{\F{run\textrm{-}step}} receives four arguments and returns a \ensuremath{\D{Maybe}} value.
-The first argument is the \emph{fuel}, used to ensure the evaluator always terminates.
-The second parameter is a \ensuremath{\D{Program}\;\V{\Pi}}, which contains information about all the program blocks.
-The third parameter is the run-time variable environment.
-And the last one is the \ensuremath{\D{Block}} to be evaluated.
-This function returns a modified run-time environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or nothing when the \emph{fuel} runs out.
-
-From now on we describe how we implement some parts\footnote{The complete evaluation function can be found in our online repository.} of the dynamic semantics (reduction rules) of the list-machine language in the function \ensuremath{\F{run\textrm{-}step}}.
-We mix the code with the explanation to make it easier for the reader.
-
-\paragraph{Out of fuel}{The interpreter stops abruptly when the \emph{fuel} counter reaches \ensuremath{\Con{zero}}, and the \ensuremath{\F{run\textrm{-}step}} function returns \ensuremath{\Con{nothing}}.
-This definition makes our evaluation function structurally recursive on the \emph{fuel} argument.
-}
+The function \ensuremath{\F{run\textrm{-}step}} receives four arguments and returns a \ensuremath{\D{Maybe}} value. The
+first argument is the \emph{fuel}, used to ensure the evaluator always terminates.
+The second parameter is a \ensuremath{\D{Program}\;\V{\Pi}}, which contains information about all the
+program blocks. The third parameter is the run-time variable environment. And the
+last one is the \ensuremath{\D{Block}} to be evaluated. This function returns a modified run-time
+environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or nothing when the \emph{fuel} runs out.
+
+From now on we describe how we implement some parts\footnote{The complete evaluation
+function can be found in our online repository.} of the dynamic semantics (reduction rules)
+of the list-machine language in the function \ensuremath{\F{run\textrm{-}step}}. We mix the code with the explanation
+to make it easier for the reader.
+
+\paragraph{Out of fuel}{The interpreter stops abruptly when the \emph{fuel} counter
+reaches \ensuremath{\Con{zero}}, and the \ensuremath{\F{run\textrm{-}step}} function returns \ensuremath{\Con{nothing}}. This definition makes our
+evaluation function structurally recursive on the \emph{fuel} argument. }
 
@@
 
-All the next pieces of code match the value \ensuremath{\Con{suc}\;\V{fuel}} for the first argument of \ensuremath{\F{run\textrm{-}step}}, meaning that there is still \emph{fuel} during the recursive processing of this function.
+All the next pieces of code match the value \ensuremath{\Con{suc}\;\V{fuel}} for the first argument of \ensuremath{\F{run\textrm{-}step}},
+meaning that there is still \emph{fuel} during the recursive processing of this function.
 
@@
 
-\paragraph{Conditional jump}{We show next only the case when the jump actually occurs, following the rule \emph{step-branch-taken}.
-In this case, variable \ensuremath{\V{v}} has value \ensuremath{\Con{nil}}, and the step of evaluation should proceed with the block instruction defined in program \ensuremath{\V{p}}, with environment \ensuremath{\V{env}} respecting the subtyping constraint.
-We use the function \ensuremath{\F{lookupA}} to obtain the block instruction with index \ensuremath{\V{i}} on program \ensuremath{\V{p}}.
-Since we use \emph{de Bruijn} indices to represent the label, only valid values are accepted by the intrinsically-typed syntax.}
+\paragraph{Conditional jump}{We show next only the case when the jump actually occurs,
+following the rule \emph{step-branch-taken}. In this case, variable \ensuremath{\V{v}} has value \ensuremath{\Con{nil}}, and
+the step of evaluation should proceed with the block instruction defined in program \ensuremath{\V{p}}, with
+environment \ensuremath{\V{env}} respecting the subtyping constraint. We use the function \ensuremath{\F{lookupA}} to obtain
+the block instruction with index \ensuremath{\V{i}} on program \ensuremath{\V{p}}. Since we use \emph{de Bruijn} indices to
+represent the label, only valid values are accepted by the intrinsically-typed syntax.}
 
@@
 
-\paragraph{Fetching list information}{The next code shows the evaluation of two syntactical constructors, both related to the \emph{step-fetch-field-0} rule.
-The first retrieves the head element of a list, and stores it in a new variable.
-The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time environment \ensuremath{\V{env}}, and this variable is added to the result environment.
-The typed \emph{de Bruijn} indices guarantee that the projected value has the type demanded, since the environment \ensuremath{\V{env}} is typed by the context \ensuremath{\V{\Gamma}}.
-Similarly, the second instruction also retrieves the head element of a list, however it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}.
-This process is done by the function \ensuremath{\F{update\textrm{-}env}}.}
+\paragraph{Fetching list information}{The next code shows the evaluation of two syntactical constructors,
+both related to the \emph{step-fetch-field-0} rule. The first retrieves the head element of a list, and
+stores it in a new variable. The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
+environment \ensuremath{\V{env}}, and this variable is added to the result environment. The typed \emph{de Bruijn}
+indices guarantee that the projected value has the type demanded, since the environment \ensuremath{\V{env}} is typed
+by the context \ensuremath{\V{\Gamma}}. Similarly, the second instruction also retrieves the head element of a list, however
+it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}. This process is done by the
+function \ensuremath{\F{update\textrm{-}env}}.}
 
@@
 
-\paragraph{List creation}{To evaluate the instruction which creates a new list and respect the expected types, we need some extra lemmas.
-First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the result type of this list is the least common subtype between these two.
-As before, we use the \ensuremath{\F{lookup}} function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we extend the run-time environment \ensuremath{\V{env}} with the type of the created list.
-To convince the Agda's type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}}, and others lemmas to deal with the least common subtying, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}}, and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}.
-These lemmas and their proofs can be found in our repository online.
-}
+\paragraph{List creation}{To evaluate the instruction which creates a new list and respect the expected
+types, we need some extra lemmas. First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the
+result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\F{lookup}}
+function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we
+extend the run-time environment \ensuremath{\V{env}} with the type of the created list. To convince the Agda's
+type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}},
+and others lemmas to deal with the least common subtying, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
+and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}. These lemmas and their proofs can be found in our repository online. }
 
@@
 
-It is worth noticing that we do not have any error treatment on this interpreter function, except for when we ran out-of-fuel.
-Since we are using an intrinsically-typed syntax, only valid instructions are accepted in each step of evaluation.
+It is worth noticing that we do not have any error treatment on this interpreter function, except for
+when we ran out-of-fuel. Since we are using an intrinsically-typed syntax, only valid instructions
+are accepted in each step of evaluation.
 
@@
 
-\paragraph{Soundness properties}{Programs written using an intrinsically-typed syntax are type-sound by construction.
-Since only well-typed programs can be expressed, the \emph{preservation} property is enforced by the host-language type-checker~\cite{Amin17}.
-By implementing the interpreter in such a total language like Agda, i.e., specifying the dynamic semantics in a functional way, instead of relational, we also show the \emph{progress} property, without the need for an extra proof~\cite{Owens2016}.
+\paragraph{Soundness properties}{Programs written using an intrinsically-typed syntax are type-sound by
+construction. Since only well-typed programs can be expressed, the \emph{preservation} property is
+enforced by the host-language type-checker~\cite{Amin17}. By implementing the interpreter in such a
+total language like Agda, i.e., specifying the dynamic semantics in a functional way, instead of
+relational, we also show the \emph{progress} property, without the need for an extra proof~\cite{Owens2016}.
 This approach is promising to be investigated when formalizing even more complex programming languages.}
@@
 
-In practice, a source-code of a programming language runs through several phases, including lexing, parsing, scope checking, and most importantly \emph{type checking}.
-This sections deals with scope and type checking.
-Since we represent programs using a intrinsically-typed syntax, scope and type checking is only a matter of elaborating an untyped syntax to a typed one.
-
-Since we use \emph{de Bruijn} indices to represent labels and variables, the first step to type and scope check them, we need to provide an index from a named variable.
-This process is done by the \ensuremath{\F{lookup\textrm{-}var}} function.
+In practice, a source-code of a programming language runs through several phases, including lexing, parsing,
+scope checking, and most importantly \emph{type checking}. This sections deals with scope and type checking.
+Since we represent programs using a intrinsically-typed syntax, scope and type checking is only a matter of
+elaborating an untyped syntax to a typed one.
+
+Since we use \emph{de Bruijn} indices to represent labels and variables, the first step to type and scope
+check them, we need to provide an index from a named variable. This process is done by the \ensuremath{\F{lookup\textrm{-}var}} function.
 
@@
 
-For space reasons, we show how we type check only one instruction.
-Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}} receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value, which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type checks.
+For space reasons, we show how we type check only one instruction. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
+receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value,
+which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type checks.
 
@@
 
-In the code above, we use the function \ensuremath{\F{lookup\textrm{-}var}} to provide the \emph{de Bruijn} indices for each variable, and match the first with its possible forms.
-The first three cases indicate type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{nothing}} it means an scope error; (2) and (3) are typing errors, since the type of variable \ensuremath{\V{v}} should be a \ensuremath{\Con{listcons}}.
-Last two cases represent that the instruction is well-typed.
-The process for type-checking different instructions follows a similar setting.
+In the code above, we use the function \ensuremath{\F{lookup\textrm{-}var}} to provide the \emph{de Bruijn} indices for
+each variable, and match the first with its possible forms. The first three cases indicate
+type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{nothing}} it means an scope error; (2) and (3) are typing errors,
+since the type of variable \ensuremath{\V{v}} should be a \ensuremath{\Con{listcons}}. Last two cases represent that the instruction
+is well-typed. The process for type-checking different instructions follows a similar setting.
+
+\section{Least common supertype}
 

### 2020-06-20 papers/SBLP2020/paper.tex
Added text for subtyping and lub

@@
 \[
-\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'~~~\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
+\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'\:\:\:\:\:\:\:\:\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
           {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[seq]
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{nil}\;{}\<[13]%
-\>[13]{}\mathbin{:}\;\D{Ty}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{list}\;{}\<[13]%
-\>[13]{}\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{listcons}\;{}\<[13]%
-\>[13]{}\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
+\>[3]{}\Con{nil}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{list}\;\Con{listcons}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
 \ColumnHook
@@
 
-To be well-typed, the list-machine syntax needs to refer to information from two sources:
-(1) a type context encoded as a list of types to store variable types; and (2) and a program
-context encoded as a vector\footnote{We use the \ensuremath{\D{Vec}} datatype indexed by a \ensuremath{\V{n}} which is
-bound on the module definition and represents the number of labels in the current program.}
-of type contexts to represent the types of the variables on entry to each basic block.
+In order to be considered well-typed, list-machine machine programs needs to refer
+to information from two sources: (1) a type context encoded as a list of types to store
+variable types; and (2) and a program context encoded as a vector\footnote{We use the
+\ensuremath{\D{Vec}} datatype indexed by a \ensuremath{\V{n}} which is bound on the module definition and represents
+the number of labels in the current program.} of type contexts to represent the types of
+the variables on entry to each basic block.
 
@@
 syntactical constructor. In Agda we use \emph{indexed inductive types} to define a intrinsically-typed syntax.
-
-Both definitions are \emph{parameterized} by a program context and a typing context, and
-\emph{indexed}\footnote{An index can vary in the result types of the different constructors,
-while a parameter cannot.} by a resulting typing context. The intuition is that, under
-program-typing \ensuremath{\V{\Pi}}, the \emph{Hoare triple} \ensuremath{\V{\Gamma}\;\{\mskip1.5mu \V{\iota}\mskip1.5mu\}\;\V{\Gamma'}} relates precondition \ensuremath{\V{\Gamma}} to
-postcondition \ensuremath{\V{\Gamma'}}. It is important to note that instead of manipulating syntax directly,
-the meta-program manipulates structures representing the type judgments as well. Such
-representation scheme makes the Agda's type-checker allow only well-typed blocks and
+The basic idea is to represent each type system rule as a constructor of the underlying type.
+Such representation scheme makes the Agda's type-checker allow only well-typed blocks and
 instructions to be created and manipulated.
+
+%Both definitions are \emph{parameterized} by a program context and a typing context, and
+%\emph{indexed}\footnote{An index can vary in the result types of the different constructors,
+%while a parameter cannot.} by a resulting typing context. The intuition is that, under
+%program-typing |Π|, the \emph{Hoare triple} |Γ{ι}Γ'| relates precondition |Γ| to
+%postcondition |Γ'|. It is important to note that instead of manipulating syntax directly,
+%the meta-program manipulates structures representing the type judgments as well.
+%Such
+%representation scheme makes the Agda's type-checker allow only well-typed blocks and
+%instructions to be created and manipulated.
 
@@
 can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the
-constructor \ensuremath{\D{\_\sqcap\_\sim\_}}\footnote{The code of this definition is omitted here, but can be found
-in our online repository.}. The resulting typing context adds information about the
-newly created list.}
+type \ensuremath{\V{\tau_1}\;\D{\sqcap}\;\V{\tau_2}\;\D{\sim}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
+The resulting typing context adds information about the newly created list.
+Section~\ref{sec:supertype} will provide details of an algorithm
+for calculating the least common supertype of two given types.}
+
+%\footnote{The code of this definition is omitted here, but can be found
+%in our online repository.}
 
@@
 In practice, a source-code of a programming language runs through several phases, including lexing, parsing,
-scope checking, and most importantly \emph{type checking}. This sections deals with scope and type checking.
-Since we represent programs using a intrinsically-typed syntax, scope and type checking is only a matter of
-elaborating an untyped syntax to a typed one.
+scope checking, and most importantly \emph{type checking}. Since we represent programs using a
+intrinsically-typed syntax, scope and type checking is only a matter of elaborating an untyped syntax
+to a typed one.
 
 Since we use \emph{de Bruijn} indices to represent labels and variables, the first step to type and scope
-check them, we need to provide an index from a named variable. This process is done by the \ensuremath{\F{lookup\textrm{-}var}} function.
+check them, we need to provide an index from a named variable. This process is done by the
+\ensuremath{\F{lookup\textrm{-}var}} function.
 
@@
 
-\section{Least common supertype}
+\subsection{Subtyping and least common supertype}\label{sec:supertype}
+
+A key feature of the list machine type system is its subypting which is exposed by the
+the least common supertype relation. We omit the Agda encoding of the subtyping relation for
+space reasons, since it is just an inductive type that encode the rules presented
+in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
+subtyping including its decidability.
+
+The least common supertype relation is used by the least machine type system to refine a
+variable type whenever it is updated by a \emph{cons} instruction. The rules of the
+super type relation are specified as the following inductive type:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{\_\sqcap\_\sim\_}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\\
+\>[3]{}\hsindent{10}{}\<[13]%
+\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcap}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\\
+\>[3]{}\hsindent{10}{}\<[13]%
+\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcap}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
+types. Also, rules ensure that the relation is commutative as proved by a simple
+Agda theorem.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{\sqcap-comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We omit the complete definition of \ensuremath{\F{\sqcap-comm}} for brevity. In the list machine benchmark definition~\cite{Appel07},
+Appel et. al. also defines that the least common supertype relation is sound and complete with respect to the
+subtyping relation. Again, we omit the definitions of these properties, whose type is presented below.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{\sqcap-subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
+\\
+\>[B]{}\F{\sqcap-least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+An algorithm to construct the least supertype from two input types is
+defined by function \ensuremath{\F{lub}}. Note that the \ensuremath{\F{lub}}'s type ensures that the
+returned type \ensuremath{\V{t}} is indeed the least common supertype of \ensuremath{\V{t1}} and \ensuremath{\V{t2}},
+thus ensuring its correctness by construction.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\D{\exists}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\F{lub}\;\Con{nil}\;\Con{nil}\;\mathrel{=}\;\Con{nil}\;\red{,}\,\;\Con{lub0}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{list}\;\V{t1})\;\Con{nil}\;\mathrel{=}\;\Con{list}\;\V{t1}\;\red{,}\,\;\Con{lub1}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{list}\;\V{t1})\;(\Con{list}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{list}\;\V{t3}\;\red{,}\,\;\Con{lub3}\;\V{p}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;\Con{nil}\;\mathrel{=}\;\Con{list}\;\V{t1}\;\red{,}\,\;\Con{lub5}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;(\Con{list}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{list}\;\V{t3}\;\red{,}\,\;\Con{lub2b}\;(\Con{lub3}\;\V{p}){}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;(\Con{listcons}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{listcons}\;\V{t3}\;\red{,}\,\;\Con{lub7}\;\V{p}{}\<[E]%
+\\
+\>[B]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\section{Comparison of Mechanized Proofs}
+
+We implemented 15 tasks from the list machine benchmark in the Agda programming language.
+The total number of lines for each task is summarized in the next table.
+
+\begin{table}[!htb]
+\begin{tabular}{rl|rrr}
+    & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
+1.  & Operational Semantics                        & 126                       & 98                      & 97                       \\
+2.  & Derive $p \Downarrow$                        & 1                         & 8                       &                          \\ \hline
+3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
+4.  & $\sqcap$ algorithm                           & *                         & *                       & 13                       \\
+5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
+                                                   & 1                         & no                      & 1                        \\
+6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
+7.  & Prove properties of $\sqcap$                 & 114                       & 21                      & 124                      \\
+8.  & State soundness theorem                      & 29                        & 15                      & *                        \\
+9.  & Prove soundness of $\vDash_{\textrm{prog}} p : \Pi$
+                                                   & 2060                      & 315                     & *                        \\ \hline
+10. & Efficient algorithm                          & 22                        & 145                     & 98                       \\
+11. & Derive $\vdash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
+                                                   & 1                         & 1                       & 1                        \\
+12. & Prove termination of $\vdash_{\textrm{prog}} p : \Pi$
+                                                   & 18                        & 0                       & 0                        \\
+13. & Scalable type-checker                        & yes                       & yes                     & yes                      \\
+14. & Prove soundness of $\vdash_{\textrm{prog}} p : \Pi$
+                                                   & 347                       & 141                     & * 
+\end{tabular}
+\end{table}
 

### 2020-06-20 papers/SBLP2020/paper.tex
More changes on text. Infrastructure to measure compilation time.

@@
 
-We implemented 15 tasks from the list machine benchmark in the Agda programming language.
+We implemented 14 tasks from the list machine benchmark in the Agda programming language.
 The total number of lines for each task is summarized in the next table.
@@
 
+The total time for parsing and proof checking our Agda implementation was around 10 seconds 
+on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5.
+
 \section{Related work}\label{sec:related}

### 2020-06-21 papers/SBLP2020/paper.tex
Finished a first version of comparison section.

@@
 
-\title{A Certified Interpreter for the List Machine Benchmark}
+\title{An Intrinsically Typed Solution for the List Machine Benchmark}
 
@@
 We implemented 14 tasks from the list machine benchmark in the Agda programming language.
-The total number of lines for each task is summarized in the next table.
+The selected tasks considered by us are the ones done by Appel et. al.~\cite{Appel07}.
+The next table summarize the total number of lines for our results together with Appel's.
 
@@
     & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
-1.  & Operational Semantics                        & 126                       & 98                      & 97                       \\
-2.  & Derive $p \Downarrow$                        & 1                         & 8                       &                          \\ \hline
+1.  & Operational Semantics                        & 126                       & 98                      & 109                      \\
+2.  & Derive $p \Downarrow$                        & 1                         & 8                       & -                        \\ \hline
 3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
@@
 5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
-                                                   & 1                         & no                      & 1                        \\
+                                                   & 1                         & no                      & -                        \\
 6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
@@
 The total time for parsing and proof checking our Agda implementation was around 10 seconds 
-on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5.
+on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5. We briefly comment 
+on our Agda coding of these 14 tasks.
+
+\begin{enumerate}
+  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we choose to
+        use a definitional interpreter for our intrinsincally-typed representation. Our implementation for the operational
+        semantics is composed by 38 lines for the type syntax and 71 for the definitional interpreter.
+  \item \textbf{Derive $p\,\Downarrow$.} Since we used an definitional intepreter for representing the semantics, we can derive
+        $p\,\Downarrow$ just by executing the interpreter on $p$.
+  \item \textbf{Represent the type system.} Our type system representation consists of the intrinsically-typed encoding of
+        the list machine programs and it is already counted in as part of the operational semantics of our solution.
+  \item \textbf{Least common supertype algorithm.} In order to implement this task, we specified the least common super type
+        as a relation and implemented the algorithm which shows that this relation is indeed a function. In this way, we
+        guarantee its soundness w.r.t. its specification.
+  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list machine program
+        is just execute the type checker over it. Since our type checker returns an intrinsincally-typed representation of
+        the input and such typed syntax is equivalent to the type system, it can be used to build type derivations.
+  \item \textbf{State properties of the least common supertype.} This is trivial, since it is just a matter to encode the
+        desired properties as Agda function types.
+  \item \textbf{Prove properties of the least common supertype.} Proofs of all proved properties about the least common supertype
+        are simple recursive functions over the least common supertype and subtyping relations definition.
+  \item \textbf{State the soundness theorem for the type system.} Following Amin et. al.~\cite{Amin17}, we represent the soundness
+        theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list machine
+        programs.
+  \item \textbf{Prove the soundness theorem for the type system.} Following Amin et. al.~\cite{Amin17}, the proof of the soundness
+        theorem is just the implementation of the definitional interpreter. Soundness is ensured by construction since the interpreter
+        produces, as result, an environment of well-typed values resulting from the execution of the input program.
+  \item \textbf{Asymptotically efficient algorithm.} Our current implementation of the typechecker takes quadratic time. The reason
+        for this inneficiency is the representation of environments as lists / vectors. The use of better data structures (like finite
+        maps implemented by efficient search trees) can solve this issue. We leave this fix to future work.
+  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it ammounts to calling the typechecker
+        on the input program.
+  \item \textbf{Prove the termination of the type checker.} This task is trivial in our setting, since all defined Agda functions
+        must be total. The totality is ensured by Agda's termination / totality checker.
+  \item \textbf{Scalable type-checker.} Agda code can be compiled to machine code using its GHC-Haskell backend. Since GHC is
+        an industrial strength compiler, the backend can generate an efficient executable for the machine interpreter and typechecker. 
+  \item \textbf{Prove soundness of type-checker.} In our approach, the soundness of the typechecker is ensured by construction,
+        since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
+\end{enumerate}
 
 \section{Related work}\label{sec:related}
+
+\paragraph{Benchmarks for PL mechanization}{
+POPLMark chalenge~\cite{Aydemir05} .... Another chalenge was proposed by Pientka et. al.~\cite{Pientka18}...
+}
+
+\paragraph{Definitional intepreters}{
+The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
+}
 

### 2020-06-23 papers/SBLP2020/paper.tex
Starting the related work

@@
 syntax\footnote{Also known as intrinsically-typed.}~\cite{Poulsen18}.
-In this work, we follow Poulsen et. al and use an intrinsically-typed
+In this work, we follow Poulsen et. al. and use an intrinsically-typed
 representation to build a definitional interpreter for a low-level virtual
@@
 
-A key feature of the list machine type system is its subypting which is exposed by the
-the least common supertype relation. We omit the Agda encoding of the subtyping relation for
+A key feature of the list machine type system is its subtyping (denoted by \ensuremath{\D{<:}}) which is
+exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
 space reasons, since it is just an inductive type that encode the rules presented
 in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
-subtyping including its decidability.
+subtyping including its decidable test.
 
@@
 The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
+Recently, the interest on such interpreters was revitalized by Amin et. al.~\cite{Amin17} which used definitional
+interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
+Unlike our work, Amin et. al. formalizations does not use intrinsically-typed representations of their syntax which cluttered
+their formalization with the need of dealing with ``stuck states'' in the semantics. Poulsen el. al. described how
+definitional interpreters for imperative languages can be much concisely implemented by using intrisincally-typed syntax by
+using a library for name binding using \emph{scope graphs}~\cite{Poulsen18}, which greatly simplifies the treatment of complex
+binding structures. Since, the list machine benchmark was designed to address other problems than binding, our representation
+using \emph{de Bruijn} indexes was sufficient to implement the desired typechecker and interpreter.
+Other recent application of definitional interpreters was proposed by Rouvoet et. al.~\cite{Rouvoet20}. The main contribution
+of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
+linear typing features in the intepreter, authors have implemented monads based on a separation algebra which they
+named \emph{Market}, supporting the main operations for accessing and update the store used by the intepreter.
+Another work using intrisincally-typed syntax for resource control was developed by Thiemann~\cite{Thiemann19}, which
+implemented an interpreter for a more realistic core functional session typed calculus including recursion and
+session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
+simple interface to a scheduler.
 }

### 2020-06-23 papers/SBLP2020/paper.tex
More updates on related work

@@
 \paragraph{Benchmarks for PL mechanization}{
-POPLMark chalenge~\cite{Aydemir05} .... Another chalenge was proposed by Pientka et. al.~\cite{Pientka18}...
-}
+The importante of benchmarks is recognized in several areas in computer science. A relevent example is the
+TPTP library for the theorem proving community~\cite{Sut17}. In the context of programming languages community,
+the POPLMark challenge~\cite{Aydemir05} was developed by a group renowned programming language researchers aiming
+the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
+was to motivate authors to formalize all of their theorems using such tools. Since the focus of the POPLMark challenge
+was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
+machine benchmark was proposed by Appel et. al.~\cite{Appel07} as an exercise in formalizing results that interest
+compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
+solution to this benchmark using intrinsically-typed syntax in Agda programming language. Representation of binding syntax
+was the subject of Felty et. al.~\cite{FeltyMP18} which proposed a set of problems and research questions for tools
+that use the high-order abstract syntax approach for name binding. Since the list machine benchmark definition avoids
+name binding issues because they are orthogonal to most of compiler related proofs~\cite{Appel07}, we just ensure
+the correct manipulation of names by following the traditional de Bruijn approach. Finally, recent problem set
+was proposed by Pientka et. al.~\cite{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
+of logical relation arguments, like strong normalization theorems.}
 
@@
 session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
-simple interface to a scheduler.
-}
+simple interface to a scheduler. Since our objective was formalizing the list machine benchmark in Agda and it
+does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
+of linear and session types as~\cite{Rouvoet20,Thiemann19}.}
 
 \section{Conclusion}\label{sec:conclusion}
+
+This paper shows the combination of intrinsically-typed syntax and definitional interpreters to formalize programming languages.
+Using such approach, we were able to provide a machine checked version of the list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
+The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM, since we were able to encode features such as jumps, mutable state, and sub-typing.
+When comparing such approach with the conventional syntactic way to prove properties, we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.
+This happens because the approach uses the power of the host language, receiving some proofs for free.
+
+As future work, we want to reuse the ideas presented in this paper to provide an intrinsically-typed formalization for languages like the eBPF and the LUA VM.
+Furthermore, one can extend the formalization presented here for other programming languages with similar settings.
 

### 2020-06-24 papers/SBLP2020/paper.tex
Before merge

--- 2020-06-21 papers/SBLP2020/paper.tex
@@
 
-\title{An Intrinsically Typed Solution for the List Machine Benchmark}
+\title{An Intrinsically Typed Solution for the List-Machine Benchmark}
 
@@
 such errors is to encode models in a dependently-typed language in order
-to ensure its ``correctness-by-construction''.
-
-In this work, we use this idea to build a verified interpreter
-for the list machine benchmark in the Agda programming language.
+to ensure its ``correctness-by-construction''. In this work, we use this
+idea to build a verified interpreter for the list-machine benchmark in the
+Agda programming language, comparing the results with the usual syntactic
+approach.
 \end{abstract}
@@
 
-The development of a new programming language design, linguistic construct
-or type system involves its careful formalization in order to express
+The development of a new programming language design, linguistic construct,
+or type system involves a careful formalization in order to express
 its core ideas in a concise way. However, such models have many details
-and complexities which can hinders its correctness assurances.
+and complexities which can hinder its correctness assurances.
 Because of such problems, the programming languages research community
@@
 meta-language~\cite{Reynolds72}. A major advantage of such approach
-is the possibility of validating the semantics through execution.
+is the possibility to validate the semantics through execution.
 Recently, definitional interpreters were used to formalize type
@@
 semantics is ensured by dependently-typed
-syntax\footnote{Also known as intrinsically-typed.}~\cite{Poulsen18}.
-In this work, we follow Poulsen et. al and use an intrinsically-typed
+syntax\footnote{Also known as intrinsically-typed.}\cite{Poulsen18}.
+In this work, we follow Poulsen et. al. and use an intrinsically-typed
 representation to build a definitional interpreter for a low-level virtual
 machine developed by Appel and Leroy as a benchmark problem closer to
-real-world implementations like typed assembly languages~\cite{CraryM99} and
+real-world implementations such as typed assembly languages~\cite{CraryM99} and
 proof carrying code~\cite{Necula97}.
@@
 \begin{itemize}
-  \item We show how all the details of the list machine type system
-        can be encoded as dependently-typed syntax which avoids, by construction,
+  \item We show how all the details of the list-machine type system
+        can be encoded in a dependently-typed syntax setting, which avoids, by construction,
         the presence of stuck states in its definitional interpreter.
-  \item We provide a provably correct implementations for testing the subtyping
+  \item We provide a provably correct implementation for testing the subtyping
         relation and to calculate the least common super type of two input
         types for the machine registers.
-  \item We provide a correct by construction implementation of a type checker for
-        list machine programs. Our type checker produces, as result, an intrinsically-typed
+  \item We provide a correct-by-construction implementation of a type checker for
+        list-machine programs. Our type checker produces, as result, an intrinsically-typed
         representation of the machine code.
+  \item We compare the results of an intrinsic approach encoded in Agda with extrinsic
+        formalizations encoded in Twelf and Coq. We show that such intrinsic encoding avoids
+        unnecessary repetitions and provides some properties for free.
 \end{itemize}
@@
 presents a brief introduction to Agda and Section~\ref{sec:list}
-reviews the list machine benchmark and presents its syntax and type system.
-We describe the intrinsically-typed representation for the list machine syntax,
-the subtyping relation and the least common super type algorithm in
-Section~\ref{sec:typing}. The list machine semantics and its realization as
-a definitional interpreter are presented in Section~\ref{sec:semantics}.
-Related work is discussed in Section~\ref{sec:related},
-and Section~\ref{sec:conclusion} concludes.
+reviews the list-machine benchmark and presents its syntax and type system.
+We describe the intrinsically-typed representation for the list-machine
+in Section~\ref{sec:typing}. The list-machine semantics and its
+realization as a definitional interpreter are presented in Section~\ref{sec:semantics}.
+The type-checking algorithm, the subtyping relation and the least-common-supertype
+algorithm are presented in Section~\ref{sec:typechecker}. Section~\ref{sec:comparison}
+compares the presented formalization in Agda with Coq and Twelf encodings. Related work is
+discussed in Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.
 
 All the source code in this article has been formalized in Agda
-version 2.6.1 using Standard Library 1.3. All source code produced,
+version 2.6.1 using the Standard Library 1.3. All source code produced,
 including the \LaTeX~source of this article, are available
@@
 a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
-that asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
+which asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
 and a vector \ensuremath{\V{xs}}, the type \ensuremath{\D{All}\;\V{P}\;\V{xs}} is used to build proofs that \ensuremath{\V{P}} holds for all
@@
 \end{hscode}\resethooks
-Using this data type,\footnote{Agda supports the definition of mixfix operators.
-We can use underscores to mark arguments positions.} we can construct expressions
-that denote terms that should not be considered well-typed like
+Using this data type\footnote{Agda supports the definition of mixfix operators.
+We can use underscores to mark arguments positions.}, we can construct expressions
+to denote terms that should not be considered well-typed like
 \ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% List machine benchmark %%
+%% List-machine benchmark %%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{The List Machine Benchmark}\label{sec:list}
-
-The list machine is a simple pointer virtual machine whose values
+\section{The List-Machine Benchmark}\label{sec:list}
+
+The list-machine, proposed by \citet{Appel07}, is a simple pointer virtual machine where values
 are empty lists and cons-cells:
@@
 \]
-We let meta-variable $a$ to denote an arbitrary value, $v$ denotes a variable
+Throughout this text, the meta-variable $a$ is used to denote an arbitrary value, $v$ denotes a variable
 and $l$ a program label. Following common practice, all meta-variables can
@@
 \]
-Programs are just a sequence of blocks which begin with a unique label.
-
-Each program variable is assigned a list type which is used to guarantee the safety of execute
+A program is just a sequence of blocks referenced by a unique label.
+
+Each program variable is assigned to a list type, \textcolor{red}{which is used to guarantee the safety of execute
 fetch-field that demands non-empty list arguments and branch instructions refine the argument type
-according to which branch is taken. In order to express such refinements, types are subject to a
-subtyping relation. We let the meta-variable $\tau$ denote an arbitrary type. 
+according to which branch is taken.} In order to express such refinements, types are subject to a
+subtyping relation. The meta-variable $\tau$ denotes an arbitrary type. 
 \[
@@
 
-Following common practice, we let meta-variable $\Gamma$ denote an
+Following common practice, the meta-variable $\Gamma$ denotes an
 environment between variable names and
@@
 \]
-We let variable $\Pi$ denote \emph{program typings}, i.e. finite mappings between
+The variable $\Pi$ is used to denote \emph{program typings}, i.e. finite mappings between
 labels and typing contexts $\Gamma$, where notation $\Pi(l) = \Gamma$ denotes that
 $\Gamma$ stores the types of variables on the entry point of the block labeled by $l$.
-Using the previously defined notations, we can define the typing rules for the list
-machine as a syntax directed judgement $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$
-which intuitively means that the instruction $\iota$ transforms an input typing environment,
-$\Gamma$, into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
-rules for the list machine instructions are as follows.
+Using the previously defined notations, the typing rules for the list-machine are defined
+as a syntax directed judgement $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
+which intuitively means that the instruction $\iota$ transforms an input typing environment
+$\Gamma$ into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
+rules for the list-machine instructions are defined as follows.
 
@@
 \]
-The type system proposed by Appel et. al.~\cite{Appel07} has three rules to deal with each of the possible
-types assigned to the branch variable in current typing context. The first rules deal with the $list$ and $listcon$
-types, which specifies that the environment associated to the label $l$, $\Pi(l) = \Gamma_1$, is greater than
+The type system proposed by \citet{Appel07} has three rules to deal with each of the possible
+types assigned to the branch variable in the current typing context. The first two rules deal with the $list$ and $listcon$
+types, specifying that the environment associated to the label $l$, $\Pi(l) = \Gamma_1$, is greater than
 $\Gamma[v := nil]$. The third rule applies whenever $\Gamma(v) = nil$ and it also demands that $\Gamma \subset \Pi(l)$.
@@
 \\ \\
-\inference{\Gamma(v) = \text{nil}~~~\Pi(l) = \Gamma_1~~~\Gamma \subset \Gamma_1}
+\inference{\Gamma(v) = \text{nil}\:\:\:\:\Pi(l) = \Gamma_1\:\:\:\:\Gamma \subset \Gamma_1}
           {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}
@@
 \]
-Next, we have the \emph{fetch} instruction which can be used to store the head / tail of a list value in
+Next, we have the \emph{fetch} instructions, which can be used to store the head / tail of a list value in
 a variable. Rule \emph{fetch-0} retrieves the head of a value stored in a variable $v$ and \emph{fetch-1}
@@
 \begin{array}{c}
-\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\tau]=\Gamma'}
+\inference{\Gamma(v) = \text{listcons}~\tau\:\:\:\:\Gamma[v':=\tau]=\Gamma'}
           {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}[fetch-0]
           \\ \\
-\inference{\Gamma(v) = \text{listcons}~\tau~~~\Gamma[v':=\text{list}~\tau]=\Gamma'}
+\inference{\Gamma(v) = \text{listcons}~\tau\:\:\:\:\Gamma[v':=\text{list}~\tau]=\Gamma'}
           {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}[fetch-1]
@@
 \]
-The final rules deal with the well-formedness of blocks and programs. The typing rules for \textbf{halt} instruction and
+The final rules deal with the well-formedness of blocks and programs. The typing rules for the \textbf{halt} instruction and
 program \textbf{end} are completely trivial. Rule \emph{block-seq} does the typing context threading between sequential
@@
 \]
-Inspired by the typing rules designed by Appel et. al., in the next section, we define an intrisinsically
-typed syntax for list machine programs which ensures that only well-typed programs can be built.
+Inspired by the presented typing rules, in the next section, we define an intrisinsically
+typed syntax for list-machine programs which ensures that only well-typed programs can be built.
 
@@
 
-In this section we present the design choices and the steps we took to represent the
+In this section we introduce the design choices and the steps we took to represent the
 intrinsically-typed syntax for the list-machine benchmark. We present here the Agda
@@
 We internalize the list-machine type judgments for blocks and instructions in Agda
-together with its syntax in such a way where only well-typed terms that satisfy typing
+together with its syntax in such a way where only well-typed terms that satisfying typing
 judgments have meaning. This approach makes the AST contain both syntactic and semantic
@@
 
-In order to be considered well-typed, list-machine machine programs needs to refer
-to information from two sources: (1) a type context encoded as a list of types to store
-variable types; and (2) and a program context encoded as a vector\footnote{We use the
-\ensuremath{\D{Vec}} datatype indexed by a \ensuremath{\V{n}} which is bound on the module definition and represents
+In order to be considered well-typed, list-machine machine programs need to refer
+to information from two sources: (1) a type context encoded as a list to store
+variable types; and (2) a program context encoded as a vector\footnote{We use the
+\ensuremath{\D{Vec}} datatype indexed by an \ensuremath{\V{n}} which is bound on the module definition and represents
 the number of labels in the current program.} of type contexts to represent the types of
@@
 to hold the well-typed terms accordingly, representing each judgment of the static semantics as a
-syntactical constructor. In Agda we use \emph{indexed inductive types} to define a intrinsically-typed syntax.
-The basic idea is to represent each type system rule as a constructor of the underlying type.
+syntactical constructor. In Agda we use \emph{indexed inductive types} to define an intrinsically-typed syntax.
+The basic idea is to represent each type system rule as a constructor typed by an output context.
 Such representation scheme makes the Agda's type-checker allow only well-typed blocks and
@@
 
-\paragraph{Conditional jump}{There are three constructors related to a conditional
+\paragraph{Conditional jump}{There are three constructors (\ensuremath{\Con{instr\textrm{-}branch\textrm{-}list}},
+\ensuremath{\Con{instr\textrm{-}branch\textrm{-}listcons}}, and \ensuremath{\Con{instr\textrm{-}branch\textrm{-}nil}}) related to a conditional
 jump. They are used to perform a jump to a label \ensuremath{\V{l}} when the received variable is \ensuremath{\Con{nil}}.
@@
 to fetch information from a given list. The constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}} receives a
-non-empty list (\ensuremath{\Con{listcons}}), and is used to retrieve the head of this list and store it
+non-empty list (of type \ensuremath{\Con{listcons}\;\V{\tau}}), and is used to retrieve the head of this list and store it
 in a fresh new variable. The resulting typing context adds the information about the new
@@
 The constructors \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}}
-are similar, fetching the tail of a list instead of the head.}
+are similar, but fetching the tail of a list instead of the head.}
 
 \paragraph{List construction}{The \ensuremath{\Con{instr\textrm{-}cons\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}cons\textrm{-}upd}} constructors are
-used to create a new list. The first creates a new variable, and the second updates a
+used to create a new list. The first creates a new variable, and the second updates an
 existing variable. The list is created from two variables, \ensuremath{(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}} and \ensuremath{(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}},
@@
 The resulting typing context adds information about the newly created list.
-Section~\ref{sec:supertype} will provide details of an algorithm
-for calculating the least common supertype of two given types.}
+\textcolor{red}{Section~\ref{sec:supertype} will provide details of an algorithm
+for calculating the least common supertype of two given types.}}
 
@@
 And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label.
-We use the \ensuremath{\D{All}} datatype to express this relationship.
+We use the \ensuremath{\D{All}} datatype to express this relation.
 
@@
 first two represent \ensuremath{\Con{nil}} values associated with type \ensuremath{\Con{nil}} and the empty \ensuremath{\Con{list}}. The last
-two represent non-empty list, considering the types \ensuremath{\Con{listcons}} and \ensuremath{\Con{list}}.
+two represent non-empty lists, considering the types \ensuremath{\Con{listcons}} and \ensuremath{\Con{list}}.
 
@@
 list of well-typed values. And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
-Both type environments are used to type block instructions and sequence of
+Both type environments are used to type block instructions and sequences of
 block instructions.
@@
 i.e., each program developed in it must terminate and all possible patterns must
-be matched. However, by using the mechanisms for jumping between labels one can
+be matched. However, by using the mechanisms for jumping between labels, one can
 write a program which never ends, making it impossible to implement a terminating
@@
 last one is the \ensuremath{\D{Block}} to be evaluated. This function returns a modified run-time
-environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or nothing when the \emph{fuel} runs out.
+environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or nothing when (and only) the \emph{fuel} runs out.
 
@@
 \paragraph{Conditional jump}{We show next only the case when the jump actually occurs,
-following the rule \emph{step-branch-taken}. In this case, variable \ensuremath{\V{v}} has value \ensuremath{\Con{nil}}, and
+following the rule \emph{step-branch-taken} of \cite{Appel07}. In this case, variable \ensuremath{\V{v}} has value \ensuremath{\Con{nil}}, and
 the step of evaluation should proceed with the block instruction defined in program \ensuremath{\V{p}}, with
@@
 \paragraph{Fetching list information}{The next code shows the evaluation of two syntactical constructors,
-both related to the \emph{step-fetch-field-0} rule. The first retrieves the head element of a list, and
+both related to the \emph{step-fetch-field-0} rule~\cite{Appel07}. The first retrieves the head element of a list, and
 stores it in a new variable. The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
@@
 it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}. This process is done by the
-function \ensuremath{\F{update\textrm{-}env}}.}
+function \ensuremath{\F{update\textrm{-}env}}\footnote{The source-code of this function can be found in our online repository.}.}
 
@@
 
-Since we use \emph{de Bruijn} indices to represent labels and variables, the first step to type and scope
+Since we use \emph{de Bruijn} indices to represent labels and variables, as the first step to type and scope
 check them, we need to provide an index from a named variable. This process is done by the
-\ensuremath{\F{lookup\textrm{-}var}} function.
+\ensuremath{\F{lookup\textrm{-}var}} function, returning an index when the named variable variable is found, and \ensuremath{\Con{nothing}} otherwise.
 
@@
 
-For space reasons, we show how we type check only one instruction. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
+For space reasons, we show how we type check only one instruction\footnote{The curious reader can refer to our online repository for a complete implementation.}.
+Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
 receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value,
@@
 \\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{nil}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error\char34}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{list}\;\V{\tau}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error\char34}{}\<[E]%
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{nil}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error~[nil]\char34}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{list}\;\V{\tau}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error~[list]\char34}{}\<[E]%
 \\
@@
 each variable, and match the first with its possible forms. The first three cases indicate
-type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{nothing}} it means an scope error; (2) and (3) are typing errors,
+type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{nothing}} it means a variable scope error; (2) and (3) are typing errors,
 since the type of variable \ensuremath{\V{v}} should be a \ensuremath{\Con{listcons}}. Last two cases represent that the instruction
@@
 
-A key feature of the list machine type system is its subypting which is exposed by the
-the least common supertype relation. We omit the Agda encoding of the subtyping relation for
+A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}) which is
+exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
 space reasons, since it is just an inductive type that encode the rules presented
 in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
-subtyping including its decidability.
-
-The least common supertype relation is used by the least machine type system to refine a
+subtyping including its decidable test.
+
+The least common supertype relation is used by the list-machine type system to refine a
 variable type whenever it is updated by a \emph{cons} instruction. The rules of the
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcap-comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-We omit the complete definition of \ensuremath{\F{\sqcap-comm}} for brevity. In the list machine benchmark definition~\cite{Appel07},
-Appel et. al. also defines that the least common supertype relation is sound and complete with respect to the
-subtyping relation. Again, we omit the definitions of these properties, whose type is presented below.
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcap-subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
-\\
-\>[B]{}\F{\sqcap-least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
+\>[B]{}\F{\sqcap\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
+\citet{Appel07} also define that the least common supertype relation is sound and complete with respect to the
+subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{\sqcap\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
+\\
+\>[B]{}\F{\sqcap\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
 \ColumnHook
@@
 
-\section{Comparison of Mechanized Proofs}
-
-We implemented 14 tasks from the list machine benchmark in the Agda programming language.
-The selected tasks considered by us are the ones done by Appel et. al.~\cite{Appel07}.
-The next table summarize the total number of lines for our results together with Appel's.
+\section{Comparison of Mechanized Proofs}\label{sec:comparison}
+
+We implemented the 14 tasks from the list-machine benchmark in the Agda programming language.
+The tasks considered by us are the same implemented and proved by~\citet{Appel07}.
+The next table summarize the total number of lines for our results together with theirs.
 
@@
 on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5. We briefly comment 
-on our Agda coding of these 14 tasks.
+on our Agda encoding of these 14 tasks.
 
 \begin{enumerate}
-  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we choose to
+  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we chose to
         use a definitional interpreter for our intrinsincally-typed representation. Our implementation for the operational
         semantics is composed by 38 lines for the type syntax and 71 for the definitional interpreter.
-  \item \textbf{Derive $p\,\Downarrow$.} Since we used an definitional intepreter for representing the semantics, we can derive
+  \item \textbf{Derive $p \Downarrow$.} Since we used a definitional intepreter for representing the semantics, we can derive
         $p\,\Downarrow$ just by executing the interpreter on $p$.
   \item \textbf{Represent the type system.} Our type system representation consists of the intrinsically-typed encoding of
-        the list machine programs and it is already counted in as part of the operational semantics of our solution.
+        the list-machine programs and it is already counted in as part of the operational semantics of our solution.
   \item \textbf{Least common supertype algorithm.} In order to implement this task, we specified the least common super type
@@
         guarantee its soundness w.r.t. its specification.
-  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list machine program
+  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list-machine program
         is just execute the type checker over it. Since our type checker returns an intrinsincally-typed representation of
@@
         are simple recursive functions over the least common supertype and subtyping relations definition.
-  \item \textbf{State the soundness theorem for the type system.} Following Amin et. al.~\cite{Amin17}, we represent the soundness
-        theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list machine
+  \item \textbf{State the soundness theorem for the type system.} Following~\citet{Amin17}, we represent the soundness
+        theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list-machine
         programs.
-  \item \textbf{Prove the soundness theorem for the type system.} Following Amin et. al.~\cite{Amin17}, the proof of the soundness
+  \item \textbf{Prove the soundness theorem for the type system.} Following \citet{Amin17}, the proof of the soundness
         theorem is just the implementation of the definitional interpreter. Soundness is ensured by construction since the interpreter
@@
 \paragraph{Benchmarks for PL mechanization}{
-POPLMark chalenge~\cite{Aydemir05} .... Another chalenge was proposed by Pientka et. al.~\cite{Pientka18}...
-}
+The importante of benchmarks is recognized in several areas in computer science. A relevent example is the
+TPTP library for the theorem proving community~\cite{Sut17}. In the context of programming languages community,
+the POPLMark challenge~\cite{Aydemir05} was developed by a renowned group of programming language researchers aiming
+the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
+was to motivate authors to formalize all of their theorems using such tools. Since the focus of the POPLMark challenge
+was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
+machine benchmark was proposed by~\citet{Appel07} as an exercise in formalizing results that interest
+compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
+solution to this benchmark using an intrinsically-typed approach in the Agda programming language. Representation of binding syntax
+was the subject of~\citet{FeltyMP18} which proposed a set of problems and research questions for tools
+that use the high-order abstract syntax approach for name binding. Since the list-machine benchmark definition avoids
+name binding issues because they are orthogonal to most of compiler related proofs~\cite{Appel07}, we just ensure
+the correct manipulation of names by following the traditional \emph{de Bruijn} approach. Finally, a recent problem set
+was proposed by~\citet{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
+of logical relation arguments, like strong normalization theorems.}
 
@@
 The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
-}
+Recently, the interest on such interpreters was revitalized by~\citet{Amin17} which used definitional
+interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
+Unlike our work, Amin and Rompf's formalizations does not use intrinsically-typed representations of their syntax which cluttered
+their formalization with the need of dealing with ``stuck states'' in the semantics. \citet{Poulsen18} described how
+definitional interpreters for imperative languages can be much concisely implemented by using intrisincally-typed syntax by
+using a library for name binding using \emph{scope graphs}, which greatly simplifies the treatment of complex
+binding structures. Since the list-machine benchmark was designed to address other problems than binding, our representation
+using \emph{de Bruijn} indices was sufficient to implement the desired type checker and interpreter.
+Other recent application of definitional interpreters was proposed by~\citet{Rouvoet20}. The main contribution
+of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
+linear typing features in the interpreter, the authors have implemented monads based on a separation algebra which they
+named \emph{Market}, supporting the main operations for accessing and update the store used by the intepreter.
+Another work using intrisincally-typed syntax for resource control was developed by~\citet{Thiemann19}, which
+implemented an interpreter for a more realistic core functional session typed calculus including recursion and
+session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
+simple interface to a scheduler. Since our objective was formalizing the list-machine benchmark in Agda and it
+does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
+of linear and session types as in~\cite{Rouvoet20,Thiemann19}.}
 
 \section{Conclusion}\label{sec:conclusion}
+
+This paper shows the combination of intrinsically-typed syntax and definitional interpreters to formalize programming languages.
+Using such approach, we were able to provide a machine checked version of the list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
+The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM, since we were able to encode features such as jumps, mutable state, and sub-typing.
+When comparing such approach with the conventional syntactic way to prove properties, we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.
+This happens because the approach uses the power of the host language, and provides some proofs for free due to the intrinsically-typed syntax.
+
+As future work, we want to reuse the ideas presented in this paper to provide an intrinsically-typed formalization for real-world low-level languages like the eBPF and the LUA VM.
+Furthermore, one can extend the formalization presented here for other programming languages with similar settings.
 

### 2020-06-26 papers/SBLP2020/paper.tex
Before merge

@@
 such errors is to encode models in a dependently-typed language in order
-to ensure its ``correctness-by-construction''. In this work, we use this
+to ensure its ``correctness-by-construction''. In this paper, we use this
 idea to build a verified interpreter for the list-machine benchmark in the
-Agda programming language, comparing the results with the usual syntactic
-approach.
+Agda programming language, comparing the results with formalizations using
+the usual syntactic approach. We formalize the 14 tasks of the benchmark
+using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
+compared to a Coq solution, even without the use of proof automation.
 \end{abstract}
@@
 
-\section{An Overview of Agda}\label{sec:agda}
-
-Agda is a dependently-typed functional programming language based on
-Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
-and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
-natural number, are built-in. Everything else is a user-defined
-type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
-``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
-\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}}, \ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}.
-We have that \ensuremath{\D{Set}\;\V{l}} is an
-element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
-stratification of types is used to keep Agda consistent as a logical
-theory~\cite{Sorensen2006}.
-
-An ordinary (non-dependent) function type is written \ensuremath{\V{A}\;\to \;\V{B}} and a
-dependent one is written \ensuremath{(\V{x}\;\mathbin{:}\;\V{A})\;\to \;\V{B}}, where type \ensuremath{\V{B}} depends on
-\ensuremath{\V{x}}, or \ensuremath{\D{\forall}\;(\V{x}\;\mathbin{:}\;\V{A})\;\to \;\V{B}}. Agda allows the definition of \emph{implicit
-parameters}, i.e.,  parameters whose values can be inferred from the
-context, by surrounding them in curly braces: \ensuremath{\D{\forall}\;\{\mskip1.5mu \V{x}\;\mathbin{:}\;\V{A}\mskip1.5mu\}\;\to \;\V{B}}. To
-avoid clutter, we'll omit implicit arguments from the source code
-presentation. The reader can safely assume that every free variable in
-a type is an implicit parameter.
-
-As an example of Agda code, consider the following data type of
-length-indexed lists, also known as vectors.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathkw{data}\;\D{\mathbb{N}}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\mathbb{N}}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{\mathbb{N}}{}\<[E]%
-\\[\blanklineskip]%
-\>[3]{}\mathkw{data}\;\D{Vec}\;(\V{A}\;\mathbin{:}\;\D{Set})\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{\lbrack\:\rbrack}\;{}\<[9]%
-\>[9]{}\mathbin{:}\;\D{Vec}\;\V{A}\;\Con{zero}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\V{A}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})
-inserts a new element in front of a vector of $n$ elements (of type
-\ensuremath{\D{Vec}\;\V{A}\;\V{n}}) and returns a value of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}. The
-\ensuremath{\D{Vec}} datatype is an example of a dependent type, i.e., a type
-that uses a value (that denotes its length). The usefulness of
-dependent types can be illustrated with the definition of a safe list
-head function: \ensuremath{\F{head}} can be defined to accept only non-empty
-vectors, i.e.,~values of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}.
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\F{head}\;\mathbin{:}\;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})\;\to \;\V{A}{}\<[E]%
-\\
-\>[3]{}\F{head}\;(\V{x}\;\Con{::}\;\V{xs})\;\mathrel{=}\;\V{x}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-In \ensuremath{\F{head}}'s definition, constructor \ensuremath{\Con{\lbrack\:\rbrack}} is not used. The
-Agda type-checker can figure out, from \ensuremath{\F{head}}'s parameter type,
-that argument \ensuremath{\Con{\lbrack\:\rbrack}} to \ensuremath{\F{head}} is not type-correct.
-
-Another useful data type is the finite type,
-\ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of
-data type constructor names. Constructor \ensuremath{\Con{zero}} can refer to
-type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the context where the name
-is used.}, which is defined in Agda's standard library as:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathkw{data}\;\D{Fin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-Type \ensuremath{\D{Fin}\;\V{n}} has exactly \ensuremath{\V{n}} inhabitants
-(elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
-An application of such type is to define a safe vector
-lookup function, which avoids the access of invalid positions.
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\F{lookup}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\V{A}{}\<[E]%
-\\
-\>[3]{}\F{lookup}\;\Con{zero}\;(\V{x}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{x}{}\<[E]%
-\\
-\>[3]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{xs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{xs}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-Thanks to the propositions-as-types principle,\footnote{It is also known as
-  Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
-types as logical formulas and terms as proofs. An example is the
-representation of equality as the following Agda type:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathkw{data}\;\D{\_ \equiv \_}\;\{\mskip1.5mu \V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{A}\;\mathbin{:}\;\D{Set}\;\V{l}\mskip1.5mu\}\;(\V{x}\;\mathbin{:}\;\V{A})\;\mathbin{:}\;\V{A}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{refl}\;\mathbin{:}\;\V{x}\;\D{\equiv}\;\V{x}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-This type is called propositional equality. It defines that there is
-a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
-which asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
-and a vector \ensuremath{\V{xs}}, the type \ensuremath{\D{All}\;\V{P}\;\V{xs}} is used to build proofs that \ensuremath{\V{P}} holds for all
-elements in \ensuremath{\V{xs}} and it is defined as:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathkw{data}\;\D{All}\;(\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
-\>[41]{}\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[3]{}\hsindent{3}{}\<[6]%
-\>[6]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\V{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
-\\
-\>[3]{}\hsindent{3}{}\<[6]%
-\>[6]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\V{P}\;\V{x}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-The first constructor specifies that \ensuremath{\D{All}\;\V{P}} holds for the empty vector and
-constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
-\ensuremath{\V{P}\;\V{x}} and \ensuremath{\D{All}\;\V{P}\;\V{xs}}. Since this type has the same structure of vectors,
-some functions on \ensuremath{\D{Vec}} have similar definitions for type \ensuremath{\D{All}}. As an example
-used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
-proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\V{P}\;\V{x}{}\<[E]%
-\\
-\>[4]{}\F{lookup}\;\Con{zero}\;(\V{px}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{px}{}\<[E]%
-\\
-\>[4]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{pxs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{pxs}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
+\section{Dependently-typed Syntax}\label{sec:agda}
+
+% Agda is a dependently-typed functional programming language based on
+% Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
+% and an infinite hierarchy of types of types, |Set l|, where |l| is a
+% natural number, are built-in. Everything else is a user-defined
+% type. The type |Set|, also known as |Set0|, is the type of all
+% ``small'' types, such as |Bool|, |String| and |List Bool|.  The type
+% |Set1| is the type of |Set| and ``others like it'', such as |Set ->
+% Bool|, |String -> Set|, and |Set -> Set|. We have that |Set l| is an
+% element of the type |Set (l+1)|, for every $l \geq 0$. This
+% stratification of types is used to keep Agda consistent as a logical
+% theory~\cite{Sorensen2006}.
+
+% An ordinary (non-dependent) function type is written |A -> B| and a
+% dependent one is written |(x : A) -> B|, where type |B| depends on
+% |x|, or |forall (x : A) -> B|. Agda allows the definition of \emph{implicit
+% parameters}, i.e.,  parameters whose values can be inferred from the
+% context, by surrounding them in curly braces: |forall {x : A} -> B|. To
+% avoid clutter, we'll omit implicit arguments from the source code
+% presentation. The reader can safely assume that every free variable in
+% a type is an implicit parameter.
+
+% As an example of Agda code, consider the following data type of
+% length-indexed lists, also known as vectors.
+
+% \begin{spec}
+%   data Nat : Set where
+%     zero : Nat
+%     succ : Nat -> Nat
+
+%   data Vec (A : Set) : Nat -> Set where
+%     []  : Vec A zero
+%     _::_ : forall {n} -> A -> Vec A n -> Vec A (succ n)
+% \end{spec}
+% %format head = "\F{head}"
+% Constructor |[]| builds empty vectors. The cons-operator (|_::_|)
+% inserts a new element in front of a vector of $n$ elements (of type
+% |Vec A n|) and returns a value of type |Vec A (succ n)|. The
+% |Vec| datatype is an example of a dependent type, i.e., a type
+% that uses a value (that denotes its length). The usefulness of
+% dependent types can be illustrated with the definition of a safe list
+% head function: |head| can be defined to accept only non-empty
+% vectors, i.e.,~values of type |Vec A (succ n)|.
+% \begin{spec}
+%   head : Vec A (succ n) -> A
+%   head (x :: xs) = x
+% \end{spec}
+% In |head|'s definition, constructor |[]| is not used. The
+% Agda type-checker can figure out, from |head|'s parameter type,
+% that argument |[]| to |head| is not type-correct.
+
+% %format _==_ = "\D{\_ \equiv \_}"
+% %format == = "\D{\equiv}"
+% %format refl = "\Con{refl}"
+% %format proj₁ = "\F{\pi_1}"
+% %format proj₂ = "\F{\pi_2}"
+% %format Fin   = "\D{Fin}"
+% %format lookup = "\F{lookup}"
+
+% Another useful data type is the finite type,
+% |Fin|\footnote{Note that Agda supports the overloading of
+% data type constructor names. Constructor |zero| can refer to
+% type |Nat| or |Fin|, depending on the context where the name
+% is used.}, which is defined in Agda's standard library as:
+% \begin{spec}
+%   data Fin : Nat -> Set where
+%     zero : forall {n} -> Fin (succ n)
+%     succ : forall {n} -> Fin n -> Fin (succ n)
+% \end{spec}
+% Type |Fin n| has exactly |n| inhabitants
+% (elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
+% An application of such type is to define a safe vector
+% lookup function, which avoids the access of invalid positions.
+% \begin{spec}
+%   lookup : forall {A n} -> Fin n -> Vec A n -> A
+%   lookup zero (x :: _) = x
+%   lookup (succ idx) (_ :: xs) = lookup idx xs
+% \end{spec}
+% Thanks to the propositions-as-types principle,\footnote{It is also known as
+%   Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
+% types as logical formulas and terms as proofs. An example is the
+% representation of equality as the following Agda type:
+
+% \begin{spec}
+%   data _==_ {l}{A : Set l}(x : A) : A -> Set where
+%     refl : x == x
+% \end{spec}
+
+% %format not = "\F{\neg}"
+% %format Dec = "\D{Dec}"
+% %format yes = "\Con{yes}"
+% %format no  = "\Con{no}"
+% %format Even = "\Con{Even}"
+% %format Odd = "\Con{Odd}"
+% %format Parity = "\D{Parity}"
+% %format parity = "\F{parity}"
+% %format natToBin = "\F{natToBin}"
+% %format false = "\Con{false}"
+% %format true = "\Con{true}"
+% %format + = "\F{+}"
+% %format ++ = "\F{++}"
+% %format Bot = "\D{\bot}"
+% %format All = "\D{All}"
+% This type is called propositional equality. It defines that there is
+% a unique evidence for equality, constructor |refl| (for reflexivity),
+% which asserts that the only value equal to |x| is itself. Given a predicate |P : A -> Set|
+% and a vector |xs|, the type |All P xs| is used to build proofs that |P| holds for all
+% elements in |xs| and it is defined as:
+% \begin{spec}
+%   data All (P : A -> Set) : Vec A n ->  Set where
+%      [] : All P []
+%      _::_ : forall {x xs} -> P x -> All P xs -> All P (x :: xs)
+% \end{spec}
+% The first constructor specifies that |All P| holds for the empty vector and
+% constructor |_::_| builds a proof of |All P (x :: xs)| from proofs of
+% |P x| and |All P xs|. Since this type has the same structure of vectors,
+% some functions on |Vec| have similar definitions for type |All|. As an example
+% used in our formalization, consider the function |lookup|, which extracts a
+% proof of |P| for the element at position |v : Fin n| in a |Vec|:
+% \begin{spec}
+%    lookup : {xs : Vec A n} -> Fin n -> All P xs -> P x
+%    lookup zero (px :: _) = px
+%    lookup (succ idx) (_ :: pxs) = lookup idx pxs
+% \end{spec}
+
 An important application of dependent types is to encode programming languages
@@
 host's language type checker. As an example, consider the following simple
-expression language.
+expression language. For this paper, we assume a basic knowledge of functional
+programming and Agda\footnote{For further information about Agda,
+see~\cite{Norell2009,Stump16}.}.
+
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
+
 Using this data type\footnote{Agda supports the definition of mixfix operators.
@@
 to denote terms that should not be considered well-typed like
-\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
+\ensuremath{(\Con{Num}\;\V{1})\;\V{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
 as another definition, which should consider all possible cases to avoid the
@@
 %A definition which uses the expression |(Num 1) + True| will be rejected by Agda's type checker automatically.
-Agda's type checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}.
+Agda's type checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\V{+}\;\Con{True}}.
 
@@
 %\end{spec}
-
-For further information about Agda, see~\cite{Norell2009,Stump16}.
 
@@
 The resulting typing context adds information about the newly created list.
-\textcolor{red}{Section~\ref{sec:supertype} will provide details of an algorithm
-for calculating the least common supertype of two given types.}}
+Subsection~\ref{sec:supertype} will provide details of an algorithm
+for calculating the least common supertype of two given types.}
 
@@
 And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label.
-We use the \ensuremath{\D{All}} datatype to express this relation.
+We use the \ensuremath{\V{All}} datatype to express this relation.
 
@@
 \\
-\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
+\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\subsection{Subtyping and least common supertype}\label{sec:supertype}
+
+A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}) which is
+exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
+space reasons, since it is just an inductive type that encode the rules presented
+in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
+subtyping including its decidable test.
+
+The least common supertype relation is used by the list-machine type system to refine a
+variable type whenever it is updated by a \emph{cons} instruction. The rules of the
+super type relation are specified as the following inductive type:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{\_\sqcap\_\sim\_}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\\
+\>[3]{}\hsindent{10}{}\<[13]%
+\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcap}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\\
+\>[3]{}\hsindent{10}{}\<[13]%
+\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcap}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
+types. Also, rules ensure that the relation is commutative as proved by a simple
+Agda theorem.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{\sqcap\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
+\citet{Appel07} also define that the least common supertype relation is sound and complete with respect to the
+subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{\sqcap\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
+\\
+\>[B]{}\F{\sqcap\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+An algorithm to construct the least supertype from two input types is
+defined by function \ensuremath{\F{lub}}. Note that the \ensuremath{\F{lub}}'s type ensures that the
+returned type \ensuremath{\V{t}} is indeed the least common supertype of \ensuremath{\V{t1}} and \ensuremath{\V{t2}},
+thus ensuring its correctness by construction.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\V{∃}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\F{lub}\;\Con{nil}\;\Con{nil}\;\mathrel{=}\;\Con{nil}\;\red{,}\,\;\Con{lub0}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{list}\;\V{t1})\;\Con{nil}\;\mathrel{=}\;\Con{list}\;\V{t1}\;\red{,}\,\;\Con{lub1}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{list}\;\V{t1})\;(\Con{list}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{list}\;\V{t3}\;\red{,}\,\;\Con{lub3}\;\V{p}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;\Con{nil}\;\mathrel{=}\;\Con{list}\;\V{t1}\;\red{,}\,\;\Con{lub5}{}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;(\Con{list}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{list}\;\V{t3}\;\red{,}\,\;\Con{lub2b}\;(\Con{lub3}\;\V{p}){}\<[E]%
+\\
+\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;(\Con{listcons}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
+\\
+\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{listcons}\;\V{t3}\;\red{,}\,\;\Con{lub7}\;\V{p}{}\<[E]%
+\\
+\>[B]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
 \ColumnHook
@@
 
-We use the datatype \ensuremath{\D{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
+We use the datatype \ensuremath{\V{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
 variable environments and well-typed programs. Thus, intuitively, \ensuremath{\D{Env}} is like a
@@
 \\
-\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\D{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
+\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\V{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
 \ColumnHook
@@
 both related to the \emph{step-fetch-field-0} rule~\cite{Appel07}. The first retrieves the head element of a list, and
-stores it in a new variable. The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
+stores it in a new variable. The \ensuremath{\V{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
 environment \ensuremath{\V{env}}, and this variable is added to the result environment. The typed \emph{de Bruijn}
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\>[3]{}\mathkw{with}\;\V{lookup}\;\V{env}\;\V{v}{}\<[E]%
 \\
@@
 types, we need some extra lemmas. First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the
-result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\F{lookup}}
+result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\V{lookup}}
 function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\F{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
+\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\V{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
+\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\V{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
 \ColumnHook
@@
 \\
-\>[B]{}\V{...}\;\mid \;\Con{yes}\;\Con{refl}\;\mathrel{=}\;\Con{just}\;(\V{\tau}\;\red{,}\,\;\Con{here}\;\Con{refl}){}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;(\Con{no}\;\V{¬p})\;\mathrel{=}\;\F{map}\;{}\<[22]%
+\>[B]{}\V{...}\;\mid \;\V{yes}\;\V{refl}\;\mathrel{=}\;\Con{just}\;(\V{\tau}\;\red{,}\,\;\Con{here}\;\V{refl}){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;(\V{no}\;\V{¬p})\;\mathrel{=}\;\F{map}\;{}\<[22]%
 \>[22]{}(\lambda\;\V{v}\;\rightarrow\;\D{proj_1}\;\V{v}\;\red{,}\,\;\Con{there}\;(\D{proj_2}\;\V{v}))\;{}\<[E]%
@@
 \\
-\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;(\text{\ttfamily \char34 variable~out~of~scope:~\char34}\;\F{++}\;\V{v}){}\<[E]%
+\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;(\text{\ttfamily \char34 variable~out~of~scope:~\char34}\;\plus \;\V{v}){}\<[E]%
 \\
@@
 
-\subsection{Subtyping and least common supertype}\label{sec:supertype}
-
-A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}) which is
-exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
-space reasons, since it is just an inductive type that encode the rules presented
-in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
-subtyping including its decidable test.
-
-The least common supertype relation is used by the list-machine type system to refine a
-variable type whenever it is updated by a \emph{cons} instruction. The rules of the
-super type relation are specified as the following inductive type:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathkw{data}\;\D{\_\sqcap\_\sim\_}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
-\\
-\>[3]{}\hsindent{10}{}\<[13]%
-\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcap}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
-\\
-\>[3]{}\hsindent{10}{}\<[13]%
-\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcap}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-Constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
-types. Also, rules ensure that the relation is commutative as proved by a simple
-Agda theorem.
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcap\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
-\citet{Appel07} also define that the least common supertype relation is sound and complete with respect to the
-subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcap\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
-\\
-\>[B]{}\F{\sqcap\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-An algorithm to construct the least supertype from two input types is
-defined by function \ensuremath{\F{lub}}. Note that the \ensuremath{\F{lub}}'s type ensures that the
-returned type \ensuremath{\V{t}} is indeed the least common supertype of \ensuremath{\V{t1}} and \ensuremath{\V{t2}},
-thus ensuring its correctness by construction.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\D{\exists}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\F{lub}\;\Con{nil}\;\Con{nil}\;\mathrel{=}\;\Con{nil}\;\red{,}\,\;\Con{lub0}{}\<[E]%
-\\
-\>[B]{}\F{lub}\;(\Con{list}\;\V{t1})\;\Con{nil}\;\mathrel{=}\;\Con{list}\;\V{t1}\;\red{,}\,\;\Con{lub1}{}\<[E]%
-\\
-\>[B]{}\F{lub}\;(\Con{list}\;\V{t1})\;(\Con{list}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
-\\
-\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{list}\;\V{t3}\;\red{,}\,\;\Con{lub3}\;\V{p}{}\<[E]%
-\\
-\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;\Con{nil}\;\mathrel{=}\;\Con{list}\;\V{t1}\;\red{,}\,\;\Con{lub5}{}\<[E]%
-\\
-\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;(\Con{list}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
-\\
-\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{list}\;\V{t3}\;\red{,}\,\;\Con{lub2b}\;(\Con{lub3}\;\V{p}){}\<[E]%
-\\
-\>[B]{}\F{lub}\;(\Con{listcons}\;\V{t1})\;(\Con{listcons}\;\V{t2})\;\mathkw{with}\;\F{lub}\;\V{t1}\;\V{t2}{}\<[E]%
-\\
-\>[B]{}\V{...|}\;\V{t3}\;\red{,}\,\;\V{p}\;\mathrel{=}\;\Con{listcons}\;\V{t3}\;\red{,}\,\;\Con{lub7}\;\V{p}{}\<[E]%
-\\
-\>[B]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
 \section{Comparison of Mechanized Proofs}\label{sec:comparison}
@@
 The tasks considered by us are the same implemented and proved by~\citet{Appel07}.
-The next table summarize the total number of lines for our results together with theirs.
+The next table summarize the total number of lines of code (LOC) for our results together with theirs.
 
@@
         maps implemented by efficient search trees) can solve this issue. We leave this fix to future work.
-  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it ammounts to calling the typechecker
+  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it is only a matter of calling the typechecker
         on the input program.
@@
 \end{enumerate}
+
+As we could notice on the table above, the approach taken in this paper avoids code repetition and decreases the number of LOCs necessary to implement the proposed language and prove its properties.
+The formalization of this benchmark took 2898 LOC to be done in Twelf, and 887 LOC in Coq, while our implementation used only 413 LOC to achieve the same result.
+Our encoding uses approximately 14\% of the Twelf LOC, and 47\% of Coq's, mostly due to the fact the soundness properties are obtained for free when combining an intrinsically-typed syntax with an implementation of a definitional interpreter in a total language like Agda.
 

### 2020-06-29 papers/SBLP2020/paper.tex
Done some re-reading andfixing some tiny things mainly on comparison section.

@@
 idea to build a verified interpreter for the list-machine benchmark in the
-Agda programming language, comparing the results with formalizations using
-the usual syntactic approach. We formalize the 14 tasks of the benchmark
+Agda programming language, comparing the results with formalizations developed
+by Appel and Leroy. We formalize the 14 tasks of the benchmark
 using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
@@
 We describe the intrinsically-typed representation for the list-machine
-in Section~\ref{sec:typing}. The list-machine semantics and its
-realization as a definitional interpreter are presented in Section~\ref{sec:semantics}.
+in Section~\ref{sec:typing}. Section~\ref{sec:semantics} briefly discuss
+the list-machine semantics and then proceeds with its
+realization as a definitional interpreter.
 The type-checking algorithm, the subtyping relation and the least-common-supertype
@@
 
-Each program variable is assigned to a list type, \textcolor{red}{which is used to guarantee the safety of execute
-fetch-field that demands non-empty list arguments and branch instructions refine the argument type
-according to which branch is taken.} In order to express such refinements, types are subject to a
+Each program variable is assigned to a list type, which is used to guarantee the safety of execute
+fetch-field that demands non-empty list arguments. In order to express such refinements, types are subject to a
 subtyping relation. The meta-variable $\tau$ denotes an arbitrary type. 
@@
    &
-
-   \inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'\\ \Pi;\Gamma'\vdash_{\text{block}} \iota_2}
-             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[block-seq]\\ \\
+   \inference{}
+             {\Pi\vdash_{\text{blocks}} \textbf{end}}[empty]\\ \\
 
@@
    &
-   \inference{}
-             {\Pi\vdash_{\text{blocks}} \textbf{end}}[empty] \\ \\
+
+   \inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'\\ \Pi;\Gamma'\vdash_{\text{block}} \iota_2}
+             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[block-seq]
+ \\ \\
 
@@
 And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label.
-We use the \ensuremath{\V{All}} datatype to express this relation.
+We use the \ensuremath{\D{All}} datatype to express this relation.
 
@@
 \\
-\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
+\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
 \ColumnHook
@@
 
-We use the datatype \ensuremath{\V{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
+We use the datatype \ensuremath{\D{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
 variable environments and well-typed programs. Thus, intuitively, \ensuremath{\D{Env}} is like a
@@
 \\
-\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\V{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
+\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\D{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
 \ColumnHook
@@
 both related to the \emph{step-fetch-field-0} rule~\cite{Appel07}. The first retrieves the head element of a list, and
-stores it in a new variable. The \ensuremath{\V{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
+stores it in a new variable. The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
 environment \ensuremath{\V{env}}, and this variable is added to the result environment. The typed \emph{de Bruijn}
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathkw{with}\;\V{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
 \\
@@
 types, we need some extra lemmas. First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the
-result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\V{lookup}}
+result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\F{lookup}}
 function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\V{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
+\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\F{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\V{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
+\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
 \ColumnHook
@@
 1.  & Operational Semantics                        & 126                       & 98                      & 109                      \\
-2.  & Derive $p \Downarrow$                        & 1                         & 8                       & -                        \\ \hline
+2.  & Derive $p \Downarrow$                        & 1                         & 8                       & 1                        \\ \hline
 3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
@@
 5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
-                                                   & 1                         & no                      & -                        \\
+                                                   & 1                         & no                      & 1                        \\
 6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
@@
   \item \textbf{Prove properties of the least common supertype.} Proofs of all proved properties about the least common supertype
-        are simple recursive functions over the least common supertype and subtyping relations definition.
+        are simple recursive functions over the least common supertype and subtyping relations definitions and were omitted
+        from this text for space reasons.
   \item \textbf{State the soundness theorem for the type system.} Following~\citet{Amin17}, we represent the soundness
         theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list-machine
-        programs.
+        programs. Therefore, this was done as part of task 9.
   \item \textbf{Prove the soundness theorem for the type system.} Following \citet{Amin17}, the proof of the soundness
@@
 
-This paper shows the combination of intrinsically-typed syntax and definitional interpreters to formalize programming languages.
-Using such approach, we were able to provide a machine checked version of the list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
-The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM, since we were able to encode features such as jumps, mutable state, and sub-typing.
-When comparing such approach with the conventional syntactic way to prove properties, we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.
-This happens because the approach uses the power of the host language, and provides some proofs for free due to the intrinsically-typed syntax.
-
-As future work, we want to reuse the ideas presented in this paper to provide an intrinsically-typed formalization for real-world low-level languages like the eBPF and the LUA VM.
-Furthermore, one can extend the formalization presented here for other programming languages with similar settings.
+This paper shows how the combination of intrinsically-typed syntax and definitional interpreters can be used to simplify
+programming languages formalization tasks. Using such approach, we were able to provide a machine checked version of the
+list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
+The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM,
+since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing such our work with the
+conventional non-dependently typed formalization strategies (like to ones used by Appel et. al. in his Coq implementation),
+we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.
+This happens because the approach uses the power of the host language, and provides some proofs for free due to the
+intrinsically-typed syntax.
+
+As future work, we want to reuse the ideas presented in this paper to provide an intrinsically-typed formalization for
+real-world low-level languages like the eBPF and the LUA VM. Furthermore, one can extend the formalization presented here
+for other programming languages with similar settings.
 

### 2020-06-30 papers/SBLP2020/paper.tex
Added System F work

@@
 does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
-of linear and session types as in~\cite{Rouvoet20,Thiemann19}.}
+of linear and session types as in~\cite{Rouvoet20,Thiemann19}. A formalization of System F$_{\omega\mu}$ was
+the subject of Chapman et.al. work~\cite{ChapmanKNW19}, which used an intrinsically-typed representation to 
+implement a normalization by evaluation for this calculus. The reason behind such formalization effort was
+the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$.}
 

### 2020-06-30 papers/SBLP2020/paper.tex
Everything fitting in 8 pages

@@
 
-\title{An Intrinsically Typed Solution for the List-Machine Benchmark}
+\title{An Intrinsically-Typed Solution for the List-Machine Benchmark}
 
@@
 \begin{abstract}
-Formal models are important tools in the programming languages research
-community. However, such models are full of intrincacies and, due to that,
+Formal models are important tools in the programming language research
+community. However, such models are full of intricacies and, due to that,
 they are subject to subtle errors. Such failures motivated the usage of
@@
 is the possibility to validate the semantics through execution.
-Recently, definitional interpreters were used to formalize type
-soundness theorems for some advanced typing features~\cite{Amin17}
-and the semantics of imperative programming languages in which static
-semantics is ensured by dependently-typed
+Recently, definitional interpreters were used to formalizing type
+soundness theorems for some advanced typing features~\cite{Amin17},
+and the semantics of imperative programming languages, in such a way the static
+semantics is ensured by a dependently-typed
 syntax\footnote{Also known as intrinsically-typed.}\cite{Poulsen18}.
-In this work, we follow Poulsen et. al. and use an intrinsically-typed
+In this work, we follow \citet{Poulsen18} by using an intrinsically-typed
 representation to build a definitional interpreter for a low-level virtual
-machine developed by Appel and Leroy as a benchmark problem closer to
+machine, developed by Appel and Leroy, as a benchmark problem closer to
 real-world implementations such as typed assembly languages~\cite{CraryM99} and
-proof carrying code~\cite{Necula97}.
+proof-carrying code~\cite{Necula97}.
 
@@
   \item We provide a provably correct implementation for testing the subtyping
-        relation and to calculate the least common super type of two input
+        relation and to calculate the least common supertype of two input
         types for the machine registers.
-  \item We provide a correct-by-construction implementation of a type checker for
-        list-machine programs. Our type checker produces, as result, an intrinsically-typed
+  \item We provide a correct-by-construction implementation of a type-checker for
+        list-machine programs. Our type-checker produces, as a result, an intrinsically-typed
         representation of the machine code.
@@
 
-The rest of this paper is organized as follows. Section~\ref{sec:agda}
-presents a brief introduction to Agda and Section~\ref{sec:list}
+The rest of this paper is organized as follows: Section~\ref{sec:agda}
+presents a brief explanation about the encoding of dependently-typed syntax in Agda and Section~\ref{sec:list}
 reviews the list-machine benchmark and presents its syntax and type system.
-We describe the intrinsically-typed representation for the list-machine
-in Section~\ref{sec:typing}. Section~\ref{sec:semantics} briefly discuss
-the list-machine semantics and then proceeds with its
-realization as a definitional interpreter.
-The type-checking algorithm, the subtyping relation and the least-common-supertype
-algorithm are presented in Section~\ref{sec:typechecker}. Section~\ref{sec:comparison}
+We describe the intrinsically-typed representation for the list-machine, as well as
+the subtyping relation, and the least common supertype algorithm in Section~\ref{sec:typing}.
+Section~\ref{sec:semantics} briefly discuss the list-machine semantics and
+then proceeds with its realization as a definitional interpreter.
+The type-checking algorithm is presented in Section~\ref{sec:typechecker}. Section~\ref{sec:comparison}
 compares the presented formalization in Agda with Coq and Twelf encodings. Related work is
@@
 constructor, leaving the responsibility of checking type safety to the
-host's language type checker. As an example, consider the following simple
+host's language type-checker. As an example, consider the following simple
 expression language. For this paper, we assume a basic knowledge of functional
-programming and Agda\footnote{For further information about Agda,
-see~\cite{Norell2009,Stump16}.}.
+programming and Agda. %\footnote{For further information about Agda, see~\cite{Norell2009,Stump16}.}.
 
@@
 %A definition which uses the expression |(Num 1) + True| will be rejected by Agda's type checker automatically.
-Agda's type checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\V{+}\;\Con{True}}.
+Agda's type-checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\V{+}\;\Con{True}}.
+
+For further information about Agda, see~\cite{Norell2009,Stump16}.
 
@@
 
-Each program variable is assigned to a list type, which is used to guarantee the safety of execute
-fetch-field that demands non-empty list arguments. In order to express such refinements, types are subject to a
+Each program variable is assigned to a list type, which is used to guarantee the safety when executing
+fetch-field instructions that demands non-empty list arguments. In order to express such refinements, types are subject to a
 subtyping relation. The meta-variable $\tau$ denotes an arbitrary type. 
@@
 \]
-We let notation $\tau \subset \tau'$
-denote the subtyping judgement which is defined as follows.
+The notation $\tau \subset \tau'$
+denotes the subtyping judgment, which is defined as follows.
 \[
@@
 subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
-The other rules specifies that type constructors $list$ and $listcons$ respect
+The other rules specify that type constructors $list$ and $listcons$ respect
 the subtyping relation. The list common supertype $\tau = \tau_1 \sqcap \tau_2$ of
@@
 Using the previously defined notations, the typing rules for the list-machine are defined
-as a syntax directed judgement $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
+as a syntax directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
 which intuitively means that the instruction $\iota$ transforms an input typing environment
@@
 \]
-Inspired by the presented typing rules, in the next section, we define an intrisinsically
-typed syntax for list-machine programs which ensures that only well-typed programs can be built.
-
-\section{Instrinsically-typed syntax}\label{sec:typing}
+Inspired by the presented typing rules, in the next section, we define an
+intrinsically-typed syntax for list-machine programs which ensures that
+only well-typed programs can be built.
+
+\section{Intrinsically-typed syntax}\label{sec:typing}
 
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{nil}\;{}\<[17]%
-\>[17]{}\mathbin{:}\;\D{Ty}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{list}\;\Con{listcons}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
+\>[3]{}\Con{nil}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{list}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{listcons}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
 \ColumnHook
@@
 We internalize the list-machine type judgments for blocks and instructions in Agda
-together with its syntax in such a way where only well-typed terms that satisfying typing
+together with its syntax in such a way where only well-typed terms that satisfy typing
 judgments have meaning. This approach makes the AST contain both syntactic and semantic
@@
 
-In order to be considered well-typed, list-machine machine programs need to refer
+In order to be considered well-typed, list-machine programs need to refer
 to information from two sources: (1) a type context encoded as a list to store
@@
 \paragraph{Conditional jump}{There are three constructors (\ensuremath{\Con{instr\textrm{-}branch\textrm{-}list}},
-\ensuremath{\Con{instr\textrm{-}branch\textrm{-}listcons}}, and \ensuremath{\Con{instr\textrm{-}branch\textrm{-}nil}}) related to a conditional
+\ensuremath{\Con{instr\textrm{-}branch\textrm{-}listcons}}, and \ensuremath{\Con{instr\textrm{-}branch\textrm{-}nil}}) related to conditional
 jump. They are used to perform a jump to a label \ensuremath{\V{l}} when the received variable is \ensuremath{\Con{nil}}.
@@
 
-A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}) which is
+A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}), which is
 exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
 space reasons, since it is just an inductive type that encode the rules presented
-in Section~\ref{sec:list}. Our code repository~\cite{list-rep} defines several lemmas about
-subtyping including its decidable test.
+in Section~\ref{sec:list}. Several lemmas about subtyping, including its decidable test,
+are defined in our source-code repository~\cite{list-rep}.
 
@@
 variable type whenever it is updated by a \emph{cons} instruction. The rules of the
-super type relation are specified as the following inductive type:
+supertype relation are specified as the following inductive type:
 
@@
 \end{hscode}\resethooks
-Constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
-types. Also, rules ensure that the relation is commutative as proved by a simple
+The presented constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
+types. Also, we have a rule to ensure that the relation is commutative, as proved by a simple
 Agda theorem.
@@
 returned type \ensuremath{\V{t}} is indeed the least common supertype of \ensuremath{\V{t1}} and \ensuremath{\V{t2}},
-thus ensuring its correctness by construction.
+thus ensuring its correctness-by-construction.
 
@@
 definitional interpreter~\cite{Reynolds72}, which evaluates an intrinsically-typed instruction,
-transforming a initial memory state into a new one, represented by a run-time environment.
+transforming an initial memory state into a new one, represented by a run-time environment.
 
@@
 i.e., each program developed in it must terminate and all possible patterns must
-be matched. However, by using the mechanisms for jumping between labels, one can
+be matched. However, by using the mechanisms for jumping between labels, one could
 write a program which never ends, making it impossible to implement a terminating
@@
 last one is the \ensuremath{\D{Block}} to be evaluated. This function returns a modified run-time
-environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or nothing when (and only) the \emph{fuel} runs out.
+environment (\ensuremath{\D{Env}\;\V{\Gamma'}}) in case of success, or \ensuremath{\Con{nothing}} when (and only) the \emph{fuel} runs out.
 
@@
 types, we need some extra lemmas. First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the
-result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\F{lookup}}
+result type of this list is the least common supertype between these two. As before, we use the \ensuremath{\F{lookup}}
 function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we
@@
 type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}},
-and others lemmas to deal with the least common subtying, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
+and others lemmas to deal with the least common supertype, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
 and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}. These lemmas and their proofs can be found in our repository online. }
@@
 It is worth noticing that we do not have any error treatment on this interpreter function, except for
-when we ran out-of-fuel. Since we are using an intrinsically-typed syntax, only valid instructions
+when we ran out-of-fuel. Since we are using an intrinsically-typed syntax, only valid (well-typed) instructions
 are accepted in each step of evaluation.
@@
 
-\section{Type Checker}\label{sec:typechecker}
+\section{Type-Checker}\label{sec:typechecker}
 
 In practice, a source-code of a programming language runs through several phases, including lexing, parsing,
-scope checking, and most importantly \emph{type checking}. Since we represent programs using a
-intrinsically-typed syntax, scope and type checking is only a matter of elaborating an untyped syntax
+scope checking, and most importantly \emph{type-checking}. Since we represent programs using a
+intrinsically-typed syntax, scope and type-checking is only a matter of elaborating an untyped syntax
 to a typed one.
@@
 check them, we need to provide an index from a named variable. This process is done by the
-\ensuremath{\F{lookup\textrm{-}var}} function, returning an index when the named variable variable is found, and \ensuremath{\Con{nothing}} otherwise.
+\ensuremath{\F{lookup\textrm{-}var}} function, returning an index when the named variable is found, and \ensuremath{\Con{nothing}} otherwise.
+
+\pagebreak
 
@@
 \\
-\>[B]{}\V{...}\;\mid \;\V{yes}\;\V{refl}\;\mathrel{=}\;\Con{just}\;(\V{\tau}\;\red{,}\,\;\Con{here}\;\V{refl}){}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;(\V{no}\;\V{¬p})\;\mathrel{=}\;\F{map}\;{}\<[22]%
+\>[B]{}\V{...}\;\mid \;\Con{yes}\;\Con{refl}\;\mathrel{=}\;\Con{just}\;(\V{\tau}\;\red{,}\,\;\Con{here}\;\Con{refl}){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;(\Con{no}\;\V{¬p})\;\mathrel{=}\;\F{map}\;{}\<[22]%
 \>[22]{}(\lambda\;\V{v}\;\rightarrow\;\D{proj_1}\;\V{v}\;\red{,}\,\;\Con{there}\;(\D{proj_2}\;\V{v}))\;{}\<[E]%
@@
 
-For space reasons, we show how we type check only one instruction\footnote{The curious reader can refer to our online repository for a complete implementation.}.
-Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
+For space reasons, we show how we type-check only one instruction\footnote{The curious reader can refer to our
+online repository for a complete implementation.}. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
 receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value,
-which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type checks.
+which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type-checks.
 
@@
 
-We implemented the 14 tasks from the list-machine benchmark in the Agda programming language.
+We implemented all 14 tasks from the list-machine benchmark in the Agda programming language.
 The tasks considered by us are the same implemented and proved by~\citet{Appel07}.
@@
 \begin{enumerate}
-  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we chose to
-        use a definitional interpreter for our intrinsincally-typed representation. Our implementation for the operational
-        semantics is composed by 38 lines for the type syntax and 71 for the definitional interpreter.
-  \item \textbf{Derive $p \Downarrow$.} Since we used a definitional intepreter for representing the semantics, we can derive
+  \item \textbf{Operational semantics.} Instead of using an inductive type for representing the operational semantics, we chose to
+        use a definitional interpreter for our intrinsically-typed representation. Our implementation for the operational
+        semantics is composed by 38 lines for the typed syntax and 71 for the definitional interpreter.
+  \item \textbf{Derive $p \Downarrow$.} Since we used a definitional interpreter for representing the semantics, we can derive
         $p\,\Downarrow$ just by executing the interpreter on $p$.
@@
         guarantee its soundness w.r.t. its specification.
-  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list-machine program
-        is just execute the type checker over it. Since our type checker returns an intrinsincally-typed representation of
+  \item \textbf{Derive an example of type-checking.} Our approach to build a derivation for a sample list-machine program
+        is just execute the type-checker over it. Since our type-checker returns an intrinsically-typed representation of
         the input and such typed syntax is equivalent to the type system, it can be used to build type derivations.
@@
         produces, as result, an environment of well-typed values resulting from the execution of the input program.
-  \item \textbf{Asymptotically efficient algorithm.} Our current implementation of the typechecker takes quadratic time. The reason
-        for this inneficiency is the representation of environments as lists / vectors. The use of better data structures (like finite
+  \item \textbf{Asymptotically efficient algorithm.} Our current implementation of the type-checker takes quadratic time. The reason
+        for this inefficiency is the representation of environments as lists / vectors. The use of better data structures (like finite
         maps implemented by efficient search trees) can solve this issue. We leave this fix to future work.
-  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it is only a matter of calling the typechecker
+  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it is only a matter of calling the type-checker
         on the input program.
-  \item \textbf{Prove the termination of the type checker.} This task is trivial in our setting, since all defined Agda functions
+  \item \textbf{Prove the termination of the type-checker.} This task is trivial in our setting, since all defined Agda functions
         must be total. The totality is ensured by Agda's termination / totality checker.
-  \item \textbf{Scalable type-checker.} Agda code can be compiled to machine code using its GHC-Haskell backend. Since GHC is
-        an industrial strength compiler, the backend can generate an efficient executable for the machine interpreter and typechecker. 
-  \item \textbf{Prove soundness of type-checker.} In our approach, the soundness of the typechecker is ensured by construction,
+  \item \textbf{Scalable type-checker.} Agda code can be compiled to machine code using its GHC-Haskell back-end. Since GHC is
+        an industrial strength compiler, the back-end can generate an efficient executable for the machine interpreter and type-checker. 
+  \item \textbf{Prove soundness of type-checker.} In our approach, the soundness of the type-checker is ensured by construction,
         since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
@@
 
-As we could notice on the table above, the approach taken in this paper avoids code repetition and decreases the number of LOCs necessary to implement the proposed language and prove its properties.
-The formalization of this benchmark took 2898 LOC to be done in Twelf, and 887 LOC in Coq, while our implementation used only 413 LOC to achieve the same result.
-Our encoding uses approximately 14\% of the Twelf LOC, and 47\% of Coq's, mostly due to the fact the soundness properties are obtained for free when combining an intrinsically-typed syntax with an implementation of a definitional interpreter in a total language like Agda.
+As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel's et. al. solution.
+Our implementation used 415 LOC to complete the tasks, while the Twelf solution demanded 2898 LOC and 887 LOC in Coq.
+Our encoding uses approximately 14\% of the LOC when compared to the Twelf formalization, and 47\% when compared to Coq's. The main
+reason for this difference is that our intrinsically-typed syntax granted us many properties for free (e.g. type soundness).
+
+%As we could notice on the previous table, the approach taken in this paper avoids code repetition and decreases the number of LOCs necessary to
+%implement the proposed language and to prove its properties. The formalization of this benchmark took 2898 LOC to be done in Twelf, and 887 LOC
+%in Coq, while our implementation used only 415 LOC to achieve the same result. Our encoding uses approximately 14\% of the LOC when compared to
+%the Twelf formalization, and 47\% when compared to Coq's. The main reason for this difference is the fact that the soundness properties are
+%obtained for free when combining an intrinsically-typed syntax with an implementation of a definitional interpreter in a total language like Agda.
 
@@
 \paragraph{Benchmarks for PL mechanization}{
-The importante of benchmarks is recognized in several areas in computer science. A relevent example is the
-TPTP library for the theorem proving community~\cite{Sut17}. In the context of programming languages community,
-the POPLMark challenge~\cite{Aydemir05} was developed by a renowned group of programming language researchers aiming
+The importance of benchmarks is recognized in several areas in computer science. A relevant example is the
+TPTP library for the theorem proving community~\cite{Sut17}. In the context of the programming languages community,
+we have the POPLMark challenge~\cite{Aydemir05}, which was developed by a renowned group of programming language researchers aiming
 the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
@@
 The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
-Recently, the interest on such interpreters was revitalized by~\citet{Amin17} which used definitional
+Recently, the interest on such interpreters was revitalized by~\citet{Amin17}, which used definitional
 interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
@@
 their formalization with the need of dealing with ``stuck states'' in the semantics. \citet{Poulsen18} described how
-definitional interpreters for imperative languages can be much concisely implemented by using intrisincally-typed syntax by
+definitional interpreters for imperative languages can be much concisely implemented by using intrinsically-typed syntax by
 using a library for name binding using \emph{scope graphs}, which greatly simplifies the treatment of complex
 binding structures. Since the list-machine benchmark was designed to address other problems than binding, our representation
-using \emph{de Bruijn} indices was sufficient to implement the desired type checker and interpreter.
+using \emph{de Bruijn} indices was sufficient to implement the desired type-checker and interpreter.
 Other recent application of definitional interpreters was proposed by~\citet{Rouvoet20}. The main contribution
@@
 linear typing features in the interpreter, the authors have implemented monads based on a separation algebra which they
-named \emph{Market}, supporting the main operations for accessing and update the store used by the intepreter.
-Another work using intrisincally-typed syntax for resource control was developed by~\citet{Thiemann19}, which
+named \emph{Market}, supporting the main operations for accessing and update the store used by the interpreter.
+Another work using intrinsically-typed syntax for resource control was developed by~\citet{Thiemann19}, which
 implemented an interpreter for a more realistic core functional session typed calculus including recursion and
@@
 implement a normalization by evaluation for this calculus. The reason behind such formalization effort was
-the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$.}
+the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$. As our work,
+Chapman's et. al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
+which avoids completely stuck states.}
 
@@
 This paper shows how the combination of intrinsically-typed syntax and definitional interpreters can be used to simplify
-programming languages formalization tasks. Using such approach, we were able to provide a machine checked version of the
+the tasks on the formalization of programming languages. Using such approach, we were able to provide a machine-checked version of the
 list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
 The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM,
-since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing such our work with the
-conventional non-dependently typed formalization strategies (like to ones used by Appel et. al. in his Coq implementation),
+since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing our work with the
+conventional non-dependently typed formalization strategies (like the one used by \citet{Appel07} in their Coq and Twelf implementations),
 we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.

### 2020-08-29 papers/SBLP2020/paper.tex
Camera ready

@@
 is the possibility to validate the semantics through execution.
-Recently, definitional interpreters were used to formalizing type
+Recently, definitional interpreters were used for formalizing type
 soundness theorems for some advanced typing features~\cite{Amin17},
@@
 
-More specifically, we contribute:
+More specifically, our contributions are:
 
@@
 the subtyping relation, and the least common supertype algorithm in Section~\ref{sec:typing}.
-Section~\ref{sec:semantics} briefly discuss the list-machine semantics and
+Section~\ref{sec:semantics} briefly discusses the list-machine semantics and
 then proceeds with its realization as a definitional interpreter.
@@
   \iota & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
-        & \mid & \text{branch-if-nil $v$}            & \text{(if $v = nil$ goto $l$)}\\
+        & \mid & \text{branch-if-nil $v$~$l$}            & \text{(if $v = nil$ goto $l$)}\\
         & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
@@
 Following common practice, the meta-variable $\Gamma$ denotes an
-environment between variable names and
-its types. Notation $\{\}$ denote an empty environment, $v : \tau , \Gamma$
-denote the operation of including a new entry for variable $v$ with type $\tau$
+environment binding variable names to their types. Notation $\{\}$ denotes an empty environment, $v : \tau , \Gamma$
+denotes the operation of including a new entry for variable $v$ with type $\tau$
 in $\Gamma$ and $\Gamma [v := \tau]$ denotes the environment which is identical to $\Gamma$, except
@@
 We internalize the list-machine type judgments for blocks and instructions in Agda
-together with its syntax in such a way where only well-typed terms that satisfy typing
+together with its syntax in such a way that only well-typed terms that satisfy typing
 judgments have meaning. This approach makes the AST contain both syntactic and semantic
@@
 \\
-\>[B]{}\D{Ctx}\;\mathrel{=}\;\D{List}\;\D{Ty}{}\<[E]%
+\>[B]{}\D{Ctx}\;\mathrel{=}\;\D{List}\;(\D{String}\;\V{×}\;\D{Ty}){}\<[E]%
 \\[\blanklineskip]%
@@
 
-As we saw in the previous section, the typing rules of the list-machine language were split
-in two segments, one for instructions and one for blocks. We defined two datatypes (\ensuremath{\D{\_\vdash\_\Rightarrow\_}} and \ensuremath{\D{Block}})
+As we saw in the previous section, the typing rules for the list-machine language were split
+into two segments, one for instructions and one for blocks. We define two datatypes (\ensuremath{\D{\_\vdash\_\Rightarrow\_}} and \ensuremath{\D{Block}})
 to hold the well-typed terms accordingly, representing each judgment of the static semantics as a
@@
 
-\paragraph{Sequencing instructions}{The constructor \ensuremath{\Con{instr\textrm{-}seq}} can be used to
+\paragraph{Sequencing instructions}{The constructor \ensuremath{\Con{instr\textrm{-}seq}} is used to
 express a sequence of instructions. From the execution of two instructions, it
@@
 \paragraph{Conditional jump}{There are three constructors (\ensuremath{\Con{instr\textrm{-}branch\textrm{-}list}},
-\ensuremath{\Con{instr\textrm{-}branch\textrm{-}listcons}}, and \ensuremath{\Con{instr\textrm{-}branch\textrm{-}nil}}) related to conditional
+\ensuremath{\Con{instr\textrm{-}branch\textrm{-}listcons}}, and \ensuremath{\Con{instr\textrm{-}branch\textrm{-}nil}}) for expressing a conditional
 jump. They are used to perform a jump to a label \ensuremath{\V{l}} when the received variable is \ensuremath{\Con{nil}}.
@@
 
-\paragraph{Fetching information from list}{There are four constructors which can be used
-to fetch information from a given list. The constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}} receives a
+\paragraph{Fetching information from list}{There are four constructors for expressing
+information fetching from a given list. The constructor \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}} receives a
 non-empty list (of type \ensuremath{\Con{listcons}\;\V{\tau}}), and is used to retrieve the head of this list and store it
@@
 The resulting typing context adds information about the newly created list.
-Subsection~\ref{sec:supertype} will provide details of an algorithm
+Subsection~\ref{sec:supertype} provides the details of an algorithm
 for calculating the least common supertype of two given types.}
@@
 similar meaning to instruction sequence, and \ensuremath{\Con{block\textrm{-}jump}} is used to perform a direct jump
-(without any condition), receiving a label and checking if the current context typing is
+(without any condition), receiving a label and checking if the current context typing is a
 subtype of the intended one.
@@
 exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
-space reasons, since it is just an inductive type that encode the rules presented
+space reasons, since it is just an inductive type that encodes the rules presented
 in Section~\ref{sec:list}. Several lemmas about subtyping, including its decidable test,
@@
 This section describes the steps to evaluate a program written using the list-machine language.
-We adapted the small-step semantics presented in Section~\ref{sec:list}, transforming it in a
+We adapted the small-step semantics presented in Section~\ref{sec:list}, transforming it into a
 definitional interpreter~\cite{Reynolds72}, which evaluates an intrinsically-typed instruction,
@@
 The tasks considered by us are the same implemented and proved by~\citet{Appel07}.
-The next table summarize the total number of lines of code (LOC) for our results together with theirs.
+The next table summarizes the total number of lines of code (LOC) for our results together with theirs.
 
@@
   \item \textbf{Derive an example of type-checking.} Our approach to build a derivation for a sample list-machine program
-        is just execute the type-checker over it. Since our type-checker returns an intrinsically-typed representation of
+        is just to execute the type-checker over it. Since our type-checker returns an intrinsically-typed representation of
         the input and such typed syntax is equivalent to the type system, it can be used to build type derivations.
@@
 
-As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel's et. al. solution.
+As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to \citet{Appel07} solution.
 Our implementation used 415 LOC to complete the tasks, while the Twelf solution demanded 2898 LOC and 887 LOC in Coq.
@@
 we have the POPLMark challenge~\cite{Aydemir05}, which was developed by a renowned group of programming language researchers aiming
-the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
+at the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
 was to motivate authors to formalize all of their theorems using such tools. Since the focus of the POPLMark challenge
@@
 
-\paragraph{Definitional intepreters}{
+\paragraph{Definitional interpreters}{
 The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
@@
 interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
-Unlike our work, Amin and Rompf's formalizations does not use intrinsically-typed representations of their syntax which cluttered
+Unlike our work, Amin and Rompf's formalizations do not use intrinsically-typed representations of their syntax, what cluttered
 their formalization with the need of dealing with ``stuck states'' in the semantics. \citet{Poulsen18} described how
-definitional interpreters for imperative languages can be much concisely implemented by using intrinsically-typed syntax by
-using a library for name binding using \emph{scope graphs}, which greatly simplifies the treatment of complex
+definitional interpreters for imperative languages can be much concisely implemented by using intrinsically-typed syntax and a
+library for name binding using \emph{scope graphs}, which greatly simplifies the treatment of complex
 binding structures. Since the list-machine benchmark was designed to address other problems than binding, our representation
@@
 linear typing features in the interpreter, the authors have implemented monads based on a separation algebra which they
-named \emph{Market}, supporting the main operations for accessing and update the store used by the interpreter.
+named \emph{Market}, supporting the main operations for accessing and updating the store used by the interpreter.
 Another work using intrinsically-typed syntax for resource control was developed by~\citet{Thiemann19}, which
@@
 session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
-simple interface to a scheduler. Since our objective was formalizing the list-machine benchmark in Agda and it
+simple interface to a scheduler. Since our objective was to formalize the list-machine benchmark in Agda and it
 does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
 of linear and session types as in~\cite{Rouvoet20,Thiemann19}. A formalization of System F$_{\omega\mu}$ was
-the subject of Chapman et.al. work~\cite{ChapmanKNW19}, which used an intrinsically-typed representation to 
+the subject of \citet{ChapmanKNW19} work, which used an intrinsically-typed representation to 
 implement a normalization by evaluation for this calculus. The reason behind such formalization effort was
 the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$. As our work,
-Chapman's et. al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
+Chapman's et al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
 which avoids completely stuck states.}
+
+\vspace{-3ex}
 
@@
 conventional non-dependently typed formalization strategies (like the one used by \citet{Appel07} in their Coq and Twelf implementations),
-we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.
+we can affirm that it requires a fewer number of lines to achieve the same result, even without the use of proof automation.
 This happens because the approach uses the power of the host language, and provides some proofs for free due to the

### 2021-03-07 papers/COLA/paper.tex
Starting the COLA paper

--- 2020-08-29 papers/SBLP2020/paper.tex
@@
-\documentclass[sigconf]{acmart}
+\documentclass[review]{elsarticle}
+
+\usepackage{lineno,hyperref}
+\modulolinenumbers[5]
 
@@
 \usepackage{proof}
-
-% Copyright
-%\setcopyright{none}
-%\setcopyright{acmcopyright}
-\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2020}
-\acmYear{2020}
-
-\acmConference[SBLP '20]{SBLP '20: Brazilian Symposium on Programming Languages}{October 19--23, 2020}{Natal, Brazil}
-\acmBooktitle{SBLP '20: Brazilian Symposium on Programming Languages,
-  October 19--23, 2020, Natal, Brazil}
 
@@
 
+\journal{Journal of Computer Languages}
+
+%% `Elsevier LaTeX' style
+\bibliographystyle{elsarticle-num}
+%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
 
-%%%%%%%%%%%%%%%%%%%%%%%
-%% Title and authors %%
-%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{An Intrinsically-Typed Solution for the List-Machine Benchmark}
+\begin{frontmatter}
+
+\title{An Intrinsically-Typed Solution for the\\ List-Machine Benchmark}
 
 \author{Samuel Feitosa}
-\authornotemark[1]
-\email{samuel.feitosa@ifsc.edu.br}
-\affiliation{%
-  \institution{Departamento de Informática}
-  \streetaddress{Instituto Federal de Santa Catarina}
-  \city{Caçador}
-  \state{Santa Catarina}
-  \country{Brazil}
+\ead{samuel.feitosa@ifsc.edu.br}
+\address{%
+  Departamento de Informática - Instituto Federal de Santa Catarina \\
+  Caçador - Santa Catarina - Brazil
 }
@@
 \author{Rodrigo Ribeiro}
-\email{rodrigo.ribeiro@ufop.edu.br}
-\affiliation{%
-  \institution{Prog. Pós Graduação em Ciência da Computação}
-  \streetaddress{Universidade Federal de Ouro Preto}
-  \city{Ouro Preto}
-  \state{Minas Gerais}
-  \country{Brazil}}
+\ead{rodrigo.ribeiro@ufop.edu.br}
+\address{
+  Prog. Pós Graduação em Ciência da Computação - Universidade Federal de Ouro Preto \\
+  Ouro Preto - Minas Gerais - Brazil}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ACM codes for areas %%
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{CCSXML}\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\V{ccs2012>}{}\<[E]%
-\\
-\>[B]{}\V{concept>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 id>10011007.10011006.10011039.10011311</concept\char95 id>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 desc>Software}\;\V{and}\;\V{its}\;\V{engineering\char126 Semantics</concept\char95 desc>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 significance>500</concept\char95 significance>}{}\<[E]%
-\\
-\>[B]{}\V{/concept>}{}\<[E]%
-\\
-\>[B]{}\V{concept>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 id>10011007.10011006.10011041.10010943</concept\char95 id>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 desc>Software}\;\V{and}\;\V{its}\;\V{engineering\char126 Interpreters</concept\char95 desc>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 significance>500</concept\char95 significance>}{}\<[E]%
-\\
-\>[B]{}\V{/concept>}{}\<[E]%
-\\
-\>[B]{}\V{concept>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 id>10003752.10003790.10011740</concept\char95 id>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 desc>Theory}\;\V{of}\;\V{computation\char126 Type}\;\V{theory</concept\char95 desc>}{}\<[E]%
-\\
-\>[B]{}\V{concept\char95 significance>500</concept\char95 significance>}{}\<[E]%
-\\
-\>[B]{}\V{/concept>}{}\<[E]%
-\\
-\>[B]{}\V{/ccs2012>}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[500]{Software and its engineering~Interpreters}
-\ccsdesc[500]{Theory of computation~Type theory}
-
-\keywords{Dependent types, formal semantics}
-
-\maketitle
+\begin{keyword}
+Dependent types, formal semantics
+\end{keyword}
+
+\end{frontmatter}
+
+\linenumbers
 
@@
 syntax\footnote{Also known as intrinsically-typed.}\cite{Poulsen18}.
-In this work, we follow \citet{Poulsen18} by using an intrinsically-typed
+In this work, we follow Poulsen et. al. by using an intrinsically-typed
 representation to build a definitional interpreter for a low-level virtual
@@
 proof-carrying code~\cite{Necula97}.
+
+In this work, we extend our previous SBLP 2020 paper by including support to
+indirect jumps, as specified in the version 2.0 of the list
+machine benchmark~\cite{AppelDL12}. We provide detailed explanations of the needed
+changes on the machine syntax, type system and interpreter to accomodate this extension, which
+allows the handling of more complex control flow, like the call/return model of function calls.
 
@@
         unnecessary repetitions and provides some properties for free.
+  \item We provide a detailed discussion on the necessary type system and interpreter modifications
+        to support indirect jumps.
 \end{itemize}
 
-The rest of this paper is organized as follows: Section~\ref{sec:agda}
-presents a brief explanation about the encoding of dependently-typed syntax in Agda and Section~\ref{sec:list}
+The rest of this paper is organized as follows: Section~\ref{sec:agda} presents an overview of
+the Agda programming language and Section~\ref{sec:syntax}
+gives a brief explanation about the encoding of dependently-typed syntax. Section~\ref{sec:list}
 reviews the list-machine benchmark and presents its syntax and type system.
@@
 
-\section{Dependently-typed Syntax}\label{sec:agda}
-
-% Agda is a dependently-typed functional programming language based on
-% Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
-% and an infinite hierarchy of types of types, |Set l|, where |l| is a
-% natural number, are built-in. Everything else is a user-defined
-% type. The type |Set|, also known as |Set0|, is the type of all
-% ``small'' types, such as |Bool|, |String| and |List Bool|.  The type
-% |Set1| is the type of |Set| and ``others like it'', such as |Set ->
-% Bool|, |String -> Set|, and |Set -> Set|. We have that |Set l| is an
-% element of the type |Set (l+1)|, for every $l \geq 0$. This
-% stratification of types is used to keep Agda consistent as a logical
-% theory~\cite{Sorensen2006}.
-
-% An ordinary (non-dependent) function type is written |A -> B| and a
-% dependent one is written |(x : A) -> B|, where type |B| depends on
-% |x|, or |forall (x : A) -> B|. Agda allows the definition of \emph{implicit
-% parameters}, i.e.,  parameters whose values can be inferred from the
-% context, by surrounding them in curly braces: |forall {x : A} -> B|. To
-% avoid clutter, we'll omit implicit arguments from the source code
-% presentation. The reader can safely assume that every free variable in
-% a type is an implicit parameter.
-
-% As an example of Agda code, consider the following data type of
-% length-indexed lists, also known as vectors.
-
-% \begin{spec}
-%   data Nat : Set where
-%     zero : Nat
-%     succ : Nat -> Nat
-
-%   data Vec (A : Set) : Nat -> Set where
-%     []  : Vec A zero
-%     _::_ : forall {n} -> A -> Vec A n -> Vec A (succ n)
-% \end{spec}
-% %format head = "\F{head}"
-% Constructor |[]| builds empty vectors. The cons-operator (|_::_|)
-% inserts a new element in front of a vector of $n$ elements (of type
-% |Vec A n|) and returns a value of type |Vec A (succ n)|. The
-% |Vec| datatype is an example of a dependent type, i.e., a type
-% that uses a value (that denotes its length). The usefulness of
-% dependent types can be illustrated with the definition of a safe list
-% head function: |head| can be defined to accept only non-empty
-% vectors, i.e.,~values of type |Vec A (succ n)|.
-% \begin{spec}
-%   head : Vec A (succ n) -> A
-%   head (x :: xs) = x
-% \end{spec}
-% In |head|'s definition, constructor |[]| is not used. The
-% Agda type-checker can figure out, from |head|'s parameter type,
-% that argument |[]| to |head| is not type-correct.
-
-% %format _==_ = "\D{\_ \equiv \_}"
-% %format == = "\D{\equiv}"
-% %format refl = "\Con{refl}"
-% %format proj₁ = "\F{\pi_1}"
-% %format proj₂ = "\F{\pi_2}"
-% %format Fin   = "\D{Fin}"
-% %format lookup = "\F{lookup}"
-
-% Another useful data type is the finite type,
-% |Fin|\footnote{Note that Agda supports the overloading of
-% data type constructor names. Constructor |zero| can refer to
-% type |Nat| or |Fin|, depending on the context where the name
-% is used.}, which is defined in Agda's standard library as:
-% \begin{spec}
-%   data Fin : Nat -> Set where
-%     zero : forall {n} -> Fin (succ n)
-%     succ : forall {n} -> Fin n -> Fin (succ n)
-% \end{spec}
-% Type |Fin n| has exactly |n| inhabitants
-% (elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
-% An application of such type is to define a safe vector
-% lookup function, which avoids the access of invalid positions.
-% \begin{spec}
-%   lookup : forall {A n} -> Fin n -> Vec A n -> A
-%   lookup zero (x :: _) = x
-%   lookup (succ idx) (_ :: xs) = lookup idx xs
-% \end{spec}
-% Thanks to the propositions-as-types principle,\footnote{It is also known as
-%   Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
-% types as logical formulas and terms as proofs. An example is the
-% representation of equality as the following Agda type:
-
-% \begin{spec}
-%   data _==_ {l}{A : Set l}(x : A) : A -> Set where
-%     refl : x == x
-% \end{spec}
-
-% %format not = "\F{\neg}"
-% %format Dec = "\D{Dec}"
-% %format yes = "\Con{yes}"
-% %format no  = "\Con{no}"
-% %format Even = "\Con{Even}"
-% %format Odd = "\Con{Odd}"
-% %format Parity = "\D{Parity}"
-% %format parity = "\F{parity}"
-% %format natToBin = "\F{natToBin}"
-% %format false = "\Con{false}"
-% %format true = "\Con{true}"
-% %format + = "\F{+}"
-% %format ++ = "\F{++}"
-% %format Bot = "\D{\bot}"
-% %format All = "\D{All}"
-% This type is called propositional equality. It defines that there is
-% a unique evidence for equality, constructor |refl| (for reflexivity),
-% which asserts that the only value equal to |x| is itself. Given a predicate |P : A -> Set|
-% and a vector |xs|, the type |All P xs| is used to build proofs that |P| holds for all
-% elements in |xs| and it is defined as:
-% \begin{spec}
-%   data All (P : A -> Set) : Vec A n ->  Set where
-%      [] : All P []
-%      _::_ : forall {x xs} -> P x -> All P xs -> All P (x :: xs)
-% \end{spec}
-% The first constructor specifies that |All P| holds for the empty vector and
-% constructor |_::_| builds a proof of |All P (x :: xs)| from proofs of
-% |P x| and |All P xs|. Since this type has the same structure of vectors,
-% some functions on |Vec| have similar definitions for type |All|. As an example
-% used in our formalization, consider the function |lookup|, which extracts a
-% proof of |P| for the element at position |v : Fin n| in a |Vec|:
-% \begin{spec}
-%    lookup : {xs : Vec A n} -> Fin n -> All P xs -> P x
-%    lookup zero (px :: _) = px
-%    lookup (succ idx) (_ :: pxs) = lookup idx pxs
-% \end{spec}
+\section{An overview of the Agda programming language}\label{sec:agda}
+
+Agda is a dependently-typed functional programming language based on
+Martin-L\"of intuitionistic type theory~\cite{Lof98}.  Function types
+and an infinite hierarchy of types of types, \ensuremath{\D{Set}\;\V{l}}, where \ensuremath{\V{l}} is a
+natural number, are built-in. Everything else is a user-defined
+type. The type \ensuremath{\D{Set}}, also known as \ensuremath{\D{Set}_{\D{0}}}, is the type of all
+``small'' types, such as \ensuremath{\D{Bool}}, \ensuremath{\D{String}} and \ensuremath{\D{List}\;\D{Bool}}.  The type
+\ensuremath{\D{Set}_{\D{1}}} is the type of \ensuremath{\D{Set}} and ``others like it'', such as \ensuremath{\D{Set}\;\to \;\D{Bool}},
+\ensuremath{\D{String}\;\to \;\D{Set}}, and \ensuremath{\D{Set}\;\to \;\D{Set}}. We have that \ensuremath{\D{Set}\;\V{l}} is an
+element of the type \ensuremath{\D{Set}\;(\V{l+1})}, for every $l \geq 0$. This
+stratification of types is used to keep Agda consistent as a logical
+theory~\cite{Sorensen2006}.
+
+An ordinary (non-dependent) function type is written \ensuremath{\V{A}\;\to \;\V{B}} and a
+dependent one is written \ensuremath{(\V{x}\;\mathbin{:}\;\V{A})\;\to \;\V{B}}, where type \ensuremath{\V{B}} depends on
+\ensuremath{\V{x}}, or \ensuremath{\D{\forall}\;(\V{x}\;\mathbin{:}\;\V{A})\;\to \;\V{B}}. Agda allows the definition of \emph{implicit
+parameters}, i.e.,  parameters whose values can be inferred from the
+context, by surrounding them in curly braces: \ensuremath{\D{\forall}\;\{\mskip1.5mu \V{x}\;\mathbin{:}\;\V{A}\mskip1.5mu\}\;\to \;\V{B}}. To
+avoid clutter, we'll omit implicit arguments from the source code
+presentation. The reader can safely assume that every free variable in
+a type is an implicit parameter.
+
+As an example of Agda code, consider the following data type of
+length-indexed lists, also known as vectors.
+
+ \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[4]{}\mathkw{data}\;\D{\mathbb{N}}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[4]{}\hsindent{2}{}\<[6]%
+\>[6]{}\Con{zero}\;\mathbin{:}\;\D{\mathbb{N}}{}\<[E]%
+\\
+\>[4]{}\hsindent{2}{}\<[6]%
+\>[6]{}\Con{suc}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{\mathbb{N}}{}\<[E]%
+\\[\blanklineskip]%
+\>[4]{}\mathkw{data}\;\D{Vec}\;(\V{A}\;\mathbin{:}\;\D{Set})\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[4]{}\hsindent{2}{}\<[6]%
+\>[6]{}\Con{\lbrack\:\rbrack}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\D{Vec}\;\V{A}\;\Con{zero}{}\<[E]%
+\\
+\>[4]{}\hsindent{2}{}\<[6]%
+\>[6]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\V{A}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})
+inserts a new element in front of a vector of $n$ elements (of type
+\ensuremath{\D{Vec}\;\V{A}\;\V{n}}) and returns a value of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}. The
+\ensuremath{\D{Vec}} datatype is an example of a dependent type, i.e., a type
+that uses a value (that denotes its length). The usefulness of
+dependent types can be illustrated with the definition of a safe list
+head function: \ensuremath{\F{head}} can be defined to accept only non-empty
+vectors, i.e.,~values of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\F{head}\;\mathbin{:}\;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})\;\to \;\V{A}{}\<[E]%
+\\
+\>[3]{}\F{head}\;(\V{x}\;\Con{::}\;\V{xs})\;\mathrel{=}\;\V{x}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+In \ensuremath{\F{head}}'s definition, constructor \ensuremath{\Con{\lbrack\:\rbrack}} is not used. The
+Agda type-checker can figure out, from \ensuremath{\F{head}}'s parameter type,
+that argument \ensuremath{\Con{\lbrack\:\rbrack}} to \ensuremath{\F{head}} is not type-correct.
+
+Another useful data type is the finite type,
+\ensuremath{\D{Fin}}\footnote{Note that Agda supports the overloading of
+data type constructor names. Constructor \ensuremath{\Con{zero}} can refer to
+type \ensuremath{\D{\mathbb{N}}} or \ensuremath{\D{Fin}}, depending on the context where the name
+is used.}, which is defined in Agda's standard library as:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathkw{data}\;\D{Fin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{zero}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{suc}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Fin}\;(\Con{suc}\;\V{n}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Type \ensuremath{\D{Fin}\;\V{n}} has exactly \ensuremath{\V{n}} inhabitants
+(elements), i.e., it is isomorphic to the set $\{0,...,n - 1\}$.
+An application of such type is to define a safe vector
+lookup function, which avoids the access of invalid positions.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\F{lookup}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{Vec}\;\V{A}\;\V{n}\;\to \;\V{A}{}\<[E]%
+\\
+\>[3]{}\F{lookup}\;\Con{zero}\;(\V{x}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{x}{}\<[E]%
+\\
+\>[3]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{xs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{xs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Thanks to the propositions-as-types principle,\footnote{It is also known as
+Curry-Howard ``isomorphism''~\cite{Sorensen2006}.} we can interpret
+types as logical formulas and terms as proofs. An example is the
+representation of equality as the following Agda type:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathkw{data}\;\D{\_ \equiv \_}\;\{\mskip1.5mu \V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{A}\;\mathbin{:}\;\D{Set}\;\V{l}\mskip1.5mu\}\;(\V{x}\;\mathbin{:}\;\V{A})\;\mathbin{:}\;\V{A}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{refl}\;\mathbin{:}\;\V{x}\;\D{\equiv}\;\V{x}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+This type is called propositional equality. It defines that there is
+a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
+which asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
+and a vector \ensuremath{\V{xs}}, the type \ensuremath{\D{All}\;\V{P}\;\V{xs}} is used to build proofs that \ensuremath{\V{P}} holds for all
+elements in \ensuremath{\V{xs}} and it is defined as:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathkw{data}\;\D{All}\;(\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
+\>[41]{}\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\V{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\V{P}\;\V{x}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The first constructor specifies that \ensuremath{\D{All}\;\V{P}} holds for the empty vector and
+constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
+\ensuremath{\V{P}\;\V{x}} and \ensuremath{\D{All}\;\V{P}\;\V{xs}}. Since this type has the same structure of vectors,
+some functions on \ensuremath{\D{Vec}} have similar definitions for type \ensuremath{\D{All}}. As an example
+used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
+proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\V{P}\;\V{x}{}\<[E]%
+\\
+\>[4]{}\F{lookup}\;\Con{zero}\;(\V{px}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{px}{}\<[E]%
+\\
+\>[4]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{pxs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{pxs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\section{Dependently-typed Syntax}\label{sec:syntax}
 
@@
 to denote terms that should not be considered well-typed like
-\ensuremath{(\Con{Num}\;\V{1})\;\V{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
+\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
 as another definition, which should consider all possible cases to avoid the
@@
 %A definition which uses the expression |(Num 1) + True| will be rejected by Agda's type checker automatically.
-Agda's type-checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\V{+}\;\Con{True}}.
+Agda's type-checker will automatically reject a definition which uses the expression \ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}.
 
@@
 \\
-\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;(\text{\ttfamily \char34 variable~out~of~scope:~\char34}\;\plus \;\V{v}){}\<[E]%
+\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;(\text{\ttfamily \char34 variable~out~of~scope:~\char34}\;\F{++}\;\V{v}){}\<[E]%
 \\
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\section*{References}
+
 \bibliography{main}
+
 \end{document}
-\endinput

### 2021-03-08 papers/COLA/paper.tex
Finished typechecker

@@
 then proceeds with its realization as a definitional interpreter.
-The type-checking algorithm is presented in Section~\ref{sec:typechecker}. Section~\ref{sec:comparison}
+The type-checking algorithm is presented in Section~\ref{sec:typechecker}.
+Section~\ref{sec:indirect} describes the necessary modifications in our formalization to
+support indirect jumps, as proposed in the version 2.0 of the list machine benchmark.
+Section~\ref{sec:comparison}
 compares the presented formalization in Agda with Coq and Twelf encodings. Related work is
@@
 
-The list-machine, proposed by \citet{Appel07}, is a simple pointer virtual machine where values
+The list-machine, proposed by Appel et. al.~\cite{AppelDL12}, is a simple pointer virtual machine where values
 are empty lists and cons-cells:
@@
 \]
-The type system proposed by \citet{Appel07} has three rules to deal with each of the possible
+The type system proposed by Appel et. al.~\cite{AppelDL12} has three rules to deal with each of the possible
 types assigned to the branch variable in the current typing context. The first two rules deal with the $list$ and $listcon$
@@
 We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
-\citet{Appel07} also define that the least common supertype relation is sound and complete with respect to the
+Appel et. al.~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
 subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
@@
 extend the run-time environment \ensuremath{\V{env}} with the type of the created list. To convince the Agda's
-type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}},
+type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{<:\textrm{-}val}} and \ensuremath{\F{list\textrm{-}<:\textrm{-}inv}},
 and others lemmas to deal with the least common supertype, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;((\F{\textrm{<:-}val}\;(\F{list\textrm{-<:-}inv}\;(\F{lub\textrm{-}subtype\textrm{-}left}{}\<[E]%
+\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;((\F{<:\textrm{-}val}\;(\F{list\textrm{-}<:\textrm{-}inv}\;(\F{lub\textrm{-}subtype\textrm{-}left}{}\<[E]%
 \\
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
+\>[5]{}\Con{::}\;\F{<:\textrm{-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
 \ColumnHook
@@
 
-\pagebreak
-
 \begin{hscode}\SaveRestoreHook
@@
 
+\section{Extending the Machine with Indirect Jumps}\label{sec:indirect}
+
+In this section we describe the necessary modifications to our formalization to include indirect jumps
+as proposed by Appel et. al.~\cite{AppelDL12}. We start by describing the changes on the type system
+and its impact on the intrinsically typed representation of programs in Section~\ref{sec:changestypes}.
+Updates on the type checker and interpreter implementations are described in Section~\ref{sec:changesintep}.
+
+\subsection{Modifications in the type system and instruction set}
+
+Since our solution relies on representing the program using intrinsically typed syntax, we describe
+first the changes on the machine type system. The first modification is the inclusion of a type for
+program labels, which denotes continuations in a program.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Ty}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{nil}\;\Con{\top}\;\Con{\bot}\;\mathbin{:}\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{list}\;\Con{listcons}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{cont}\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;\D{Ty}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+The type \ensuremath{\Con{cont}} $\Gamma$ for a label express that the machine can safely jump when its registers
+satisfies $\Gamma$. The inclusion of continuation types makes the type and context syntax mutually
+inductive.
+
+Since we have changed the type syntax, we need to adapt the definition of the subtyping relation.
+The continuation type substantially changes the subtyping relation,
+since now we have ``incomparable'' types: list types are not subtypes neither supertypes for continuations.
+The solution is to complete the subtyping relation with a bottom and top types. The subtyping rules for these
+new types are presented next.
+
+\[
+\begin{array}{cc}
+  \infer{\bot <: \tau}{} &
+  \infer{\tau <: \top}{} \\ & \\
+  \multicolumn{2}{c}{
+    \infer{cont\,\Gamma_1\:<:\:cont\,\Gamma_2}{\Gamma_2 \subset \Gamma_1}
+  }
+\end{array}
+\]
+
+We omit its representation in Agda because its just a direct encoding of these previous rules
+in the subtyping relation inductive type. Notice that both subtyping relations for types and
+contexts are mutually inductive.
+
+Another major modification in the typing for the list machine is the definition of least upper bounds.
+First, we complete the least upper bound relation by specifying that the lub for incomparable types is
+$\top$.
+
+\[
+\begin{array}{cc}
+  \infer{cont\:\Gamma\:\sqcup\:nil = \top}{} & \infer{cont\:\Gamma\:\sqcup\:(list\:\tau) = \top}\\ & \\
+  \multicolumn{2}{c}{\infer{cont\:\Gamma\:\sqcup\:(listcons\:\tau) = \top}}
+\end{array}
+\]
+
+We omit rules that ensure the commutativity of the least upper bound relation for brevity.
+Next, we present the rules for the least upper bound relation for top and bottom types.
+
+\[
+\begin{array}{cc}
+  \infer{\tau \sqcup \bot = \tau}{} & \infer{\tau \sqcup \top = \top}{}
+\end{array}
+\]
+
+The last set of rules for the least upper bound relation deals with continuations.
+Because of the contravariance of the \ensuremath{\Con{cont}} type, we also need to define greatest lower bounds for both
+types and typing contexts. First, we present the rule for the least upper bound for continuation types.
+
+\[
+\begin{array}{c}
+  \infer{cont\:\Gamma_1\:\sqcap\: cont\:\Gamma_2=cont\:\Gamma_3}{\Gamma_1\cup\Gamma_2 = \Gamma_3}
+\end{array}
+\]
+
+The rule specifies that the least upper bound for types \ensuremath{\Con{cont}} $\Gamma_1$ and \ensuremath{\Con{cont}} $\Gamma_2$ is
+\ensuremath{\Con{cont}} $\Gamma_3$, where $\Gamma_3$ is greatest lower bound for typing contexts $\Gamma_1$ and $\Gamma_2$.
+The definition of the greatest lower bound for contexts is presented next.
+\[
+  \begin{array}{cc}
+    \infer{\{\} \cup \Gamma = \Gamma}{} &
+    \infer{\Gamma \cup \{\} = \Gamma}{} \\ & \\
+    \multicolumn{2}{c}{
+      \infer{(\Gamma_1 , x : \tau_1) \cup (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
+            {\tau_1 \sqcup \tau_2 = \tau_3 & \Gamma_1 \cup \Gamma_2 = \Gamma_3}
+    }
+  \end{array}
+\]
+The definition of the greatest lower bound for types is as follows. The first set of rules specify the reflexivity,
+that $`\top$ type is an identity and $`\bot$ is for greastest lower bound for any type $\tau$.
+Again, rules to ensure commutativity of the relation are ommited for brevity.
+
+\[
+\begin{array}{ccc}
+  \infer{\tau \sqcup \tau = \tau}{} & \infer{\top \sqcup \tau = \tau}{} & \infer{\bot \sqcup \tau = \bot}{}\\
+\end{array}
+\]
+Next, we define the greatest lower bound relation for list types. The first rule specifies that the bottom
+type is the greatest lower bound for the empty list type and non-empty list type. Two rules specifies the
+compatibility of \ensuremath{\Con{list}} and \ensuremath{\Con{listcons}} type constructors with the greatest lower bound relation. The last
+rule show that \ensuremath{\Con{listcons}} $\tau_3$ is the lower bound for \ensuremath{\Con{list}} $\tau_1$ and \ensuremath{\Con{listcons}} $\tau_2$.
+\[
+\begin{array}{cc}
+  \infer{nil \sqcup (listcons\: \tau) = \bot}{} &
+  \infer{(list\: \tau_1)\sqcup (list\:\tau_2) = list\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
+   \infer{(listcons\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} &
+  \infer{(list\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
+\end{array}
+\]
+Rules for the greatest lower bound relation for continuation types shows that bottom type is the
+lower bound for list types and continuation one. The other rule for stabishes the compatibility of
+the continuation type with the lower bound relation and it uses the least upper bound for typing
+contexts relation.
+\[
+\begin{array}{ccc}
+  \infer{nil\sqcup(cont\:\Gamma) = \bot}{} &
+  \infer{(list\:\tau)\sqcup(cont\:\Gamma) = \bot}{} &
+  \infer{(listcons\:\tau)\sqcup(cont\:\Gamma) = \bot}{} \\ & & \\
+  \multicolumn{3}{c}{
+    \infer{(cont\:\Gamma_1)\sqcup(cont\:\Gamma_2) = (cont\:\Gamma_3)}{\Gamma_1\cap\Gamma_2 = \Gamma_3}
+  }
+\end{array}
+\]
+We omit several rules for the greatest lower bound relation that ensure the commutativity of the
+relation and that $\bot$ type is a lower bound for any type. Rules for the least upper bound for
+typing contexts are simply the dual version for greatest lower bound for contexts, as shown next.
+\[
+  \begin{array}{cc}
+    \infer{\{\} \cap \Gamma = \{\}}{} &
+    \infer{\Gamma \cap \{\} = \{\}}{} \\ & \\
+    \multicolumn{2}{c}{
+      \infer{(\Gamma_1 , x : \tau_1) \cap (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
+            {\tau_1 \sqcap \tau_2 = \tau_3 & \Gamma_1 \cap \Gamma_2 = \Gamma_3}
+    }
+  \end{array}
+\]
+The Agda encoding of these relations is an immediate translation of the mathematical notation
+to code and are omitted for brevity.
+
+\subsection{Modifications in the type-checker and interpreter}\label{sec:changesintep}
+
 \section{Comparison of Mechanized Proofs}\label{sec:comparison}
@@
 We implemented all 14 tasks from the list-machine benchmark in the Agda programming language.
-The tasks considered by us are the same implemented and proved by~\citet{Appel07}.
+The tasks considered by us are the same implemented and proved by Appel et. al.~\cite{Appel07}.
 The next table summarizes the total number of lines of code (LOC) for our results together with theirs.
@@
         from this text for space reasons.
-  \item \textbf{State the soundness theorem for the type system.} Following~\citet{Amin17}, we represent the soundness
+  \item \textbf{State the soundness theorem for the type system.} Following~\cite{Amin17}, we represent the soundness
         theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list-machine
         programs. Therefore, this was done as part of task 9.
-  \item \textbf{Prove the soundness theorem for the type system.} Following \citet{Amin17}, the proof of the soundness
+  \item \textbf{Prove the soundness theorem for the type system.} Following \cite{Amin17}, the proof of the soundness
         theorem is just the implementation of the definitional interpreter. Soundness is ensured by construction since the interpreter
@@
 
-As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to \citet{Appel07} solution.
+As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel et. al.
+\cite{Appel07} solution.
 Our implementation used 415 LOC to complete the tasks, while the Twelf solution demanded 2898 LOC and 887 LOC in Coq.
@@
 
-\section*{References}
-
 \bibliography{main}

### 2021-03-09 papers/COLA/paper.tex
Little modifications on text

@@
 
+The modifications on the subtyping and least upper bound induced changes on all theorems
+about these relations. On our first implementation, these results are just straightforward
+inductive proofs. Because of the mutually inductive nature of the new relation versions, all
+these proofs needed to be defined by mutually recursive functions.
+
 \subsection{Modifications in the type-checker and interpreter}\label{sec:changesintep}

### 2021-03-09 papers/COLA/paper.tex
Fixes on the code

@@
 and its impact on the intrinsically typed representation of programs in Section~\ref{sec:changestypes}.
-Updates on the type checker and interpreter implementations are described in Section~\ref{sec:changesintep}.
-
-\subsection{Modifications in the type system and instruction set}
+Updates on the type checker and interpreter implementations are described in Sections~\ref{sec:changestypechecker}
+and ~\ref{sec:changesintep}, respectively.
+
+\subsection{Modifications in the type system and instruction set}\label{sec:changestypes}
 
@@
 
-\subsection{Modifications in the type-checker and interpreter}\label{sec:changesintep}
+Once we have modified the subtyping and least upper bound relations, most of the type system
+rules remain unchanged from the original version. In order to support indirect jumps, Appel et. al.
+approach is to modify the jump instruction and add a new instruction for loading label values into
+machine registers. The typing rules for get-label and jump instructions are as follows.
+
+\[
+\begin{array}{cc}
+  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:L_0\}\Gamma'}
+        {\Gamma[v\,:=\,nil] = \Gamma'} &
+  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:l\}\Gamma'}
+        {\Pi(l) = \Gamma_1 & \Gamma[v\,:=\,cont\:\Gamma_1] = \Gamma'} \\ & \\
+  \multicolumn{2}{c}{
+    \infer{\Pi;\Gamma\vdash_{block}\,\textrm{jump}\:v}
+          {\Gamma(v) = cont\:\Gamma_1 & \Gamma \subset \Gamma_1}
+  }
+\end{array}
+\]
+The first rule show that the type of label $L_0$, the starting label of a complete machine program,
+is $nil$ and the second specifies that the type for any other label is a continuation type formed by
+the context associated with label $l$ in $\Pi$. The last rule show that typing a jump instruction
+requires that its register has a continuation type, $cont\:\Gamma_1$, that should be a subtype of
+current block context $\Gamma$. Based on these rules, we add the following constructors to our
+instruction typed syntax:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{40}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{\_\vdash\_\Rightarrow\_}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx})\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\orange{\texttt{-- unchanged from the previous code version.}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}getlabel\textrm{-}0}\;{}\<[25]%
+\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[25]{}\hsindent{10}{}\<[35]%
+\>[35]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{nil})){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{instr\textrm{-}getlabel}\;{}\<[25]%
+\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\tau}\;\V{\Gamma_1}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[25]{}\hsindent{15}{}\<[40]%
+\>[40]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;{}\<[E]%
+\\
+\>[25]{}\hsindent{15}{}\<[40]%
+\>[40]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{cont}\;\V{\Gamma_1})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} encodes the typing rule for the instruction \textrm{get-label} when
+the label involved is the special label $L_0$. Intutively, the constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} type
+ensures that the type associated with the register for the instruction is the empty list type.
+Next, constructor \ensuremath{\Con{instr\textrm{-}getlabel}} show that any other label $l$ should be typed with a continuation
+holding the typing context associated with $l$ in $\Pi$.
+
+The needed modifications on block syntax are presented next.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{42}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Block}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx})\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;{}\<[42]%
+\>[42]{}\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\orange{\texttt{-- unchanged from the previous code version}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{block\textrm{-}jump}\;{}\<[25]%
+\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{x}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{cont}\;\V{\Gamma_1})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[25]{}\rightarrow\;\V{\Gamma}\;\D{\subset}\;\V{\Gamma_1}\;{}\<[E]%
+\\
+\>[25]{}\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma_1}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We only need to adapt the type of constructor for jump by requiring that the register has a
+continuation type whose typing context is a subtype of the current block type, which allows
+the safe jump to the next block.
+
+\subsection{Modifications in the type-checker}\label{sec:changestypechecker}
+
+\subsection{Modification in the intepreter}\label{sec:changesintep}
 

### 2021-03-10 papers/COLA/paper.tex
More updates

@@
 
-\subsection{Modification in the intepreter}\label{sec:changesintep}
+Most of the code for the type-checker is ummodified in our new version. We just need to change
+the type checking for jumps, which now should validate if a machine register has an appropriate
+continuation type.
+
+We also need to include code to check the well-typedness for the get-label instruction. First,
+we include a new constructor on type \ensuremath{\D{UInstr}} to represent untyped get-label instruction and
+add its corresponding equation on function \ensuremath{\F{forge\textrm{-}types\textrm{-}instr}}, which erases type information
+from the instrinsically typed syntax. This function is used by the type-checker to ensure its
+correctness.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{UInstr}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\orange{\texttt{-- same code from before.}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{get\textrm{-}label}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;(\V{v}\;\mathbin{:}\;\V{ℕ})\;\rightarrow\;\D{Label}\;\rightarrow\;\D{UInstr}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}{}\<[E]%
+\\
+\>[B]{}\orange{\texttt{-- same code as before...}}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\Con{L_{0}}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v}\;\V{p})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\V{l}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Constructor \ensuremath{\Con{get\textrm{-}label}} stores the name for a register and a label which is loaded into the register.
+Equations for erasing type information are immediate. The only peculiarity is that when the instruction
+\ensuremath{\Con{get\textrm{-}label}} is typed using rule \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} we consider the initial label \ensuremath{\Con{L_{0}}}. Next, we implement
+the type checker logic for \ensuremath{\Con{get\textrm{-}label}}. Function \ensuremath{\F{check\textrm{-}get\textrm{-}label}} verifies if the label is present at
+program global typing context and if the variable is present on current block typing context.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\mathbin{:}\;\D{\forall}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;{}\<[30]%
+\>[30]{}\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{get\textrm{-}label}\;\V{v}\;\V{l})){}\<[E]%
+\\
+\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mathkw{with}\;\F{inspect}\;(\F{lookup}\;\V{\Pi}\;\V{l}){}\<[E]%
+\\
+\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mid \;\V{\Gamma_1}\;\F{with\equiv}\;\V{p}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 check-get-label:~variable~out~of~scope\char34}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{just}\;(\V{\tau}\;\red{,}\,\;\V{idx})\;\mathrel{=}\;\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}getlabel}\;\V{idx}\;(\F{lookup⇒[]=}\;\V{l}\;\V{\Pi}\;\V{p}))){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\subsection{Modifications in the intepreter}\label{sec:changesintep}
+
+The final piece of our formalization for the version 2.0 of the list machine benchmark is modifying
+the interpreter implementation. Our first step is to change the definition of values to reflect
+the changes on the type syntax, since now we have a new value for continuation types. A continuation
+value simply stores the label of the corresponding block.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathkw{data}\;\D{Val}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\orange{\texttt{-- same code as before}}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Con{cont}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;\D{Val}\;(\Con{cont}\;\V{\Gamma_1}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Since the value definition has changed, we need to modify the lemmas for value and environment
+subsumption accordingly. Next, we change the interpreter itself by adding equations for the
+\ensuremath{\Con{get\textrm{-}label}} instruction and fixing the \ensuremath{\Con{jump}} implementation.
+Equations for \ensuremath{\Con{get\textrm{-}label}} simply update the current environment using the label. When the
+typed syntax is built using constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}}, the environment is updated with
+value \ensuremath{\Con{nil}}, which denotes the initial continuation for the program execution.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{penv}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\V{idx})\;\V{b})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{penv}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{idx}\;\Con{nil})\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{penv}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}getlabel}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x})\;\V{b})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{penv}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{idx}\;(\Con{cont}\;\V{x}))\;\V{b}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The final piece of changes in the interpreter is implementing jump instruction. In order to
+execute an indirect jump, first we need to get the continuation value associated with its
+variable using \ensuremath{\F{lookup}\;\V{env}\;\V{v}}. Using the label associated with the retrieved continuation value
+we jump to its corresponding block and its entry execution environment.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{penv}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}jump}\;\V{v}\;\V{x₁})\;\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\V{...|}\;(\Con{cont}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{k})\;\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{k})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{penv}\;\V{p}\;(\F{lookupA}\;\V{l}\;\V{penv})\;(\F{lookupA}\;\V{l}\;\V{p}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 
+Appel's list machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
+correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics. We stick to our initial
+proposal based on dependently typed syntax and definitional interpreters.
+
 %As we could notice on the previous table, the approach taken in this paper avoids code repetition and decreases the number of LOCs necessary to

### 2021-03-10 papers/COLA/paper.tex
Finished the paper first version

@@
 
-In this work, we extend our previous SBLP 2020 paper by including support to
+In this work, we extend our previous SBLP 2020 paper~\cite{Feitosa2020} by including support to
 indirect jumps, as specified in the version 2.0 of the list
@@
 changes on the machine syntax, type system and interpreter to accomodate this extension, which
-allows the handling of more complex control flow, like the call/return model of function calls.
-
-More specifically, our contributions are:
+allows the handling of more complex control flow structures, like the call/return model of function calls.
+
+More specifically, we contribute:
 
@@
         unnecessary repetitions and provides some properties for free.
-  \item We provide a detailed discussion on the necessary type system and interpreter modifications
+  \item We provide a detailed discussion on the necessary modifications to the type system,
+        type checker and interpreter
         to support indirect jumps.
@@
 Appel's list machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
-correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics. We stick to our initial
-proposal based on dependently typed syntax and definitional interpreters.
+correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics.
+We stick to our initial proposal based on dependently typed syntax and definitional interpreters. Appel's formalization for
+the version 2.0 of the benchmark demanded around 1750 LOC without considering his library for the
+``very modal model''~\cite{Appel07}. Our formalization demanded 1134 LOC and it depends only on the Agda standard library.
+Compared to our solution to the first version of the benchmark (which was implemented in 612 LOC), the inclusion of
+indirect jumps increased the formalization size by around 85\%.
 
@@
 was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
-machine benchmark was proposed by~\citet{Appel07} as an exercise in formalizing results that interest
+machine benchmark was proposed by~\cite{Appel07} as an exercise in formalizing results that interest
 compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
 solution to this benchmark using an intrinsically-typed approach in the Agda programming language. Representation of binding syntax
-was the subject of~\citet{FeltyMP18} which proposed a set of problems and research questions for tools
+was the subject of~\cite{FeltyMP18} which proposed a set of problems and research questions for tools
 that use the high-order abstract syntax approach for name binding. Since the list-machine benchmark definition avoids
@@
 the correct manipulation of names by following the traditional \emph{de Bruijn} approach. Finally, a recent problem set
-was proposed by~\citet{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
+was proposed by~\cite{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
 of logical relation arguments, like strong normalization theorems.}
@@
 The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
-Recently, the interest on such interpreters was revitalized by~\citet{Amin17}, which used definitional
+Recently, the interest on such interpreters was revitalized by~\cite{Amin17}, which used definitional
 interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
 Unlike our work, Amin and Rompf's formalizations do not use intrinsically-typed representations of their syntax, what cluttered
-their formalization with the need of dealing with ``stuck states'' in the semantics. \citet{Poulsen18} described how
+their formalization with the need of dealing with ``stuck states'' in the semantics. \cite{Poulsen18} described how
 definitional interpreters for imperative languages can be much concisely implemented by using intrinsically-typed syntax and a
@@
 using \emph{de Bruijn} indices was sufficient to implement the desired type-checker and interpreter.
-Other recent application of definitional interpreters was proposed by~\citet{Rouvoet20}. The main contribution
+Other recent application of definitional interpreters was proposed by~\cite{Rouvoet20}. The main contribution
 of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
@@
 named \emph{Market}, supporting the main operations for accessing and updating the store used by the interpreter.
-Another work using intrinsically-typed syntax for resource control was developed by~\citet{Thiemann19}, which
+Another work using intrinsically-typed syntax for resource control was developed by~\cite{Thiemann19}, which
 implemented an interpreter for a more realistic core functional session typed calculus including recursion and
@@
 of linear and session types as in~\cite{Rouvoet20,Thiemann19}. A formalization of System F$_{\omega\mu}$ was
-the subject of \citet{ChapmanKNW19} work, which used an intrinsically-typed representation to
+the subject of \cite{ChapmanKNW19} work, which used an intrinsically-typed representation to
 implement a normalization by evaluation for this calculus. The reason behind such formalization effort was
@@
 This paper shows how the combination of intrinsically-typed syntax and definitional interpreters can be used to simplify
-the tasks on the formalization of programming languages. Using such approach, we were able to provide a machine-checked version of the
-list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
+the tasks on the formalization of programming languages. Using such approach, we were able to provide a machine-checked
+version of the two versions of the list-machine benchmark in Agda, showing that the approach is useful to
+formalize both high-level and low-level languages.
 The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM,
 since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing our work with the
-conventional non-dependently typed formalization strategies (like the one used by \citet{Appel07} in their Coq and Twelf implementations),
+conventional non-dependently typed formalization strategies (like the one used in~\cite{Appel07} in their Coq and Twelf implementations),
 we can affirm that it requires a fewer number of lines to achieve the same result, even without the use of proof automation.

### 2021-03-11 papers/COLA/paper.tex
Fixes based on Samuel suggestions

@@
 using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
-compared to a Coq solution, even without the use of proof automation.
+compared to a Coq solution, even without the use of proof automation. We also
+describe a solution to the second version of the benchmark and compare it with
+Appel's and Leroy Coq-based solution.
 \end{abstract}
@@
 \begin{keyword}
-Dependent types, formal semantics
+Dependent types, formal semantics, Agda
 \end{keyword}
@@
   \item We provide a detailed discussion on the necessary modifications to the type system,
-        type checker and interpreter
-        to support indirect jumps.
+        type checker and interpreter to support indirect jumps.
 \end{itemize}
@@
 
-Using this data type\footnote{Agda supports the definition of mixfix operators.
+With this data type\footnote{Agda supports the definition of mixfix operators.
 We can use underscores to mark arguments positions.}, we can construct expressions
@@
         & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
-        & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in v')} \\
+        & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in $v'$)} \\
         & \mid & \text{halt}                         & \text{(finishes execution)}\\
@@
 
-Following common practice, the meta-variable $\Gamma$ denotes an
+Following the common practice, the meta-variable $\Gamma$ denotes an
 environment binding variable names to their types. Notation $\{\}$ denotes an empty environment, $v : \tau , \Gamma$
@@
 Using the previously defined notations, the typing rules for the list-machine are defined
-as a syntax directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
+as a syntax-directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
 which intuitively means that the instruction $\iota$ transforms an input typing environment
@@
 \]
-The type system proposed by Appel et. al.~\cite{AppelDL12} has three rules to deal with each of the possible
+The type system proposed by Appel and Leroy~\cite{AppelDL12} has three rules to deal with each of the possible
 types assigned to the branch variable in the current typing context. The first two rules deal with the $list$ and $listcon$
@@
 exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
-space reasons, since it is just an inductive type that encodes the rules presented
+brevity, since it is just an inductive type that encodes the rules presented
 in Section~\ref{sec:list}. Several lemmas about subtyping, including its decidable test,
@@
 We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
-Appel et. al.~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
+Appel and Leroy~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
 subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
@@
 
-For space reasons, we show how we type-check only one instruction\footnote{The curious reader can refer to our
+For brevity, we show how we type-check only one instruction\footnote{The curious reader can refer to our
 online repository for a complete implementation.}. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
@@
 
-\section{Extending the Machine with Indirect Jumps}\label{sec:indirect}
+\section{Extending the List-Machine with Indirect Jumps}\label{sec:indirect}
 
 In this section we describe the necessary modifications to our formalization to include indirect jumps
-as proposed by Appel et. al.~\cite{AppelDL12}. We start by describing the changes on the type system
+as proposed by Appel and Leroy~\cite{AppelDL12}. We start by describing the changes on the type system
 and its impact on the intrinsically typed representation of programs in Section~\ref{sec:changestypes}.
-Updates on the type checker and interpreter implementations are described in Sections~\ref{sec:changestypechecker}
+Updates on the type checker and the interpreter implementations are described in Sections~\ref{sec:changestypechecker}
 and ~\ref{sec:changesintep}, respectively.
@@
 Since our solution relies on representing the program using intrinsically typed syntax, we describe
-first the changes on the machine type system. The first modification is the inclusion of a type for
+first the changes on the list-machine type system. The first modification is the inclusion of a type for
 program labels, which denotes continuations in a program.
@@
 
-We omit its representation in Agda because its just a direct encoding of these previous rules
+We omit their representation in Agda because they are just a direct encoding of these previous rules
 in the subtyping relation inductive type. Notice that both subtyping relations for types and
@@
 Next, we define the greatest lower bound relation for list types. The first rule specifies that the bottom
-type is the greatest lower bound for the empty list type and non-empty list type. Two rules specifies the
+type is the greatest lower bound for the empty list type and non-empty list type. Two rules specify the
 compatibility of \ensuremath{\Con{list}} and \ensuremath{\Con{listcons}} type constructors with the greatest lower bound relation. The last
@@
 \]
-Rules for the greatest lower bound relation for continuation types shows that bottom type is the
-lower bound for list types and continuation one. The other rule for stabishes the compatibility of
+Rules for the greatest lower bound relation for continuation types show that the bottom type is the
+lower bound for the list and continuation types. The other rule stabishes the compatibility of
 the continuation type with the lower bound relation and it uses the least upper bound for typing
@@
 
-Once we have modified the subtyping and least upper bound relations, most of the type system
-rules remain unchanged from the original version. In order to support indirect jumps, Appel et. al.
+Once we have only modified the subtyping and the least upper bound relations, most of the type system
+rules remained unchanged from the original version. In order to support indirect jumps, Appel and Leroy.
 approach is to modify the jump instruction and add a new instruction for loading label values into
@@
 \]
-The first rule show that the type of label $L_0$, the starting label of a complete machine program,
+The first rule shows that the type of label $L_0$, the starting label of a complete machine program,
 is $nil$ and the second specifies that the type for any other label is a continuation type formed by
@@
 requires that its register has a continuation type, $cont\:\Gamma_1$, that should be a subtype of
-current block context $\Gamma$. Based on these rules, we add the following constructors to our
+the current block context $\Gamma$. Based on these rules, we add the following constructors to our
 instruction typed syntax:
@@
 ensures that the type associated with the register for the instruction is the empty list type.
-Next, constructor \ensuremath{\Con{instr\textrm{-}getlabel}} show that any other label $l$ should be typed with a continuation
+Next, constructor \ensuremath{\Con{instr\textrm{-}getlabel}} shows that any other label $l$ should be typed with a continuation
 holding the typing context associated with $l$ in $\Pi$.
@@
 
-Most of the code for the type-checker is ummodified in our new version. We just need to change
+Most of the code for the type-checker is unmodified in our new version. We just need to change
 the type checking for jumps, which now should validate if a machine register has an appropriate
@@
 we include a new constructor on type \ensuremath{\D{UInstr}} to represent untyped get-label instruction and
-add its corresponding equation on function \ensuremath{\F{forge\textrm{-}types\textrm{-}instr}}, which erases type information
+add its corresponding equation on function \ensuremath{\F{forget\textrm{-}types\textrm{-}instr}}, which erases type information
 from the instrinsically typed syntax. This function is used by the type-checker to ensure its
@@
 \\[\blanklineskip]%
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}{}\<[E]%
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}{}\<[E]%
 \\
@@
 \\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\Con{L_{0}}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v}\;\V{p})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\V{l}{}\<[E]%
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\Con{L_{0}}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v}\;\V{p})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\V{l}{}\<[E]%
 \ColumnHook
@@
 the type checker logic for \ensuremath{\Con{get\textrm{-}label}}. Function \ensuremath{\F{check\textrm{-}get\textrm{-}label}} verifies if the label is present at
-program global typing context and if the variable is present on current block typing context.
+the program global typing context and if the variable is present on the current block typing context.
 \begin{hscode}\SaveRestoreHook
@@
 
-The final piece of our formalization for the version 2.0 of the list machine benchmark is modifying
-the interpreter implementation. Our first step is to change the definition of values to reflect
-the changes on the type syntax, since now we have a new value for continuation types. A continuation
+The final piece of our formalization for the version 2.0 of the list-machine benchmark is the modification of
+the interpreter implementation. Our first step changed the definition of values to reflect
+the changes on the typed syntax, since now we have a new value for continuation types. A continuation
 value simply stores the label of the corresponding block.
@@
 subsumption accordingly. Next, we change the interpreter itself by adding equations for the
-\ensuremath{\Con{get\textrm{-}label}} instruction and fixing the \ensuremath{\Con{jump}} implementation.
+\ensuremath{\Con{get\textrm{-}label}} instruction and adjusting the \ensuremath{\Con{jump}} implementation.
 Equations for \ensuremath{\Con{get\textrm{-}label}} simply update the current environment using the label. When the
@@
 \end{hscode}\resethooks
-The final piece of changes in the interpreter is implementing jump instruction. In order to
+The final change in the interpreter is the implementation of the jump instruction. In order to
 execute an indirect jump, first we need to get the continuation value associated with its
@@
 We implemented all 14 tasks from the list-machine benchmark in the Agda programming language.
-The tasks considered by us are the same implemented and proved by Appel et. al.~\cite{Appel07}.
+The tasks considered by us are the same implemented and proved by Appel and Leroy.~\cite{Appel07}.
 The next table summarizes the total number of lines of code (LOC) for our results together with theirs.
@@
         are simple recursive functions over the least common supertype and subtyping relations definitions and were omitted
-        from this text for space reasons.
+        from this text for brevity.
   \item \textbf{State the soundness theorem for the type system.} Following~\cite{Amin17}, we represent the soundness
@@
 
-As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel et. al.
+As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel and Leroy's
 \cite{Appel07} solution.
@@
 
-Appel's list machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
+Appel and Leroy's list-machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
 correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics.
-We stick to our initial proposal based on dependently typed syntax and definitional interpreters. Appel's formalization for
+We stick to our initial proposal based on dependently typed syntax and definitional interpreters. Appel and Leroy's formalization for
 the version 2.0 of the benchmark demanded around 1750 LOC without considering his library for the

### 2021-07-23 papers/COLA/paper.tex
Before revision

@@
 Agda programming language, comparing the results with formalizations developed
-by Appel and Leroy. We formalize the 14 tasks of the benchmark
+by Appel et. al. We formalize the 14 tasks of the benchmark
 using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
@@
 describe a solution to the second version of the benchmark and compare it with
-Appel's and Leroy Coq-based solution.
+Appel'et. al. Coq-based solution.
 \end{abstract}
@@
 representation to build a definitional interpreter for a low-level virtual
-machine, developed by Appel and Leroy, as a benchmark problem closer to
+machine, developed by Appel et. al., as a benchmark problem closer to
 real-world implementations such as typed assembly languages~\cite{CraryM99} and
@@
 
-More specifically, we contribute:
+Let us summarize our contributions. %More specifically, we contribute:
 
@@
 
-Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})
+Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})\footnote{Agda supports the definition of mixfix operators.
+We can use underscores to mark arguments positions.}
 inserts a new element in front of a vector of $n$ elements (of type
@@
 a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
-which asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
-and a vector \ensuremath{\V{xs}}, the type \ensuremath{\D{All}\;\V{P}\;\V{xs}} is used to build proofs that \ensuremath{\V{P}} holds for all
-elements in \ensuremath{\V{xs}} and it is defined as:
+which asserts that the only value equal to \ensuremath{\V{x}} is itself.
+
+Dependently typed pattern matching is built by using the so-called
+\ensuremath{\mathkw{with}} construct, that allows for matching intermediate
+values~\cite{McBride2004}. If the matched value has a dependent type,
+then its result can affect the form of other values. For example,
+consider the following code that defines a type for natural number
+parity. If the natural number is even, it can be represented as the
+sum of two equal natural numbers; if it is odd, it is equal to one
+plus the sum of two equal values. Pattern matching on a value of
+\ensuremath{\D{Parity}\;\V{n}} allows to discover if $n = j + j$ or $n = S (k + k)$,
+for some $j$ and $k$ in each branch of \ensuremath{\mathkw{with}}.  Note that the
+value of $n$ is specialized accordingly, using information ``learned''
+by the type-checker.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Parity}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Con{Even}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{n}\;\F{+}\;\V{n}){}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Con{Odd}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{S}\;(\V{n}\;\F{+}\;\V{n})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\F{parity}\;\mathbin{:}\;(\V{n}\;\mathbin{:}\;\D{\mathbb{N}})\;\to \;\D{Parity}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\F{parity}\;\mathrel{=}\orange{\texttt{-- definition omitted}}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\F{natToBin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{List}\;\D{Bool}{}\<[E]%
+\\
+\>[B]{}\F{natToBin}\;\Con{zero}\;\mathrel{=}\;\Con{\lbrack\:\rbrack}{}\<[E]%
+\\
+\>[B]{}\F{natToBin}\;\V{k}\;\mathkw{with}\;(\F{parity}\;\V{k}){}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\F{natToBin}\;(\V{j}\;\F{+}\;\V{j})\;{}\<[25]%
+\>[25]{}\mid \;\Con{Even}\;\mathrel{=}\;\Con{false}\;\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\F{natToBin}\;(\Con{suc}\;(\V{j}\;\F{+}\;\V{j}))\;\mid \;\Con{Odd}\;{}\<[35]%
+\>[35]{}\mathrel{=}\;\Con{true}\;{}\<[43]%
+\>[43]{}\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+A important application of Agda's dependently-typed pattern matching is
+to rewrite a equality. As an example, consider the following code which
+postulates the commutativity of natural number addition and a predicate
+over natural numbers, \ensuremath{\D{P}}.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{postulate}\;\F{plus\textrm{-}comm}\;\mathbin{:}\;(\V{a}\;\V{b}\;\mathbin{:}\;\D{\mathbb{N}})\;\V{→}\;\V{a}\;\F{+}\;\V{b}\;\D{\equiv}\;\V{b}\;\F{+}\;\V{a}{}\<[E]%
+\\
+\>[B]{}\mathkw{postulate}\;\D{P}\;\mathbin{:}\;\D{\mathbb{N}}\;\V{→}\;\D{Set}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\F{foo}\;\mathbin{:}\;(\V{a}\;\V{b}\;\mathbin{:}\;\D{\mathbb{N}})\;\V{→}\;\D{P}\;(\V{a}\;\F{+}\;\V{b})\;\V{→}\;\D{P}\;(\V{b}\;\F{+}\;\V{a}){}\<[E]%
+\\
+\>[B]{}\F{foo}\;\V{a}\;\V{b}\;\V{t}\;\mathkw{with}\;{}\<[18]%
+\>[18]{}\V{a}\;\F{+}\;\V{b}\;{}\<[25]%
+\>[25]{}\mid \;\F{plus\textrm{-}comm}\;\V{a}\;\V{b}{}\<[E]%
+\\
+\>[B]{}\F{foo}\;\V{a}\;\V{b}\;\V{t}\;{}\<[14]%
+\>[14]{}\mid \;.\;(\V{b}\;\F{+}\;\V{a})\;\mid \;\Con{refl}\;\mathrel{=}\;\V{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Function \ensuremath{\F{foo}} uses the \ensuremath{\mathkw{with}} construct to pattern-match on the equality produced
+by \ensuremath{\F{plus\textrm{-}comm}},  \ensuremath{\V{a}\;\F{+}\;\V{b}\;\D{\equiv}\;\V{b}\;\F{+}\;\V{a}}. Since the propositional equality type has only one
+constructor, \ensuremath{\Con{refl}}, matching over it makes Agda type checker to replace
+all occurrences of \ensuremath{\V{a}\;\F{+}\;\V{b}} to \ensuremath{\V{b}\;\F{+}\;\V{a}} in the current equation context. We can notice
+this by the use of the \empth{dotted-pattern} \ensuremath{.\;(\V{b}\;\F{+}\;\V{a})} which indicates that the
+only way that this equation is correct is when the first result of the \ensuremath{\mathkw{with}} construct
+is \ensuremath{\V{b}\;\F{+}\;\V{a}}. The use of dependent pattern matching to rewrite equalities in functions is
+so common in Agda that the language provides a syntax sugar that simplifies such structure.
+The \ensuremath{\mathkw{rewrite}} construct allows the substitution of an equality in all values in a certain
+equation. The \ensuremath{\F{foo}} function could be represented more concisely, using \ensuremath{\mathkw{rewrite}}, as:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{foo}\;\mathbin{:}\;(\V{a}\;\V{b}\;\mathbin{:}\;\D{\mathbb{N}})\;\V{→}\;\D{P}\;(\V{a}\;\F{+}\;\V{b})\;\V{→}\;\D{P}\;(\V{b}\;\F{+}\;\V{a}){}\<[E]%
+\\
+\>[B]{}\F{foo}\;\V{a}\;\V{b}\;\V{t}\;\mathkw{rewrite}\;\F{plus\textrm{-}comm}\;\V{a}\;\V{b}\;\mathrel{=}\;\V{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Another inductive predicate used in our formalization is the type \ensuremath{\D{All}\;\D{P}\;\V{xs}} which
+ensures that all elements of a vector \ensuremath{\V{xs}} satisfy a predicate \ensuremath{\D{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}.
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathkw{data}\;\D{All}\;(\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
+\>[3]{}\mathkw{data}\;\D{All}\;(\D{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
 \>[41]{}\D{Set}\;\mathkw{where}{}\<[E]%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\V{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
+\>[5]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\D{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\V{P}\;\V{x}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-The first constructor specifies that \ensuremath{\D{All}\;\V{P}} holds for the empty vector and
-constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
-\ensuremath{\V{P}\;\V{x}} and \ensuremath{\D{All}\;\V{P}\;\V{xs}}. Since this type has the same structure of vectors,
+\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\D{P}\;\V{x}\;\to \;\D{All}\;\D{P}\;\V{xs}\;\to \;\D{All}\;\D{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The first constructor specifies that \ensuremath{\D{All}\;\D{P}} holds for the empty vector and
+constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\D{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
+\ensuremath{\D{P}\;\V{x}} and \ensuremath{\D{All}\;\D{P}\;\V{xs}}. Since this type has the same structure of vectors,
 some functions on \ensuremath{\D{Vec}} have similar definitions for type \ensuremath{\D{All}}. As an example
 used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
-proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
+proof of \ensuremath{\D{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\V{P}\;\V{x}{}\<[E]%
+\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\D{P}\;\V{xs}\;\to \;\D{P}\;\V{x}{}\<[E]%
 \\
@@
 constructor, leaving the responsibility of checking type safety to the
-host's language type-checker. As an example, consider the following simple
+host's language type-checker. Before considering our solution to encode list-machine
+programs as instrisincally-typed syntax, we review this approach in a reduced example.
+
+Consider the following simple
 expression language. For this paper, we assume a basic knowledge of functional
@@
 
-With this data type\footnote{Agda supports the definition of mixfix operators.
-We can use underscores to mark arguments positions.}, we can construct expressions
+With this data type, we can construct expressions
 to denote terms that should not be considered well-typed like
@@
 
-%Dependently typed pattern matching is built by using the so-called
-%|with| construct, that allows for matching intermediate
-%values~\cite{McBride2004}. If the matched value has a dependent type,
-%then its result can affect the form of other values. For example,
-%consider the following code that defines a type for natural number
-%parity. If the natural number is even, it can be represented as the
-%sum of two equal natural numbers; if it is odd, it is equal to one
-%plus the sum of two equal values. Pattern matching on a value of
-%|Parity n| allows to discover if $n = j + j$ or $n = S (k + k)$,
-%for some $j$ and $k$ in each branch of |with|.  Note that the
-%value of $n$ is specialized accordingly, using information ``learned''
-%by the type-checker.
-%\begin{spec}
-%data Parity : Nat -> Set where
-%   Even : forall {n : Nat} -> Parity (n + n)
-%   Odd  : forall {n : Nat} -> Parity (S (n + n))
-%
-%parity : (n : Nat) -> Parity n
-%parity = -- definition omitted
-
-%natToBin : Nat -> List Bool
-%natToBin zero = []
-%natToBin k with (parity k)
-%   natToBin (j + j)     | Even = false :: natToBin j
-%   natToBin (succ (j + j)) | Odd  = true  :: natToBin j
-%\end{spec}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The list-machine, proposed by Appel et. al.~\cite{AppelDL12}, is a simple pointer virtual machine where values
+In this section, we review the syntax, semantics and typing rules for the
+list-machine, as originally proposed by Appel et. al.~\cite{AppelDL12}.
+The machine is a simple pointer virtual machine where values
 are empty lists and cons-cells:
@@
 appear primed or subscripted. The syntax of the virtual machine instructions are
-presented next and their meaning is as usual.
+presented next.
+\begin{figure}[H]
 \[
@@
 \]
+\centering
+\caption{Syntax of list-machine programs}
+\label{fig:list-syntax}
+\end{figure}
 A program is just a sequence of blocks referenced by a unique label.
 
-Each program variable is assigned to a list type, which is used to guarantee the safety when executing
+Each program variable is assigned to a list type (Figure~\ref{fig:type-syntax}), which is used to guarantee the safety of executing
 fetch-field instructions that demands non-empty list arguments. In order to express such refinements, types are subject to a
 subtyping relation. The meta-variable $\tau$ denotes an arbitrary type.
+\begin{figure}[H]
 \[
@@
 \]
+\centering
+\caption{Type syntax}
+\label{fig:type-syntax}
+\end{figure}
 The notation $\tau \subset \tau'$
 denotes the subtyping judgment, which is defined as follows.
+\begin{figure}[H]
 \[
-\begin{array}{ccc}
+\begin{array}{c}
   \inference{}
             {\tau\subset \tau}
-            [refl]
-  &
+            [subtype-refl]
+  \\ \\
   \inference{}
-            {nil \subset \tau}
-            [nil]
-  &
+            {nil \subset list\:\tau}
+            [subtype-nil]
+  \\ \\
   \inference{\tau \subset \tau'}
             {list\:\tau\subset list\:\tau'}
-            [list]\\\\
+            [subtype-list]\\ \\
   \multicolumn{3}{c}{
@@
             {listcons\:\tau\subset list\:\tau'}
-            [listcons]} \\\\
+            [subtype-listmixed]} \\ \\
   \multicolumn{3}{c}{
@@
             {listcons\:\tau\subset listcons\:\tau'}
-            [mixed]} \\\\
+            [subtype-listcons]} \\ \\
 \end{array}
 \]
+\centering
+\caption{Subtyping relation}
+\label{fig:subtyping}
+\end{figure}
 Basically, the subtyping relation specifies that $nil$ (empty list type) is
@@
 The other rules specify that type constructors $list$ and $listcons$ respect
-the subtyping relation. The list common supertype $\tau = \tau_1 \sqcap \tau_2$ of
+the subtyping relation. The least common supertype $\tau = \tau_1 \sqcap \tau_2$ of
 $\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
@@
 Subtyping is extended to contexts as follows.
+\begin{figure}[H]
 \[
@@
 \]
+\centering
+\caption{Subtyping relation for typing contexts.}
+\label{fig:subtyping-context}
+\end{figure}
 The variable $\Pi$ is used to denote \emph{program typings}, i.e. finite mappings between
@@
 second.
+\begin{figure}[H]
 \[
 \inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'\:\:\:\:\:\:\:\:\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[seq]
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[check-instr-seq]
 \]
-The type system proposed by Appel and Leroy~\cite{AppelDL12} has three rules to deal with each of the possible
+\centering
+\caption{Typing rule for instruction sequencing}
+\label{fig:typing-seq}
+\end{figure}
+The type system proposed by Appel et. al.~\cite{AppelDL12} has three rules to deal with each of the possible
 types assigned to the branch variable in the current typing context. The first two rules deal with the $list$ and $listcon$
@@
 $\Gamma[v := nil]$. The third rule applies whenever $\Gamma(v) = nil$ and it also demands that $\Gamma \subset \Pi(l)$.
-
+\begin{figure}[H]
 \[
@@
           {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}(v: \text{listcons}~\tau,~\Gamma')}
-          [branch-list]
+          [check-instr-branch-list]
 \\ \\
@@
            \Gamma[v:=\text{nil}]=\Gamma'\:\:\:\:\:\:\:\:\Gamma' \subset_{env} \Gamma_1}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}[branch-listcons]
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}
+          [check-instr-branch-listcons]
 
@@
           {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{branch-if-nil}~v~l \}\Gamma}
-          [branch-nil]
+          [check-instr-branch-nil]
 \end{array}
 \]
+\centering
+\caption{Typing rules for branching instructions}
+\label{fig:typing-branch}
+\end{figure}
 Next, we have the \emph{fetch} instructions, which can be used to store the head / tail of a list value in
@@
 $\tau$.
+\begin{figure}[H]
 \[
@@
 \inference{\Gamma(v) = \text{listcons}~\tau\:\:\:\:\Gamma[v':=\tau]=\Gamma'}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}[fetch-0]
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~0~v' \}\Gamma'}
+          [check-instr-fetch-0]
           \\ \\
 \inference{\Gamma(v) = \text{listcons}~\tau\:\:\:\:\Gamma[v':=\text{list}~\tau]=\Gamma'}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}[fetch-1]
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{fetch-field}~v~1~v' \}\Gamma'}
+          [check-instr-fetch-1]
 \end{array}
 \]
+\centering
+\caption{Typing rules for field fetching instructions}
+\label{fig:typing-fetching}
+\end{figure}
 The \emph{cons} instruction allows us to build a non-empty list value and this rule uses the least
 common supertype operator to check if the result of the operation is really a list type.
+\begin{figure}[H]
 \[
 \inference{\Gamma(v_0) = \tau_0\:\:\:\:\:\:\:\:\Gamma(v_1) = \tau_1 \\
-           (\text{list}~\tau_0) \sqcap \tau_1=\text{list}~\tau\:\:\:\:\:\:\:\:\Gamma[v:=\text{listcons}~\tau]=\Gamma'}
-          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{cons}~v_0~v_1~v \}\Gamma'}[cons]
+           (\text{list}~\tau_0) \sqcup \tau_1=\text{list}~\tau\:\:\:\:\:\:\:\:\Gamma[v:=\text{listcons}~\tau]=\Gamma'}
+          {\Pi \vdash_{\text{instr}} \Gamma\{ \textbf{cons}~v_0~v_1~v \}\Gamma'}
+          [check-instr-cons]
 \]
+\centering
+\caption{Typing rule for cons instruction}
+\label{fig:typing-fetching}
+\end{figure}
 The final rules deal with the well-formedness of blocks and programs. The typing rules for the \textbf{halt} instruction and
@@
 input program.
-
+\begin{figure}[H]
 \[
-\begin{array}{cc}
+\begin{array}{c}
    \inference{}
-             {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[halt]
-   &
+             {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[check-block-halt]
+   \\ \\
    \inference{}
-             {\Pi\vdash_{\text{blocks}} \textbf{end}}[empty]\\ \\
+             {\Pi\vdash_{\text{blocks}} \textbf{end}}[check-block-empty]\\ \\
 
    \inference{\Pi(l)=\Gamma_1\\ \Gamma \subset_{env} \Gamma_1}
-             {\Pi;\Gamma\vdash_{\text{block}} \textbf{jump}~l}[jump]
-
-   &
+             {\Pi;\Gamma\vdash_{\text{block}} \textbf{jump}~l}[check-block-jump]
+
+   \\ \\
 
    \inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'\\ \Pi;\Gamma'\vdash_{\text{block}} \iota_2}
-             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[block-seq]
+             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[check-block-seq]
  \\ \\
 
-   \multicolumn{2}{c}{
    \inference{\Pi(l)=\Gamma\:\:\:\:\:\:\:\:\Pi;\Gamma\vdash_{\text{block}} \iota\:\:\:\:\:\:\:\:\Pi\vdash_{\text{blocks}} p}
-     {\Pi\vdash_{\text{blocks}} l: \iota;~p}[blocks-label]
-    }
+     {\Pi\vdash_{\text{blocks}} l: \iota;~p}[check-blocks-label]
 \end{array}
 \]
+\centering
+\caption{Typing rules for program blocks}
+\label{fig:typing-block}
+\end{figure}
 Inspired by the presented typing rules, in the next section, we define an
-intrinsically-typed syntax for list-machine programs which ensures that
-only well-typed programs can be built.
+intrinsically-typed syntax for list-machine programs which allows
+only well-typed pro\-grams.
 
@@
 \\
-\>[B]{}\D{Ctx}\;\mathrel{=}\;\D{List}\;(\D{String}\;\V{×}\;\D{Ty}){}\<[E]%
+\>[B]{}\D{Ctx}\;\mathrel{=}\;\D{List}\;\D{Ty}{}\<[E]%
 \\[\blanklineskip]%
-\>[B]{}\D{PCtx}\;\mathbin{:}\;\D{Set}{}\<[E]%
-\\
-\>[B]{}\D{PCtx}\;\mathrel{=}\;\D{Vec}\;\D{Ctx}\;\V{n}{}\<[E]%
+\>[B]{}\D{PCtx}\;\mathbin{:}\;\D{\mathbb{N}}\;\rightarrow\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{PCtx}\;\V{n}\;\mathrel{=}\;\D{Vec}\;\D{Ctx}\;\V{n}{}\<[E]%
 \ColumnHook
@@
 The basic idea is to represent each type system rule as a constructor typed by an output context.
-Such representation scheme makes the Agda's type-checker allow only well-typed blocks and
+Such a representation scheme makes Agda's type-checker allow only well-typed blocks and
 instructions to be created and manipulated.
@@
 \>[3]{}\Con{instr\textrm{-}branch\textrm{-}list}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{l}\;\V{\Gamma'}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\Con{list}\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma'}\;\rightarrow\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{nil}))\;\D{\subset}\;\V{\Gamma'}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{l}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;\Con{list}\;\V{\tau}\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma'}\;\rightarrow\;(\V{idx}\;\F{::=}\;\Con{nil})\;\D{\subset}\;\V{\Gamma'}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;\Con{listcons}\;\V{\tau}){}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}branch\textrm{-}listcons}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{l}\;\V{\Gamma_1}\;\V{x}\mskip1.5mu\}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{nil}))\;\D{\subset}\;\V{\Gamma_1}\;{}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{l}\;\V{\Gamma_1}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;(\V{idx}\;\F{::=}\;\Con{nil})\;\D{\subset}\;\V{\Gamma_1}\;{}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}branch\textrm{-}nil}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma_1}\;\V{l}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{nil})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma_1}\;\V{l}\mskip1.5mu\}\;\rightarrow\;\V{x}\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;((\V{x'}\;\red{,}\,\;\V{\tau})\;\Con{::}\;\V{\Gamma}){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\mskip1.5mu\}\;\rightarrow\;\Con{listcons}\;\V{\tau}\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{\tau}\;\Con{::}\;\V{\Gamma}){}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau'}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x'}\;\red{,}\,\;\V{\tau})){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau'}\mskip1.5mu\}\;\rightarrow\;\Con{listcons}\;\V{\tau}\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;\V{\tau'}\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;\V{\tau}){}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;((\V{x'}\;\red{,}\,\;\Con{list}\;\V{\tau})\;\Con{::}\;\V{\Gamma}){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\Con{list}\;\V{\tau}\;\Con{::}\;\V{\Gamma}){}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau'}\;\V{x}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x'}\;\red{,}\,\;\Con{list}\;\V{\tau})){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau'}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{\tau})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;(\V{idx}\;\mathbin{:}\;\V{\tau'}\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;\Con{list}\;\V{\tau}){}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}cons\textrm{-}new}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau_0}\;\V{\tau_1}\;\V{x₀}\;\V{x₁}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcap}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;((\V{x'}\;\red{,}\,\;\Con{listcons}\;\V{\tau})\;\Con{::}\;\V{\Gamma}){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau_0}\;\V{\tau_1}\mskip1.5mu\}\;\rightarrow\;\V{\tau_0}\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\tau_1}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\Con{listcons}\;\V{\tau})\;\Con{::}\;\V{\Gamma}{}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}cons\textrm{-}upd}\;{}\<[26]%
-\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau_0}\;\V{\tau_1}\;\V{\tau_2}\;\V{x₀}\;\V{x₁}\;\V{x'}\mskip1.5mu\}\;\rightarrow\;(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau_2})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcap}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
-\\
-\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x'}\;\red{,}\,\;\Con{listcons}\;\V{\tau})){}\<[E]%
+\>[26]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{\tau_0}\;\V{\tau_1}\;\V{\tau_2}\mskip1.5mu\}\;\rightarrow\;\V{\tau_0}\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\tau_1}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;\V{\tau_2}\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
+\\
+\>[26]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;\Con{listcons}\;\V{\tau}){}\<[E]%
 \ColumnHook
@@
 position-dependent naming scheme. For example, we use a well-typed \emph{de Bruijn}
-index \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma}}, which witnesses the existence of an element \ensuremath{(\V{x}\;\red{,}\,\;\V{\tau})} in \ensuremath{\V{\Gamma}},
+index \ensuremath{\V{\tau}\;\D{\in}\;\V{\Gamma}}, which witnesses the existence of an element \ensuremath{\V{\tau}} in \ensuremath{\V{\Gamma}},
 as defined by the standard library \ensuremath{\D{\_\hspace{-2pt}\in\hspace{-2pt}\_}} operator. This technique is well-known for
@@
 list, however storing its value in an existing variable, represented by the \emph{de Bruijn}
-index \ensuremath{\V{idx}\;\mathbin{:}\;(\V{x'}\;\red{,}\,\;\V{\tau'})\;\D{\in}\;\V{\Gamma}}.
+index \ensuremath{\V{idx}\;\mathbin{:}\;\V{\tau'}\;\D{\in}\;\V{\Gamma}}.
 The constructors \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}}
@@
 used to create a new list. The first creates a new variable, and the second updates an
-existing variable. The list is created from two variables, \ensuremath{(\V{x₀}\;\red{,}\,\;\V{\tau_0})\;\D{\in}\;\V{\Gamma}} and \ensuremath{(\V{x₁}\;\red{,}\,\;\V{\tau_1})\;\D{\in}\;\V{\Gamma}},
+existing variable. The list is created from two variables, \ensuremath{\V{\tau_0}\;\D{\in}\;\V{\Gamma}} and \ensuremath{\V{\tau_1}\;\D{\in}\;\V{\Gamma}},
 which are represented as \emph{de Bruijn} indices. The type of the new list is defined by
@@
 can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the
-type \ensuremath{\V{\tau_1}\;\D{\sqcap}\;\V{\tau_2}\;\D{\sim}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
+type \ensuremath{\V{\tau_1}\;\D{\sqcup}\;\V{\tau_2}\;\D{\sim}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
 The resulting typing context adds information about the newly created list.
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{13}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathkw{data}\;\D{\_\sqcap\_\sim\_}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\>[B]{}\mathkw{data}\;\D{\_\sqcup\_\sim\_}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub4}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{list}\;\V{t})\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub2}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\\
+\>[3]{}\hsindent{8}{}\<[11]%
+\>[11]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;\V{t'}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub2b}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\\
+\>[3]{}\hsindent{9}{}\<[12]%
+\>[12]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
 \\
 \>[3]{}\hsindent{10}{}\<[13]%
-\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcap}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub6}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{listcons}\;\V{t})\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
 \\
 \>[3]{}\hsindent{10}{}\<[13]%
-\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcap}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
+\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
 \ColumnHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcap\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
-Appel and Leroy~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
+\>[B]{}\F{\sqcup\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We omit the complete definition of \ensuremath{\F{\sqcup\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
+Appel et. al.~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
 subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcap\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
-\\
-\>[B]{}\F{\sqcap\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
+\>[B]{}\F{\sqcup\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
+\\
+\>[B]{}\F{\sqcup\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
 \ColumnHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\V{∃}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t}){}\<[E]%
+\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\V{∃}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t}){}\<[E]%
 \\
@@
 list of well-typed values. And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
-Both type environments are used to type block instructions and sequences of
+Both typing contexts are used to type block instructions and sequences of
 block instructions.
@@
 \\
-\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\D{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
+\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\D{All}\;\D{Val}\;\V{\Gamma}{}\<[E]%
 \ColumnHook
@@
 From now on we describe how we implement some parts\footnote{The complete evaluation
-function can be found in our online repository.} of the dynamic semantics (reduction rules)
+function can be found in our online repository~\cite{list-rep}.} of the dynamic semantics (reduction rules)
 of the list-machine language in the function \ensuremath{\F{run\textrm{-}step}}. We mix the code with the explanation
@@
 \paragraph{Conditional jump}{We show next only the case when the jump actually occurs,
-following the rule \emph{step-branch-taken} of \cite{Appel07}. In this case, variable \ensuremath{\V{v}} has value \ensuremath{\Con{nil}}, and
+following the rule \emph{step-branch-taken} of \cite{Appel07}. In this case, variable \ensuremath{\V{v}}
+has value \ensuremath{\Con{nil}}, and
 the step of evaluation should proceed with the block instruction defined in program \ensuremath{\V{p}}, with
@@
 the block instruction with index \ensuremath{\V{i}} on program \ensuremath{\V{p}}. Since we use \emph{de Bruijn} indices to
-represent the label, only valid values are accepted by the intrinsically-typed syntax.}
-
+represent the label, only valid values are accepted by the intrinsically-typed syntax. We use
+Agda's standard library function \ensuremath{\F{[]\textrm{=}{\Rightarrow}lookup}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{[]\textrm{=}{\Rightarrow}lookup}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{px}\;\mathbin{:}\;\D{P}\;\V{x}\mskip1.5mu\}\;\{\mskip1.5mu \V{pxs}\;\mathbin{:}\;\D{All}\;\D{P}\;\V{xs}\mskip1.5mu\}\;\{\mskip1.5mu \V{i}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs}\mskip1.5mu\}\;\rightarrow\;{}\<[E]%
+\\
+\>[B]{}\hsindent{14}{}\<[14]%
+\>[14]{}\V{pxs}\;\D{[}\;\V{i}\;\D{]=}\;\V{px}\;\rightarrow\;{}\<[E]%
+\\
+\>[B]{}\hsindent{14}{}\<[14]%
+\>[14]{}\F{lookupA}\;\V{pxs}\;\V{i}\;\D{\equiv}\;\V{px}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+which rewrites a lookup predicate (\ensuremath{\V{pxs}\;\D{[}\;\V{i}\;\D{]=}\;\V{px}}) into an equality using the \ensuremath{\F{lookupA}}
+function, \ensuremath{\F{lookupA}\;\V{pxs}\;\V{i}\;\D{\equiv}\;\V{px}}.}
 \begin{hscode}\SaveRestoreHook
@@
 it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}. This process is done by the
-function \ensuremath{\F{update\textrm{-}env}}\footnote{The source-code of this function can be found in our online repository.}.}
+function \ensuremath{\F{update\textrm{-}env}}\footnote{The source-code of this function can be found in our online
+repository~\cite{list-rep}.}.}
 
@@
 and others lemmas to deal with the least common supertype, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
-and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}. These lemmas and their proofs can be found in our repository online. }
+and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}. These lemmas and their proofs can be found in our repository online~\cite{list-rep}. }
 
@@
 
-In practice, a source-code of a programming language runs through several phases, including lexing, parsing,
-scope checking, and most importantly \emph{type-checking}. Since we represent programs using a
-intrinsically-typed syntax, scope and type-checking is only a matter of elaborating an untyped syntax
-to a typed one.
-
-Since we use \emph{de Bruijn} indices to represent labels and variables, as the first step to type and scope
-check them, we need to provide an index from a named variable. This process is done by the
-\ensuremath{\F{lookup\textrm{-}var}} function, returning an index when the named variable is found, and \ensuremath{\Con{nothing}} otherwise.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{lookup\textrm{-}var}\;{}\<[13]%
-\>[13]{}\mathbin{:}\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\rightarrow\;(\V{x}\;\mathbin{:}\;\D{String})\;{}\<[E]%
-\\
-\>[13]{}\rightarrow\;\D{Maybe}\;(\D{\exists}\;\lambda\;\V{\tau}\;\rightarrow\;(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma}){}\<[E]%
-\\
-\>[B]{}\F{lookup\textrm{-}var}\;\Con{\lbrack\:\rbrack}\;\V{x}\;\mathrel{=}\;\Con{nothing}{}\<[E]%
-\\
-\>[B]{}\F{lookup\textrm{-}var}\;((\V{y}\;\red{,}\,\;\V{\tau})\;\Con{::}\;\V{\Gamma})\;\V{x}\;\mathkw{with}\;\V{x}\;\D{=}\;\V{y}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{yes}\;\Con{refl}\;\mathrel{=}\;\Con{just}\;(\V{\tau}\;\red{,}\,\;\Con{here}\;\Con{refl}){}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;(\Con{no}\;\V{¬p})\;\mathrel{=}\;\F{map}\;{}\<[22]%
-\>[22]{}(\lambda\;\V{v}\;\rightarrow\;\D{proj_1}\;\V{v}\;\red{,}\,\;\Con{there}\;(\D{proj_2}\;\V{v}))\;{}\<[E]%
-\\
-\>[22]{}(\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{x}){}\<[E]%
+In practice, a source-code of a programming language runs through several phases,
+including lexing, parsing, scope checking, and most importantly \emph{type-checking}.
+Since we represent programs using a intrinsically-typed syntax, scope and
+type-checking is only a matter of elaborating an untyped syntax to a typed one.
+The definition of our type-checker follows McBride's~\cite{McBride2004} approach which
+specifies the type-checker function type as its correctness property: it should return
+a type annotated syntax such that it's erasure is equal to the type-checker input program.
+Before considering machine instructions, we will detail the type-checker correctness
+approach in the context of variables.
+Since we use \emph{de Bruijn} indices to represent labels and variables, as the first
+step to type and scope check them, we need to get the index from a given variable.
+The function \ensuremath{\F{index}} return the list position corresponding to a membership proof (variable):
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{index}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{A}\;\mathbin{:}\;\D{Set}\mskip1.5mu\}\;\{\mskip1.5mu \V{t}\;\mathbin{:}\;\V{A}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Gamma}\;\mathbin{:}\;\D{List}\;\V{A}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\D{\mathbb{N}}{}\<[E]%
+\\
+\>[B]{}\F{index}\;(\Con{here}\;\V{px})\;\mathrel{=}\;\Con{zero}{}\<[E]%
+\\
+\>[B]{}\F{index}\;(\Con{there}\;\V{p})\;\mathrel{=}\;\Con{suc}\;(\F{index}\;\V{p}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Next, we follow Norell's~\cite{Norell2009} approach and define a type that relates a
+natural number with a list as follows. Either the number correspond to some list position,
+in which it is of the form \ensuremath{\F{index}\;\V{p}} for some proof \ensuremath{\V{p}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs}} or it is invalid position.
+The type \ensuremath{\D{Lookup}} encodes this relation between lists and natural numbers.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{Lookup}\;\{\mskip1.5mu \V{A}\;\mathbin{:}\;\D{Set}\mskip1.5mu\}\;(\V{xs}\;\mathbin{:}\;\D{List}\;\V{A})\;\mathbin{:}\;\D{\mathbb{N}}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{inside}\;\mathbin{:}\;(\V{x}\;\mathbin{:}\;\V{A})\;(\V{p}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs})\;\rightarrow\;\D{Lookup}\;\V{xs}\;(\F{index}\;\V{p}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{outside}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\rightarrow\;\D{Lookup}\;\V{xs}\;\V{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Using this infrastructure we can build the function \ensuremath{\F{lookup\textrm{-}var}} which
+returns a value of type \ensuremath{\D{Lookup}\;\V{xs}\;\V{n}} that specifies whether \ensuremath{\V{n}} is
+a valid list position or not.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{lookup\textrm{-}var}\;\mathbin{:}\;\{\mskip1.5mu \V{A}\;\mathbin{:}\;\D{Set}\mskip1.5mu\}\;(\V{xs}\;\mathbin{:}\;\D{List}\;\V{A})\;(\V{n}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{Lookup}\;\V{xs}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;\Con{\lbrack\:\rbrack}\;\V{n}\;\mathrel{=}\;\Con{outside}{}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;(\V{x}\;\Con{::}\;\V{xs})\;\Con{zero}\;\mathrel{=}\;\Con{inside}\;\V{x}\;(\Con{here}\;\Con{refl}){}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;(\V{x}\;\Con{::}\;\V{xs})\;(\Con{suc}\;\V{n})\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{xs}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;(\V{x}\;\Con{::}\;\V{xs})\;(\Con{suc}\;.\;(\F{index}\;\V{p}))\;\mid \;\Con{inside}\;\V{y}\;\V{p}\;\mathrel{=}\;\Con{inside}\;\V{y}\;(\Con{there}\;\V{p}){}\<[E]%
+\\
+\>[B]{}\F{lookup\textrm{-}var}\;(\V{x}\;\Con{::}\;\V{xs})\;(\Con{suc}\;\anonymous )\;\mid \;\Con{outside}\;\mathrel{=}\;\Con{outside}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+In order to type-check instructions using this approach, we need to first define a
+datatype for untyped programs.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{UInstr}\;(\V{n}\;\mathbin{:}\;\D{\mathbb{N}})\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{jump}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;(\V{x}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{Label}\;\V{n}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{branch\textrm{-}if\textrm{-}nil}\;\mathbin{:}\;(\V{v}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{Label}\;\V{n}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{fetch\textrm{-}field\textrm{-}0}\;\mathbin{:}\;(\V{v}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;(\V{v′}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{fetch\textrm{-}field\textrm{-}1}\;\mathbin{:}\;(\V{v}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;(\V{v′}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{get\textrm{-}label}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;(\V{v}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{Label}\;\V{n}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{cons}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;(\V{v₀}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;(\V{v₁}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;(\V{v′}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{halt}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{seq}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;\D{UInstr}\;\V{n}\;\rightarrow\;\D{UInstr}\;\V{n}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Type \ensuremath{\D{UInstr}} has a constructor for each of the list machine instructions,
+type \ensuremath{\D{Label}\;\V{n}} denote a program label and it is represented by
+a value of type \ensuremath{\D{Fin}\;\V{n}}. Next, we define an erasure function for
+our intrinsically typed syntax:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}seq}\;\V{p}\;\V{p₁})\;\mathrel{=}\;\Con{seq}\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{p})\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{p₁}){}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}nil}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{x}\;\V{x₁}\;\V{x₂})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{x})\;\V{l}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}0}\;(\F{index}\;\V{x})\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{x}\;\V{idx})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}0}\;(\F{index}\;\V{x})\;(\F{index}\;\V{idx}){}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}1}\;(\F{index}\;\V{x})\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;\V{x}\;\V{idx})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}1}\;(\F{index}\;\V{x})\;(\F{index}\;\V{idx}){}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x}\;\V{x₁}\;\V{x₂})\;\mathrel{=}\;\Con{cons}\;(\F{index}\;\V{x})\;(\F{index}\;\V{x₁})\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}cons\textrm{-}upd}\;\V{x}\;\V{x₁}\;\V{idx}\;\V{x₂})\;\mathrel{=}\;\Con{cons}\;(\F{index}\;\V{x})\;(\F{index}\;\V{x₁})\;(\F{index}\;\V{idx}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Finally, we can define a view of an untyped instruction (\ensuremath{\D{UInstr}\;\V{n}}) as being the
+erasure of an intrinsically-typed one using the type \ensuremath{\D{CheckedInstr}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{CheckedInstr}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n})\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\mathbin{:}\;\D{UInstr}\;\V{n}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{ok}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;(\V{i}\;\mathbin{:}\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'})\;\rightarrow\;\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{i}){}\<[E]%
 \ColumnHook
@@
 For brevity, we show how we type-check only one instruction\footnote{The curious reader can refer to our
-online repository for a complete implementation.}. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
-receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value,
-which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type-checks.
-
+online repository~\cite{list-rep} for a complete implementation.}. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
+receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\D{TC}\;\mathbin{:}\;\D{Set}\;\rightarrow\;\D{Set}{}\<[E]%
+\\
+\>[B]{}\D{TC}\;\V{i}\;\mathrel{=}\;\D{Either}\;\D{TypeError}\;\V{i}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+which is either an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type-checks.
+The \ensuremath{\D{TypeError}} contains one constructor for each of the possible failures causes for an
+input program.
 \begin{hscode}\SaveRestoreHook
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\mathbin{:}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx})\;\rightarrow\;(\V{\Gamma}\;\mathbin{:}\;\D{Ctx})\;\rightarrow\;(\V{v}\;\mathbin{:}\;\D{String})\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\rightarrow\;(\V{v'}\;\mathbin{:}\;\D{String})\;\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{fetch\textrm{-}field\textrm{-}0}\;\V{v}\;\V{v'})){}\<[E]%
-\\
-\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{v'}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}\;\red{,}\,\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v'}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;(\text{\ttfamily \char34 variable~out~of~scope:~\char34}\;\F{++}\;\V{v}){}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{nil}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error~[nil]\char34}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{list}\;\V{\tau}\;\red{,}\,\;\anonymous )\;\red{,}\,\;\anonymous \;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 type~error~[list]\char34}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{listcons}\;\V{\tau}\;\red{,}\,\;\V{idx})\;\red{,}\,\;{}\<[34]%
-\>[34]{}\Con{nothing}\;\mathrel{=}\;{}\<[E]%
-\\
-\>[34]{}\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{idx})){}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\Con{listcons}\;\V{\tau}\;\red{,}\,\;\V{idx})\;\red{,}\,\;{}\<[34]%
-\>[34]{}\Con{just}\;(\V{\tau'}\;\red{,}\,\;\V{idx'})\;\mathrel{=}\;{}\<[E]%
-\\
-\>[34]{}\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{idx}\;\V{idx'})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-In the code above, we use the function \ensuremath{\F{lookup\textrm{-}var}} to provide the \emph{de Bruijn} indices for
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{data}\;\D{TypeError}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{UnexpectedTy}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{TypeError}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{UndefinedVar}\;\mathbin{:}\;\D{\mathbb{N}}\;{}\<[21]%
+\>[21]{}\rightarrow\;\D{TypeError}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{UndefinedLabel}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\rightarrow\;\D{Label}\;\V{n}\;\rightarrow\;\D{TypeError}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+In the code below, we use the function \ensuremath{\F{lookup\textrm{-}var}} to provide the \emph{de Bruijn} indices for
 each variable, and match the first with its possible forms. The first three cases indicate
-type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{nothing}} it means a variable scope error; (2) and (3) are typing errors,
+type errors: (1) when \ensuremath{\V{v}} is \ensuremath{\Con{outside}} it means a variable scope error; (2) and (3) are typing errors,
 since the type of variable \ensuremath{\V{v}} should be a \ensuremath{\Con{listcons}}. Last two cases represent that the instruction
 is well-typed. The process for type-checking different instructions follows a similar setting.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n})\;\V{\Gamma}\;\V{v}\;\V{v'}\;\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{fetch\textrm{-}field\textrm{-}0}\;\V{v}\;\V{v'})){}\<[E]%
+\\
+\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{v'}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}\;\mid \;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{outside}\;\mid \;\anonymous \;\mathrel{=}\;\Con{undefined\textrm{-}var}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{inside}\;\Con{nil}\;\V{p}\;\mid \;\V{q}\;\mathrel{=}\;\Con{type\textrm{-}error}\;\Con{nil}{}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{inside}\;(\Con{list}\;\V{x})\;\V{p}\;\mid \;\V{q}\;\mathrel{=}\;\Con{type\textrm{-}error}\;(\Con{list}\;\V{x}){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{inside}\;(\Con{listcons}\;\V{x})\;\V{p}\;\mid \;\Con{inside}\;\V{x₁}\;\V{p₁}\;\mathrel{=}\;\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{p}\;\V{p₁})){}\<[E]%
+\\
+\>[B]{}\V{...}\;\mid \;\Con{inside}\;(\Con{listcons}\;\V{x})\;\V{p}\;\mid \;\Con{outside}\;\mathrel{=}\;\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{p})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Finally, functions \ensuremath{\Con{type\textrm{-}error}} and \ensuremath{\Con{undefined\textrm{-}var}} just build the corresponding
+type-checker error as follows:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Con{type\textrm{-}error}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{i}\mskip1.5mu\}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{TC}\;\V{i}{}\<[E]%
+\\
+\>[B]{}\Con{type\textrm{-}error}\;\mathrel{=}\;\Con{left}\;\V{∘}\;\Con{UnexpectedTy}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Con{undefined\textrm{-}var}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{i}\mskip1.5mu\}\;\rightarrow\;\D{\mathbb{N}}\;\rightarrow\;\D{TC}\;\V{i}{}\<[E]%
+\\
+\>[B]{}\Con{undefined\textrm{-}var}\;\mathrel{=}\;\Con{left}\;\V{∘}\;\Con{UndefinedVar}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 In this section we describe the necessary modifications to our formalization to include indirect jumps
-as proposed by Appel and Leroy~\cite{AppelDL12}. We start by describing the changes on the type system
+as proposed by Appel et. al.~\cite{AppelDL12}. We start by describing the changes on the type system
 and its impact on the intrinsically typed representation of programs in Section~\ref{sec:changestypes}.
@@
 
-Since our solution relies on representing the program using intrinsically typed syntax, we describe
-first the changes on the list-machine type system. The first modification is the inclusion of a type for
+Since our solution relies on representing the program using intrinsically typed syntax, we first
+describe the changes on the list-machine type system. The first modification is the inclusion of a type for
 program labels, which denotes continuations in a program.
@@
 
-The type \ensuremath{\Con{cont}} $\Gamma$ for a label express that the machine can safely jump when its registers
-satisfies $\Gamma$. The inclusion of continuation types makes the type and context syntax mutually
-inductive.
+The type \ensuremath{\Con{cont}} $\Gamma$ is given to program labels and it
+express that the machine can safely jump when its registers
+satisfies $\Gamma$. The inclusion of continuation types makes
+the type and context syntax mutually inductive.
 
@@
 since now we have ``incomparable'' types: list types are not subtypes neither supertypes for continuations.
-The solution is to complete the subtyping relation with a bottom and top types. The subtyping rules for these
+The solution is to complete the subtyping relation with bottom and top types. The subtyping rules for these
 new types are presented next.
@@
 \begin{array}{cc}
-  \infer{\bot <: \tau}{} &
-  \infer{\tau <: \top}{} \\ & \\
+  \inference{}{\bot <: \tau}[subtype-bot] &
+  \inference{}{\tau <: \top}[subtype-top] \\ & \\
   \multicolumn{2}{c}{
-    \infer{cont\,\Gamma_1\:<:\:cont\,\Gamma_2}{\Gamma_2 \subset \Gamma_1}
+    \inference{\Gamma_2 \subset \Gamma_1}{cont\,\Gamma_1\:<:\:cont\,\Gamma_2}[subtype-cont]
   }
@@
 \[
-\begin{array}{cc}
-  \infer{cont\:\Gamma\:\sqcup\:nil = \top}{} & \infer{cont\:\Gamma\:\sqcup\:(list\:\tau) = \top}\\ & \\
-  \multicolumn{2}{c}{\infer{cont\:\Gamma\:\sqcup\:(listcons\:\tau) = \top}}
+\begin{array}{c}
+  \inference{}{cont\:\Gamma\:\sqcup\:nil = \top}[lub-cont-nil] \\ \\
+  \inference{}{cont\:\Gamma\:\sqcup\:(list\:\tau) = \top}[lub-cont-list]\\ & \\
+  \inference{}{cont\:\Gamma\:\sqcup\:(listcons\:\tau) = \top}[lub-cont-listcons]
 \end{array}
@@
 \[
-\begin{array}{cc}
-  \infer{\tau \sqcup \bot = \tau}{} & \infer{\tau \sqcup \top = \top}{}
+\begin{array}{c}
+  \inference{}{\tau \sqcup \bot = \tau}[lub-bot] \\ \\
+  \inference{}{\tau \sqcup \top = \top}[lub-top]
 \end{array}
@@
 The last set of rules for the least upper bound relation deals with continuations.
-Because of the contravariance of the \ensuremath{\Con{cont}} type, we also need to define greatest lower bounds for both
-types and typing contexts. First, we present the rule for the least upper bound for continuation types.
+Because of the contravariance of the \ensuremath{\Con{cont}} type, we also need to define the greatest
+lower bound relation for both
+types and typing contexts.
+First, we present the rule for the least upper bound for continuation types.
 
@@
 \begin{array}{c}
-  \infer{cont\:\Gamma_1\:\sqcap\: cont\:\Gamma_2=cont\:\Gamma_3}{\Gamma_1\cup\Gamma_2 = \Gamma_3}
+  \inference{\Gamma_1\cap\Gamma_2 = \Gamma_3}
+            {cont\:\Gamma_1\:\sqcup\: cont\:\Gamma_2=cont\:\Gamma_3}
+            [lub-cont]
 \end{array}
@@
 \ensuremath{\Con{cont}} $\Gamma_3$, where $\Gamma_3$ is greatest lower bound for typing contexts $\Gamma_1$ and $\Gamma_2$.
-The definition of the greatest lower bound for contexts is presented next.
+The definition of the greatest lower bound for contexts is as follows:
 \[
-  \begin{array}{cc}
-    \infer{\{\} \cup \Gamma = \Gamma}{} &
-    \infer{\Gamma \cup \{\} = \Gamma}{} \\ & \\
-    \multicolumn{2}{c}{
-      \infer{(\Gamma_1 , x : \tau_1) \cup (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
-            {\tau_1 \sqcup \tau_2 = \tau_3 & \Gamma_1 \cup \Gamma_2 = \Gamma_3}
-    }
+  \begin{array}{c}
+    \inference{}{\{\} \cup \Gamma = \Gamma}[glb-empty-left] \\ \\
+    \inference{}{\Gamma \cup \{\} = \Gamma}[glb-empty-right] \\ \\
+      \inference{\tau_1 \sqcup \tau_2 = \tau_3 & \Gamma_1 \cup \Gamma_2 = \Gamma_3}
+                {(\Gamma_1 , x : \tau_1) \cup (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
+                [glb-cons]
   \end{array}
@@
 The definition of the greatest lower bound for types is as follows. The first set of rules specify the reflexivity,
-that $`\top$ type is an identity and $`\bot$ is for greastest lower bound for any type $\tau$.
-Again, rules to ensure commutativity of the relation are ommited for brevity.
+that $\top$ type is an identity and $\bot$ is for greastest lower bound for any type $\tau$.
+Again, rules to ensure commutativity of the relation are omitted for brevity.
 
 \[
-\begin{array}{ccc}
-  \infer{\tau \sqcup \tau = \tau}{} & \infer{\top \sqcup \tau = \tau}{} & \infer{\bot \sqcup \tau = \bot}{}\\
+\begin{array}{c}
+  \inference{}{\tau \sqcap \tau = \tau}[glb-refl] \\ \\
+  \inference{}{\top \sqcap \tau = \tau}[glb-top] \\ \\
+  \inference{}{\bot \sqcap \tau = \bot}[glb-bot]
 \end{array}
@@
 \[
-\begin{array}{cc}
-  \infer{nil \sqcup (listcons\: \tau) = \bot}{} &
-  \infer{(list\: \tau_1)\sqcup (list\:\tau_2) = list\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
-   \infer{(listcons\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} &
-  \infer{(list\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
+\begin{array}{c}
+  \inference{}{nil \sqcap (listcons\: \tau) = \bot}[glb-nil-listcons] \\ \\
+  \inference{\tau_1\sqcap \tau_2 = \tau_3}
+            {(list\: \tau_1)\sqcap (list\:\tau_2) = list\:\tau_3}
+            [glb-list]\\ \\
+   \inference{\tau_1\sqcap \tau_2 = \tau_3}
+             {(listcons\: \tau_1)\sqcap (listcons\:\tau_2) = listcons\:\tau_3}
+             [glb-listcons] \\ \\
+
+   \inference{\tau_1\sqcap \tau_2 = \tau_3}
+             {(list\: \tau_1)\sqcap (listcons\:\tau_2) = listcons\:\tau_3}
+             [glb-list-listcons]
 \end{array}
@@
 Rules for the greatest lower bound relation for continuation types show that the bottom type is the
-lower bound for the list and continuation types. The other rule stabishes the compatibility of
+lower bound for the list and continuation types. The other rule estabishes the compatibility of
 the continuation type with the lower bound relation and it uses the least upper bound for typing
@@
 \[
-\begin{array}{ccc}
-  \infer{nil\sqcup(cont\:\Gamma) = \bot}{} &
-  \infer{(list\:\tau)\sqcup(cont\:\Gamma) = \bot}{} &
-  \infer{(listcons\:\tau)\sqcup(cont\:\Gamma) = \bot}{} \\ & & \\
-  \multicolumn{3}{c}{
-    \infer{(cont\:\Gamma_1)\sqcup(cont\:\Gamma_2) = (cont\:\Gamma_3)}{\Gamma_1\cap\Gamma_2 = \Gamma_3}
-  }
+\begin{array}{c}
+  \inference{}{nil\sqcap(cont\:\Gamma) = \bot}[glb-cont-nil] \\ \\
+  \inference{}{(list\:\tau)\sqcap(cont\:\Gamma) = \bot}[glb-cont-list] \\ \\
+  \inference{}{(listcons\:\tau)\sqcap(cont\:\Gamma) = \bot}[glb-cont-listcons] \\ \\
+  \inference{\Gamma_1\cup\Gamma_2 = \Gamma_3}
+            {(cont\:\Gamma_1)\sqcap(cont\:\Gamma_2) = (cont\:\Gamma_3)}
+            [glb-cont]
 \end{array}
@@
 \[
-  \begin{array}{cc}
-    \infer{\{\} \cap \Gamma = \{\}}{} &
-    \infer{\Gamma \cap \{\} = \{\}}{} \\ & \\
-    \multicolumn{2}{c}{
-      \infer{(\Gamma_1 , x : \tau_1) \cap (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
-            {\tau_1 \sqcap \tau_2 = \tau_3 & \Gamma_1 \cap \Gamma_2 = \Gamma_3}
-    }
+  \begin{array}{c}
+    \inference{}{\{\} \cap \Gamma = \{\}}[glb-empty-left] \\ \\
+    \inference{}{\Gamma \cap \{\} = \{\}}[glb-empty-right] \\ \\
+    \inference{\tau_1 \sqcap \tau_2 = \tau_3 & \Gamma_1 \cap \Gamma_2 = \Gamma_3}
+              {(\Gamma_1 , x : \tau_1) \cap (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
+              [glb-cons]
   \end{array}
 \]
-The Agda encoding of these relations is an immediate translation of the mathematical notation
-to code and are omitted for brevity.
+The Agda encoding of these relations is an immediate translation of the mathematical notation.
 
@@
 Once we have only modified the subtyping and the least upper bound relations, most of the type system
-rules remained unchanged from the original version. In order to support indirect jumps, Appel and Leroy.
+rules remained unchanged from the original version. In order to support indirect jumps, Appel's
 approach is to modify the jump instruction and add a new instruction for loading label values into
@@
 \[
-\begin{array}{cc}
-  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:L_0\}\Gamma'}
-        {\Gamma[v\,:=\,nil] = \Gamma'} &
-  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:l\}\Gamma'}
-        {\Pi(l) = \Gamma_1 & \Gamma[v\,:=\,cont\:\Gamma_1] = \Gamma'} \\ & \\
-  \multicolumn{2}{c}{
-    \infer{\Pi;\Gamma\vdash_{block}\,\textrm{jump}\:v}
-          {\Gamma(v) = cont\:\Gamma_1 & \Gamma \subset \Gamma_1}
-  }
+\begin{array}{c}
+  \inference{\Gamma[v\,:=\,nil] = \Gamma'}
+            {\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:L_0\}\Gamma'}
+            [check-instr-get-label-0] \\ \\
+  \inference{\Pi(l) = \Gamma_1 & \Gamma[v\,:=\,cont\:\Gamma_1] = \Gamma'}
+            {\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:l\}\Gamma'}
+            [check-instr-get-label]\\ \\
+  \inference{\Gamma(v) = cont\:\Gamma_1 & \Gamma \subset \Gamma_1}
+            {\Pi;\Gamma\vdash_{block}\,\textrm{jump}\:v}
+            [check-block-jump]
 \end{array}
@@
 is $nil$ and the second specifies that the type for any other label is a continuation type formed by
-the context associated with label $l$ in $\Pi$. The last rule show that typing a jump instruction
+the context associated with label $l$ in $\Pi$. The last rule shows that typing a jump instruction
 requires that its register has a continuation type, $cont\:\Gamma_1$, that should be a subtype of
 the current block context $\Gamma$. Based on these rules, we add the following constructors to our
-instruction typed syntax:
+typed instruction syntax:
 
@@
 \column{25}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{40}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\Con{instr\textrm{-}getlabel\textrm{-}0}\;{}\<[25]%
-\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[25]{}\hsindent{10}{}\<[35]%
-\>[35]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{nil})){}\<[E]%
+\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\tau}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;\V{\tau}\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[25]{}\hsindent{8}{}\<[33]%
+\>[33]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;\Con{nil}){}\<[E]%
 \\
@@
 \>[3]{}\Con{instr\textrm{-}getlabel}\;{}\<[25]%
-\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\tau}\;\V{\Gamma_1}\;\V{x}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;(\V{x}\;\red{,}\,\;\V{\tau})\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
-\\
-\>[25]{}\hsindent{15}{}\<[40]%
-\>[40]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;{}\<[E]%
-\\
-\>[25]{}\hsindent{15}{}\<[40]%
-\>[40]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;(\V{x}\;\red{,}\,\;\Con{cont}\;\V{\Gamma_1})){}\<[E]%
+\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\tau}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;(\V{idx}\;\mathbin{:}\;\V{\tau}\;\D{\in}\;\V{\Gamma})\;{}\<[E]%
+\\
+\>[25]{}\hsindent{13}{}\<[38]%
+\>[38]{}\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;{}\<[E]%
+\\
+\>[25]{}\hsindent{13}{}\<[38]%
+\>[38]{}\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;(\V{idx}\;\F{::=}\;\Con{cont}\;\V{\Gamma_1}){}\<[E]%
 \ColumnHook
@@
 \>[3]{}\Con{block\textrm{-}jump}\;{}\<[25]%
-\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{x}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;(\V{x}\;\red{,}\,\;\Con{cont}\;\V{\Gamma_1})\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
+\>[25]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;\Con{cont}\;\V{\Gamma_1}\;\D{\in}\;\V{\Gamma}\;{}\<[E]%
 \\
@@
 we include a new constructor on type \ensuremath{\D{UInstr}} to represent untyped get-label instruction and
-add its corresponding equation on function \ensuremath{\F{forget\textrm{-}types\textrm{-}instr}}, which erases type information
-from the instrinsically typed syntax. This function is used by the type-checker to ensure its
-correctness.
+add its corresponding equation on function \ensuremath{\F{forget\textrm{-}types\textrm{-}instr}}.
 \begin{hscode}\SaveRestoreHook
@@
 \>[3]{}\Con{get\textrm{-}label}\;{}\<[17]%
-\>[17]{}\mathbin{:}\;(\V{v}\;\mathbin{:}\;\V{ℕ})\;\rightarrow\;\D{Label}\;\rightarrow\;\D{UInstr}{}\<[E]%
+\>[17]{}\mathbin{:}\;(\V{v}\;\mathbin{:}\;\D{\mathbb{N}})\;\rightarrow\;\D{Label}\;\rightarrow\;\D{UInstr}{}\<[E]%
 \\[\blanklineskip]%
@@
 \\
-\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\Con{L_{0}}{}\<[E]%
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\Con{zero}{}\<[E]%
 \\
@@
 Equations for erasing type information are immediate. The only peculiarity is that when the instruction
-\ensuremath{\Con{get\textrm{-}label}} is typed using rule \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} we consider the initial label \ensuremath{\Con{L_{0}}}. Next, we implement
-the type checker logic for \ensuremath{\Con{get\textrm{-}label}}. Function \ensuremath{\F{check\textrm{-}get\textrm{-}label}} verifies if the label is present at
+\ensuremath{\Con{get\textrm{-}label}} is typed using rule \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} we consider the initial label \ensuremath{\Con{zero}}.
+Next, we implement the type checker logic for \ensuremath{\Con{get\textrm{-}label}}. Function \ensuremath{\F{check\textrm{-}get\textrm{-}label}} verifies if the label is present at
 the program global typing context and if the variable is present on the current block typing context.
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{30}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \\
-\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mid \;\V{\Gamma_1}\;\F{with\equiv}\;\V{p}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 check-get-label:~variable~out~of~scope\char34}{}\<[E]%
-\\
-\>[B]{}\V{...}\;\mid \;\Con{just}\;(\V{\tau}\;\red{,}\,\;\V{idx})\;\mathrel{=}\;\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}getlabel}\;\V{idx}\;(\F{lookup⇒[]=}\;\V{l}\;\V{\Pi}\;\V{p}))){}\<[E]%
+\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mid \;{}\<[28]%
+\>[28]{}\V{\Gamma_1}\;\F{with\equiv}\;\V{p}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}{}\<[E]%
+\\
+\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;.\;(\F{index}\;\V{p₁})\;\V{l}\;\mid \;\V{\Gamma_1}\;\F{with\equiv}\;\V{p}\;\mid \;\Con{inside}\;\V{x}\;\V{p₁}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathrel{=}\;\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}getlabel}\;\V{p₁}\;(\F{lookup⇒[]=}\;\V{l}\;\V{\Pi}\;\V{p}))){}\<[E]%
+\\
+\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mid \;\V{\Gamma_1}\;\F{with\equiv}\;\V{p}\;\mid \;\Con{outside}\;\mathrel{=}\;\Con{undefined\textrm{-}var}\;\V{v}{}\<[E]%
 \ColumnHook
@@
 the changes on the typed syntax, since now we have a new value for continuation types. A continuation
-value simply stores the label of the corresponding block.
+value stores the environment for its corresponding block.
 
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Con{cont}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;\D{Val}\;(\Con{cont}\;\V{\Gamma_1}){}\<[E]%
+\>[5]{}\Con{cont}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;\D{Env}\;\V{\Gamma_1}\;\rightarrow\;\D{Val}\;(\Con{cont}\;\V{\Gamma_1}){}\<[E]%
 \ColumnHook
@@
 We implemented all 14 tasks from the list-machine benchmark in the Agda programming language.
-The tasks considered by us are the same implemented and proved by Appel and Leroy.~\cite{Appel07}.
+The tasks considered by us are the same implemented and proved by Appel et. al.~\cite{Appel07}.
 The next table summarizes the total number of lines of code (LOC) for our results together with theirs.
@@
     & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
-1.  & Operational Semantics                        & 126                       & 98                      & 109                      \\
+1.  & Operational Semantics                        & 126                       & 98                      & 101                      \\
 2.  & Derive $p \Downarrow$                        & 1                         & 8                       & 1                        \\ \hline
-3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
+3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 50                       \\
 4.  & $\sqcap$ algorithm                           & *                         & *                       & 13                       \\
@@
                                                    & 1                         & no                      & 1                        \\
-6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
-7.  & Prove properties of $\sqcap$                 & 114                       & 21                      & 124                      \\
+6.  & State properties of $\sqcup$                 & 12                        & 13                      & 6                        \\
+7.  & Prove properties of $\sqcup$                 & 114                       & 21                      & 124                      \\
 8.  & State soundness theorem                      & 29                        & 15                      & *                        \\
@@
 14. & Prove soundness of $\vdash_{\textrm{prog}} p : \Pi$
-                                                   & 347                       & 141                     & *
+                                                   & 347                       & 141                     & *                        \\
+15. & Generate~\LaTeX                              & no                        & no                      & no
 \end{tabular}
@@
 on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5. We briefly comment
-on our Agda encoding of these 14 tasks.
+on our Agda encoding of these 15 tasks.
 
@@
         since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
+  \item \textbf{Generate~\LaTeX.} Agda's compiler does have a limited support for typesetting~\LaTeX code.
+        However, we find it lacking some important features like omitting code from typesetted output
+        and use~\LaTeX math-commands instead of unicode characters for mathematical symbols.
 \end{enumerate}
@@
 
-Appel and Leroy's list-machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
+Appel et. al. list-machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
 correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics.
-We stick to our initial proposal based on dependently typed syntax and definitional interpreters. Appel and Leroy's formalization for
-the version 2.0 of the benchmark demanded around 1750 LOC without considering his library for the
-``very modal model''~\cite{Appel07}. Our formalization demanded 1134 LOC and it depends only on the Agda standard library.
-Compared to our solution to the first version of the benchmark (which was implemented in 612 LOC), the inclusion of
-indirect jumps increased the formalization size by around 85\%.
-
-%As we could notice on the previous table, the approach taken in this paper avoids code repetition and decreases the number of LOCs necessary to
-%implement the proposed language and to prove its properties. The formalization of this benchmark took 2898 LOC to be done in Twelf, and 887 LOC
-%in Coq, while our implementation used only 415 LOC to achieve the same result. Our encoding uses approximately 14\% of the LOC when compared to
-%the Twelf formalization, and 47\% when compared to Coq's. The main reason for this difference is the fact that the soundness properties are
-%obtained for free when combining an intrinsically-typed syntax with an implementation of a definitional interpreter in a total language like Agda.
+However, we are not able to prove a lemma which stabilishes a coercion between continuation types:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{postulate}\;\V{sub-env}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Gamma'}\;\D{\subset}\;\V{\Gamma}\;\rightarrow\;\D{Env}\;\V{\Gamma}\;\rightarrow\;\D{Env}\;\V{\Gamma'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Appel et. al. was able to prove this property in Coq by enconding the machine syntax and semantics using
+a shallow embedding based on his ``very modal model''~\cite{Appel07} and its complete formalization
+demanded around 1750 LOC without considering his auxiliar library.
+Our formalization demanded 1134 LOC and it depends only on the Agda standard library.
+Compared to our solution to the first version of the benchmark (which was implemented in 612 LOC),
+the inclusion of indirect jumps increased the formalization size by around 85\%.
 
@@
 
-\paragraph{Definitional interpreters}{
+\paragraph{Definitional interpreters and intrinsically-typed syntax}{
 The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
@@
 Chapman's et al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
-which avoids completely stuck states.}
+which avoids completely stuck states. Another application of intrisincally-typed syntax is Pardo's et. al. work on correct-by-construction
+compilers~\cite{PardoGPV18}. The approach used by the authors is to index the syntax of
+a language with its types and semantics in order to automatically derive a correctness proof for the
+compiler. Authors apply their metodology in an arithmetic expression languages and a imperative
+language with top-level variables, sequencing and looping statements. }
 

### 2021-11-26 papers/COLA/paper.tex
finishing review for COLA

@@
 describe a solution to the second version of the benchmark and compare it with
-Appel'et. al. Coq-based solution.
+Appel's et. al. Coq-based solution.
 \end{abstract}
@@
 machine benchmark~\cite{AppelDL12}. We provide detailed explanations of the needed
-changes on the machine syntax, type system and interpreter to accomodate this extension, which
+changes on the machine syntax, type system and interpreter to accommodate this extension, which
 allows the handling of more complex control flow structures, like the call/return model of function calls.
@@
         types for the machine registers.
-  \item We provide a correct-by-construction implementation of a type-checker for
+  \item We provide a correct-by-construction implementation of a type-checker for the
         list-machine programs. Our type-checker produces, as a result, an intrinsically-typed
@@
 A important application of Agda's dependently-typed pattern matching is
-to rewrite a equality. As an example, consider the following code which
+to rewrite an equality. As an example, consider the following code which
 postulates the commutativity of natural number addition and a predicate
@@
 constructor, leaving the responsibility of checking type safety to the
-host's language type-checker. Before considering our solution to encode list-machine
-programs as instrisincally-typed syntax, we review this approach in a reduced example.
+host language type-checker. Before considering our solution to encode list-machine
+programs as intrinsically-typed syntax, we review this approach in a reduced example.
 
@@
 The other rules specify that type constructors $list$ and $listcons$ respect
-the subtyping relation. The least common supertype $\tau = \tau_1 \sqcap \tau_2$ of
+the subtyping relation. The least common supertype $\tau = \tau_1 \sqcup \tau_2$ of
 $\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
@@
 
-\section{Intrinsically-typed syntax}\label{sec:typing}
+\section{Intrinsically-Typed Syntax}\label{sec:typing}
 
@@
 \\
-\>[26]{}\rightarrow\;\V{\tau_1}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
+\>[26]{}\rightarrow\;\V{\tau_1}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{=}\;\Con{list}\;\V{\tau}\;{}\<[E]%
 \\
@@
 \\
-\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%
+\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{=}\;\Con{list}\;\V{\tau}\;{}\<[E]%
 \\
@@
 can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the
-type \ensuremath{\V{\tau_1}\;\D{\sqcup}\;\V{\tau_2}\;\D{\sim}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
+type \ensuremath{\V{\tau_1}\;\D{\sqcup}\;\V{\tau_2}\;\D{=}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
 The resulting typing context adds information about the newly created list.
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathkw{data}\;\D{\_\sqcup\_\sim\_}\;\mathkw{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub4}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{list}\;\V{t})\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub2}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\>[B]{}\mathkw{data}\;\D{\_\sqcup\_=\_}\;\mathkw{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t}\;\D{=}\;\V{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub4}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{list}\;\V{t})\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub2}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%
 \\
 \>[3]{}\hsindent{8}{}\<[11]%
-\>[11]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;\V{t'}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub2b}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\>[11]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{=}\;\V{t'}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub2b}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%
 \\
 \>[3]{}\hsindent{9}{}\<[12]%
-\>[12]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\>[12]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;\V{t'}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%
 \\
 \>[3]{}\hsindent{10}{}\<[13]%
-\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub6}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{listcons}\;\V{t})\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
+\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;(\Con{list}\;\V{t'}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub6}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{listcons}\;\V{t})\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%
 \\
 \>[3]{}\hsindent{10}{}\<[13]%
-\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
+\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{=}\;(\Con{listcons}\;\V{t'}){}\<[E]%
 \ColumnHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcup\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
+\>[B]{}\F{\sqcup\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcup}\;\V{t1}\;\D{=}\;\V{t3}{}\<[E]%
 \ColumnHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{\sqcup\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
-\\
-\>[B]{}\F{\sqcup\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
+\>[B]{}\F{\sqcup\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
+\\
+\>[B]{}\F{\sqcup\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
 \ColumnHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\V{∃}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t}){}\<[E]%
+\>[B]{}\F{lub}\;\mathbin{:}\;(\V{t1}\;\V{t2}\;\mathbin{:}\;\D{Ty})\;\rightarrow\;\V{∃}\;(\lambda\;\V{t}\;\rightarrow\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t}){}\<[E]%
 \\
@@
 including lexing, parsing, scope checking, and most importantly \emph{type-checking}.
-Since we represent programs using a intrinsically-typed syntax, scope and
+Since we represent programs using an intrinsically-typed syntax, scope and
 type-checking is only a matter of elaborating an untyped syntax to a typed one.
@@
 Next, we follow Norell's~\cite{Norell2009} approach and define a type that relates a
-natural number with a list as follows. Either the number correspond to some list position,
-in which it is of the form \ensuremath{\F{index}\;\V{p}} for some proof \ensuremath{\V{p}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs}} or it is invalid position.
+natural number with a list as follows. Either the number corresponds to some list position,
+in which it is of the form \ensuremath{\F{index}\;\V{p}} for some proof \ensuremath{\V{p}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs}} or it is an invalid position.
 The type \ensuremath{\D{Lookup}} encodes this relation between lists and natural numbers.
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}seq}\;\V{p}\;\V{p₁})\;\mathrel{=}\;\Con{seq}\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{p})\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{p₁}){}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}nil}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{x}\;\V{x₁}\;\V{x₂})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{x})\;\V{l}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}0}\;(\F{index}\;\V{x})\;\V{v'}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{x}\;\V{idx})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}0}\;(\F{index}\;\V{x})\;(\F{index}\;\V{idx}){}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}1}\;(\F{index}\;\V{x})\;\V{v'}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;\V{x}\;\V{idx})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}1}\;(\F{index}\;\V{x})\;(\F{index}\;\V{idx}){}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x}\;\V{x₁}\;\V{x₂})\;\mathrel{=}\;\Con{cons}\;(\F{index}\;\V{x})\;(\F{index}\;\V{x₁})\;\V{v'}{}\<[E]%
-\\
-\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}cons\textrm{-}upd}\;\V{x}\;\V{x₁}\;\V{idx}\;\V{x₂})\;\mathrel{=}\;\Con{cons}\;(\F{index}\;\V{x})\;(\F{index}\;\V{x₁})\;(\F{index}\;\V{idx}){}\<[E]%
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}seq}\;\V{p}\;\V{p₁})\;\mathrel{=}\;\Con{seq}\;(\F{forget\textrm{-}types\textrm{-}instr}\;\V{p})\;(\F{forget\textrm{-}types\textrm{-}instr}\;\V{p₁}){}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}nil}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{x}\;\V{x₁}\;\V{x₂})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{x})\;\V{l}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}0}\;(\F{index}\;\V{x})\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{x}\;\V{idx})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}0}\;(\F{index}\;\V{x})\;(\F{index}\;\V{idx}){}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}1}\;(\F{index}\;\V{x})\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}fetch\textrm{-}1\textrm{-}upd}\;\V{x}\;\V{idx})\;\mathrel{=}\;\Con{fetch\textrm{-}field\textrm{-}1}\;(\F{index}\;\V{x})\;(\F{index}\;\V{idx}){}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\{\mskip1.5mu \V{v'}\;\mathrel{=}\;\V{v'}\mskip1.5mu\}\;\V{x}\;\V{x₁}\;\V{x₂})\;\mathrel{=}\;\Con{cons}\;(\F{index}\;\V{x})\;(\F{index}\;\V{x₁})\;\V{v'}{}\<[E]%
+\\
+\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}cons\textrm{-}upd}\;\V{x}\;\V{x₁}\;\V{idx}\;\V{x₂})\;\mathrel{=}\;\Con{cons}\;(\F{index}\;\V{x})\;(\F{index}\;\V{x₁})\;(\F{index}\;\V{idx}){}\<[E]%
 \ColumnHook
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Con{ok}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;(\V{i}\;\mathbin{:}\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'})\;\rightarrow\;\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{i}){}\<[E]%
+\>[3]{}\Con{ok}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;(\V{i}\;\mathbin{:}\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'})\;\rightarrow\;\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\F{forget\textrm{-}types\textrm{-}instr}\;\V{i}){}\<[E]%
 \ColumnHook
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n})\;\V{\Gamma}\;\V{v}\;\V{v'}\;\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{fetch\textrm{-}field\textrm{-}0}\;\V{v}\;\V{v'})){}\<[E]%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;(\V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n})\;\V{\Gamma}\;\V{v}\;\V{v'}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{21}{}\<[21]%
+\>[21]{}\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{fetch\textrm{-}field\textrm{-}0}\;\V{v}\;\V{v'})){}\<[E]%
 \\
@@
 The type \ensuremath{\Con{cont}} $\Gamma$ is given to program labels and it
-express that the machine can safely jump when its registers
+expresses that the machine can safely jump when its registers
 satisfies $\Gamma$. The inclusion of continuation types makes
@@
 
-Since we have changed the type syntax, we need to adapt the definition of the subtyping relation.
+Since we have changed the typed syntax, we need to adapt the definition of the subtyping relation.
 The continuation type substantially changes the subtyping relation,
@@
 The rule specifies that the least upper bound for types \ensuremath{\Con{cont}} $\Gamma_1$ and \ensuremath{\Con{cont}} $\Gamma_2$ is
-\ensuremath{\Con{cont}} $\Gamma_3$, where $\Gamma_3$ is greatest lower bound for typing contexts $\Gamma_1$ and $\Gamma_2$.
+\ensuremath{\Con{cont}} $\Gamma_3$, where $\Gamma_3$ is the greatest lower bound for typing contexts $\Gamma_1$ and $\Gamma_2$.
 The definition of the greatest lower bound for contexts is as follows:
@@
 The definition of the greatest lower bound for types is as follows. The first set of rules specify the reflexivity,
-that $\top$ type is an identity and $\bot$ is for greastest lower bound for any type $\tau$.
+that $\top$ type is an identity and $\bot$ is for greatest lower bound for any type $\tau$.
 Again, rules to ensure commutativity of the relation are omitted for brevity.
@@
 compatibility of \ensuremath{\Con{list}} and \ensuremath{\Con{listcons}} type constructors with the greatest lower bound relation. The last
-rule show that \ensuremath{\Con{listcons}} $\tau_3$ is the lower bound for \ensuremath{\Con{list}} $\tau_1$ and \ensuremath{\Con{listcons}} $\tau_2$.
+rule shows that \ensuremath{\Con{listcons}} $\tau_3$ is the lower bound for \ensuremath{\Con{list}} $\tau_1$ and \ensuremath{\Con{listcons}} $\tau_2$.
 \[
@@
 Rules for the greatest lower bound relation for continuation types show that the bottom type is the
-lower bound for the list and continuation types. The other rule estabishes the compatibility of
-the continuation type with the lower bound relation and it uses the least upper bound for typing
-contexts relation.
+lower bound for the list and continuation types. The other rule establishes the compatibility of
+the continuation type with the lower bound relation and it uses the least upper bound for the typing
+context relation.
 \[
@@
 Once we have only modified the subtyping and the least upper bound relations, most of the type system
-rules remained unchanged from the original version. In order to support indirect jumps, Appel's
+rules remained unchanged from the original version. In order to support indirect jumps, Appel's et. al.
 approach is to modify the jump instruction and add a new instruction for loading label values into
@@
 Constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} encodes the typing rule for the instruction \textrm{get-label} when
-the label involved is the special label $L_0$. Intutively, the constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} type
+the label involved is the special label $L_0$. Intuitively, the constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} type
 ensures that the type associated with the register for the instruction is the empty list type.
@@
 
-The needed modifications on block syntax are presented next.
+The needed modifications on the block syntax are presented next.
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-We only need to adapt the type of constructor for jump by requiring that the register has a
+We only need to adapt the type of the constructor for jump by requiring that the register has a
 continuation type whose typing context is a subtype of the current block type, which allows
@@
 
-\subsection{Modifications in the intepreter}\label{sec:changesintep}
+\subsection{Modifications in the interpreter}\label{sec:changesintep}
 
@@
         since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
-  \item \textbf{Generate~\LaTeX.} Agda's compiler does have a limited support for typesetting~\LaTeX code.
-        However, we find it lacking some important features like omitting code from typesetted output
-        and use~\LaTeX math-commands instead of unicode characters for mathematical symbols.
+  \item \textbf{Generate~\LaTeX.} Agda's compiler does provide support for typesetting~\LaTeX~code.
+        However, we find it lacking some fine-grained formatting features like use~\LaTeX~math-commands instead of
+        unicode characters for mathematical symbols.
 \end{enumerate}
 
-As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel and Leroy's
+As we could notice, our approach avoids code repetition and decreases the needed LOCs, when compared to Appel's et. al.
 \cite{Appel07} solution.
@@
 reason for this difference is that our intrinsically-typed syntax granted us many properties for free (e.g. type soundness).
-
 Appel et. al. list-machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
 correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics.
-However, we are not able to prove a lemma which stabilishes a coercion between continuation types:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathkw{postulate}\;\V{sub-env}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Gamma'}\;\D{\subset}\;\V{\Gamma}\;\rightarrow\;\D{Env}\;\V{\Gamma}\;\rightarrow\;\D{Env}\;\V{\Gamma'}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-Appel et. al. was able to prove this property in Coq by enconding the machine syntax and semantics using
-a shallow embedding based on his ``very modal model''~\cite{Appel07} and its complete formalization
-demanded around 1750 LOC without considering his auxiliar library.
-Our formalization demanded 1134 LOC and it depends only on the Agda standard library.
+However, we are not able to prove a lemma which establishes a coercion between continuation types:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathkw{postulate}\;\F{sub\textrm{-}env}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Gamma'}\;\D{\subset}\;\V{\Gamma}\;\rightarrow\;\D{Env}\;\V{\Gamma}\;\rightarrow\;\D{Env}\;\V{\Gamma'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Appel et. al. was able to prove this property in Coq by encoding the machine syntax and semantics using
+a shallow embedding based on his ``very modal model''~\cite{Appel07}. Using a semantics-based characterization
+of typing, like Appel's solution to the list machine benchmark, has the benefit of allowing a direct proof
+of the coercion property for continuation types. While is certainly possible to construct a modal logic
+inspired model and from it proving this type coercion result, such approach deviates from our main objective:
+construct a solution to the list machine benchmark using intrinsically-typed syntax and definitional interpreters.
+For this reason, we decided to postulate this coercion property and stick with our original proposal.
+Our formalization demanded 1134 LOC and it depends only on the Agda standard library. Appel et. al.
+solution demanded around 1750 LOC without considering his ``very modal model'' library.
 Compared to our solution to the first version of the benchmark (which was implemented in 612 LOC),
@@
 was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
-machine benchmark was proposed by~\cite{Appel07} as an exercise in formalizing results that interest
+machine benchmark was proposed by Appel and Leroy~\cite{Appel07} as an exercise in formalizing results that interest
 compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
@@
 the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$. As our work,
-Chapman's et al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
-which avoids completely stuck states. Another application of intrisincally-typed syntax is Pardo's et. al. work on correct-by-construction
+Chapman's et al. formalization is an example of how intrinsically-typed syntax leads to clear interpreter code
+which avoids completely stuck states. Another application of intrinsically-typed syntax is Pardo's et. al. work on correct-by-construction
 compilers~\cite{PardoGPV18}. The approach used by the authors is to index the syntax of
 a language with its types and semantics in order to automatically derive a correctness proof for the
-compiler. Authors apply their metodology in an arithmetic expression languages and a imperative
+compiler. Authors apply their methodology in an arithmetic expression languages and a imperative
 language with top-level variables, sequencing and looping statements. }
